Scheduled feeding improves behavioral outcomes and reduces inflammation in a mouse model of Fragile X syndrome

Version 1

Abstract

Abstract

Fragile X syndrome (FXS) is a neurodevelopmental disorder caused by the abnormal expansion of CGG repeats in the fragile X mental retardation 1 (FMR1) gene. Many FXS patients experience sleep disruptions, and we sought to explore these symptoms along with the possible benefits of a scheduled feeding intervention using the Fmr1 knockout (KO) mouse model. These mutants displayed clear evidence for sleep and circadian disturbances including delay in the onset of sleep and fragmented activity rhythms with increases in cycle-to-cycle variability. Importantly, the Fmr1 KO mice exhibited deficits in their circadian behavioral response to light with reduced masking, longer time to resetting to shifts in the Light-Dark cycle, altered synchronization to a skeleton photoperiod and lower magnitude light-induced phase shifts of activity rhythms. Investigation of the retinal input to the surprachiasmatic nucleus (SCN) with the neurotracer cholera toxin (β subunit) and quantification of the light-evoked cFos expression in the SCN revealed an abnormal retinal innervation of the SCN in the Fmr1 KO, providing a possible mechanistic explanation for the observed behavioral deficits. Interestingly, disruptions in social and repetitive behaviors correlated with sleep duration and fragmentation. Understanding the nature of the deficits, we decided to apply a scheduled feeding regimen (6-hr/18-hr feed/fast cycle) as a circadian-based strategy to boast circadian rhythms independently of light. This intervention significantly improved the activity rhythms and sleep in the mutants. Strikingly, the scheduled feeding ameliorated social interactions and reduced repetitive behaviors as well as the levels of Interferon-gamma and Interleukin-12 in the Fmr1 KO mutants, suggesting that timed eating may be an effective way to lessen inflammation. Collectively, this work adds support to efforts to develop circadian based interventions to help with symptoms of neurodevelopmental disorders.

Introduction

Fragile X syndrome (FXS) is a relatively common inherited cause of intellectual disability with a prevalence of about 1 in 4,000 males and 1 in 8,000 females (Hersh and Saul, 2011), with the males more severely affected. FXS present with an abnormal expansion of CGG repeats in the fragile X mental retardation 1 (FMR1) gene. Such trinucleotide repeat expansion causes transcriptional silencing of the FMR1 gene and reduction in the levels of the fragile X mental retardation protein (FMRP), an RNA-binding protein essential for synaptic plasticity among other functions (Zalfa et al., 2006; Soden et al., 2010; Kute et al., 2019). There is no cure for FXS; current treatments primarily focus on managing the symptoms, among which the most common include social deficits, repetitive behaviors and sleep disruptions (Budimirovic et al., 2022; Mendez and Mendez, 2024). Given the central role of sleep and circadian rhythms in restoration and recovery, a key issue is whether improving the sleep/wake cycle could ameliorate other symptoms of FXS.

Animal models are helpful to better understand the mechanisms underlying circadian and sleep disorders as well as to develop strategies for disease management focused on restoring the sleep/wake cycle. Several models of FXS have been put forward, each with a distinct set of advantages and disadvantages (Sandoval et al., 2024); among them the Fmr1 knockout (KO) model has been validated for elucidating disease mechanisms and is widely used for preclinical development of drug candidates including those focused on sleep (Thomas et al., 2012; Kazdoba et al., 2014; Kat et al., 2022; Saré et al., 2022; Martinez et al., 2024). Prior work has found that both the Fmr1/Fxr2 double KO and Fmr1 KO/Fxr2 heterozygous animals exhibit a loss of rhythmic activity in the light/dark (LD) cycle (Zhang et al., 2008). More recent work has found evidence for deficits in sleep (Saré et al., 2017) and activity (Bonasera et al., 2017; Angelakos et al., 2019) rhythms in the Fmr1 KO model. This prior work suggests that, like the FSX patients (Kronk et al., 2009; Budimirovic et al., 2022), sleep/wake disturbances can be seen in the mouse models. Still numerous gaps remain in our knowledge in regard to the nature of the deficits and if correcting the altered sleep/wake cycles would better other aspects of the Fmr1 KO phenotype.

In this study, we sought to comprehensively examine the sleep/circadian disturbances in the Fmr1 KO mice and to answer the question of whether or not these are associated with their behavioral alterations. It is worth noting that much of the literature concerning behavioral deficits in the Fmr1 KO line were conducted in the middle of the day when the mice were normally asleep, and this daytime protocol may cause sleep disruption. To address this issue, we sought to make use of non-invasive monitoring systems to assess the sleep behavior and the activity rhythms of the animals in their home cages. The response of WT and mutants to different behavioral tests was assessed during the active (dark) phase to minimize disruptions to their sleep/wake rhythms. The same mice that were recorded for their sleep/wake cycles were also evaluated with other behavioral tests so that correlations could be examined. Guided by our understanding of the deficits in the Fmr1 KO line, we utilized a circadian-based intervention (schedule feeding, 6-hr feeding/18-hr fast) and determined its impact on sleep/wake rhythms as well as social deficits, and repetitive behavior.

Results

Fmr1 KOs exhibited shorter and fragmented sleep in the light phase

To determine if the Fmr1 KO had deficits in immobility-defined sleep behavior, the animals were examined using a combination of video recording and mouse tracking system under LD (12:12 hr) conditions. All the mice exhibited robust day-night rhythms in sleep with higher levels of sleep during the day and the lowest levels during the first half of the night (Fig. 1). Looking at the total sleep values within a 24 hr cycle, the total amount of sleep and the number of sleep bout did not vary between genotypes, although the Fmr1 KO did exhibit a reduction in the average duration of each sleep bout (Table 1). A two-way ANOVA used to analyze the temporal pattern of sleep (1-hr bins) in WT and mutants indicated significant effects of time (F(23, 287) = 31.94; P < 0.001) and genotype (F(23, 287) = 11.95; P < 0.001). A number of diurnal differences emerged when day and night sleep parameters were analyzed with a two-way ANOVA (Table 1). Focusing our attention on the resting phase (day), the Fmr1 KO mutants exhibited less sleep behavior (Fig. 1B) with more bouts of a shorter duration (Fig. 1C, D). In addition, maximum sleep bout duration was about 20 minutes shorter in the mutant mice. The reduced sleep time, shorter bout length, and a greater number of sleep bouts seen during the day are all evidence of a fragmented sleep pattern in the Fmr1 KO mutants.

The Fmr1 KO mice exhibited shorter and fragmented sleep in the light phase.

(A) Waveforms of daily rhythms in sleep behavior under standard 12:12 h light-dark (LD) cycles in both WT (blue circle) and Fmr1 KO (yellow triangle) mice. By definition, ZT 0 is when lights turn on and ZT 12 is when lights turn off. The sleep waveform (1 hr bins) of each genotype were analyzed using a two-way ANOVA with genotype and time as factors, followed by the Holm-Sidak’s multiple comparisons test. Significant differences (P < 0.05) are indicated with an asterisk (*). Both genotypes exhibited clear rhythms in sleep, with reductions in the mutants mostly found in the light phase. The white/black bar on the top indicates the LD cycle, and the gray shading in the waveforms indicates the dark phase time-period. (B-D) Measures of immobility-defined sleep in the light phase. Fmr1 KO mice display more sleep bouts of shorter duration. Values are shown as the means ± SEM. Genotypic differences were analyzed with a t-test and significant differences (P < 0.05) are indicated with an asterisk (*). See Table 1.

Reduced rhythmic strength and nocturnality in the Fmr1 KOs

Next, we investigated the presence of deficits in locomotor activity rhythms in the Fmr1 KO mice showed. Age-matched WT and Fmr1 KO were housed in cages equipped with running-wheels under LD cycles for 2 weeks and then released into constant darkness (DD) to measure their endogenous rhythms. All of the mice exhibited a robust daily and circadian rhythms in wheel running activity (Fig. 2; Table 2). In the LD environment, some indication of the commonly reported hyper-activity in the Fmr1 KO mice emerged (Fig. 2B), although the activity levels over 24 hours were not significantly different (Fig. 2C) between the genotypes. Analysis of the diurnal activity patterns with two-way ANOVA confirmed the significant effects of time (F(23, 287) = 8.84; P = 0.003) and genotype (F(23, 287) = 39.75; P < 0.001) on locomotor activity. In LD conditions (Fig. 2C), the mutants exhibited lower power rhythms with increased activity during lights-on and higher cycle-to cycle variability; while, when the animals were released in DD (Fig. 2D), both the WT and the mutants displayed very similar free-running period (tau), but again, the latter presented with weaker rhythmic strength and increased cycle-to-cycle variability (Table 2). Together, our data suggest that the Fmr1 KO mice exhibit deficits in the circadian regulation of locomotor activity and raise the possibility of problems in the synchronization to the LD cycle.

The Fmr1 KO mice exhibited unstable locomotor activity rhythms and reduced nocturnality.

(A) Representative wheel-running actograms of daily rhythms in cage activity under LD cycles followed by constant darkness (DD) in both WT (left) and Fmr1 KO (right) mice. The activity levels in the actograms were normalized to the same scale (85% of the maximum of the most active individual). Each row represents two consecutive days, and the second day is repeated at the beginning of the next row. (B) Waveforms of daily rhythms in cage activity in WT (blue circle) and Fmr1 KO (yellow triangle) mice under the LD cycles. The activity waveform (1 hr bins) was analyzed using a two-way ANOVA with genotype and time as factors followed by the Holm-Sidak’s multiple comparisons test. Significant differences (P < 0.05) are indicated with an asterisk (*). There were significant effects of both time (F = 8.84; P = 0.003) and genotype (F = 39.75; P < 0.001) on the temporal pattern of the locomotor activity rhythms. Note that genotypic differences were found before and after dawn. Measures of locomotor activity rhythm parameters under LD (C) and DD (D). The strength of the rhythms is significantly lower in the mutants in both conditions. Histograms show the means ± SEM with the values from individual animals overlaid, and the genotypic differences were analyzed by a t-test (*P < 0.05). The white/black bars on the top of the actograms and waveforms indicate the LD cycle, and the gray shading in the waveforms indicates the time of dark exposure. See Table 2.

Fmr1 KOs displayed an aberrant behavioral response to photic timing cues

The likelihood of the Fmr1 KO mice presenting deficits in their response to light was tested using four behavioral assays and measuring wheel running activity. First, the ability of light (4500K, 50lx) to suppress or “mask” the locomotor activity of nocturnal mice was evaluated. For this assay, the level of locomotor activity between Zeitgeber Time (ZT) 14 and 15 was first measured under baseline LD conditions, and no differences were observed between genotypes (WT: 2229 ± 162 revolutions (rev); KO: 2639 ± 328 rev; t(18) = -1.123, P = 0.277). The next day the mice were exposed to light for 1 hour at this same phase and the levels of activity measured. A two-way ANOVA indicated significant effects of both genotype (F(1, 39) = 12.002; P = 0.001) and light exposure (F(1, 39) = 24.253; P < 0.001). The percentage of suppression in activity elicited by the light exposure was higher in the WT compared to the mutants (Fig. 3A; Table 3), although a couple of KO mice exhibited levels of masking comparable to WT (Fig. 3B). Next, we determined the number of cycles needed by the mice to re-entrain to a 6 hours phase advance of the LD cycle. While the WT mice re-synchronized in 5.9 ± 0.6 days to the new phase, it took 11.3 ± 0.4 days for the Fmr1 KO to adjust (Fig. 3C, D; Table 3). Third, we evaluated the ability of the mice to entrain to a skeleton photoperiod (SPP) in which the full 12 hours of light are replaced by two 1-hr light exposures separated by 11 hours of dark (Fig. 4A). In these experiments, the mice were first entrained to the standard LD cycle and then released into the SPP for 2 weeks (Fig. 4A). The activity recordings demonstrated that the WT were able to entrain to this challenging environment and exhibited robust circadian rhythms with a tau of 24.0 ± 0.0 hr (Table 3). The Fmr1 KO exhibited a tau of 23.7 ± 0.2 hr with the shorter period driven by three mutant mice who failed to stably entrain to the SPP (Table 3). Compared to the WT, the mutants under SPP showed reduced rhythmic strength, increased light-phase activity, and larger cycle-to-cycle onset variability (Fig. 4B). Finally, the direct light-induced phase shift of the circadian system of the mice was measured by exposing them to light (300 lx) for 15min while in constant darkness at circadian time (CT) 16. The WT showed a phase delay of 135.6 ± 26.9 min, whilst, in the Fmr1 KO mutants, the delay in the activity onset on the next day was about half the magnitude (64.0 ± 0.1 min) (Fig. 4C, D; Table 3). These findings provide evidence for the presence of a defective circadian response to light in the Fmr1 KO mice.

The Fmr1 KO mice showed deficits in light-regulated circadian behaviors.

(A, B) Photic-suppression (masking) of activity in mice exposed to light at 300 lx (4500K) for one hour at ZT 14 (lights off; n=10/genotype). The activity level during the light exposure was compared to the activity level during the equivalent hour (ZT 14-15) on the day before the treatment (baseline activity). (A) The genotypic difference in the fold change was determined by t-test, with the mutants showing a significantly reduced suppression of activity as compared to the WT (*P = 0.05). (B) Changes in the activity levels of each individual mice during the baseline window and the light masking were analyzed using a paired t-test in WT (P < 0.001) and Fmr1 KO (P = 0.12). (C, D) Entrainment induced by a 6 hr-phase advanced LD cycle. Examples of light-induced phase shifts of wheel-running activity rhythms (C) of a WT (left) and an Fmr1 KO (right) mice are shown. The white/black bars on the top of actograms indicate the LD cycle before (upper) and after (lower) the 6 hr phase advance. The gray shading in the waveforms indicates the dark phase time-period. The arrows next to the actograms indicate the day when the 6 hr-phase advance was applied. Two-way ANOVA confirmed significant effects of genotype (F(1, 285) = 130.157, P < 0.001). The entrainment shifting in the WT (blue circle) and the Fmr1 KO (yellow triangle) was quantified by the difference between the activity onset and the new ZT12 on each day (D). The yellow and blue arrow heads in the graph indicate the day when the activity rhythms are considered well entrained. The Fmr1 KO took significantly longer to re-entrain to the new LD cycle (P < 0.001). See Table 3.

The Fmr1 KO mice exhibited difficulty in adapting to the skeleton photic period (SPP).

(A) Representative actograms of daily rhythms in cage activity under standard LD cycles (2 weeks) followed by the SPP challenge (1hr:11hr:1hr:11hr LD cycles) in both WT (left) and Fmr1 KO (right) mice. The white/black bars on the top of actograms indicate the baseline LD cycle (upper) and the SPP LD cycles (lower). The gray shading in the waveforms indicates the time of the dark phases. (B) Measures of locomotor activity rhythms under the SPP environment. Many of the parameters measured were significantly different between the genotypes with the mutants being more impacted. Histograms show the means ± SEM with the values from each individual animal overlaid. Significant differences (P < 0.05), determined by t-test or Mann-Whitney Rank sum test, are indicated with an asterisk. See also Table 3. (C, D) Light-induced phase delay of free-running activity rhythms in mice exposed to light (300 lx, 4500K, 15 mins) at circadian time (CT) 16. Mice were held in constant darkness, by definition, CT 12 is the beginning of the activity cycle in DD for a nocturnal organism. Examples of light-induced phase shifts of wheel-running activity rhythms (C) of WT (left) and Fmr1 KO (right) and quantified phase delay (D) In the representative actograms, the yellow lines indicate the best-fit line of the activity onset across the 10 days before and after the light pulse. The amount of phase delay is determined by the difference between the two lines on the day after the light pulse. The sunny-shape symbols indicate when the mice were exposed to light (CT16). Compared to WT, the Fmr1 KO showed reduced phase shift of their activity rhythms (Mann Whitney U *P = 0.011). See Table 3.

Fmr1 KOs exhibited subtle deficits in the retinal afferent innervation to the SCN

The results described above suggest that Fmr1 KO mice could have an anomalous retinal input to the SCN. The fluorescence-conjugated neurotracer Cholera Toxin (β subunit) has been used to map the projections of melanopsin-expressing intrinsically photoreceptive retinal ganglion cells (ipRGCs) from the retina to the SCN (Muscat et al., 2003; Hattar et al., 2006), hence, the WT and Fmr1 KO received a bilateral intravitreal injection of this tracer. A lower fluorescent signal could be observed in the mutant mice both laterally and medially to the ventral SCN (Fig. 5A) as well as beneath in the optic chiasm as compared to the WT. Analysis of the intensity and distribution (Supp. Fig. 1) of the labelled retino-hypothalamic processes entering the ventral SCN showed a reduction, particularly evident in the lateral side of both the left and right mutant SCN (Fig. 5B, C). Likewise, a subtle decrease in the intensity of the labelled fibers was found in the whole SCN (Table 4) of the Fmr1 KO mice as compared to WT.

Abnormal Retinal-Suprachiasmatic Nucleus connectivity in the Fmr1 KO mice.

To trace the projections from the retina to the suprachiasmatic nucleus (SCN) via the Retino-Hypothalamic tract (RHT), WT and Fmr1 KO mice received a bilateral intravitreal injection of Cholera Toxin (β-subunit) conjugated to Alexa Fluor555 and were perfused 72 hours later. (A) Lower intensity of the fluorescently labelled RHT projections can be observed both laterally and medially to the ventral part of the SCN in the Fmr1 KO mice as compared to WT, suggesting a loss of afferent projections to the SCN. (B, C) Densitometric analysis of the distribution of the cholera toxin fluorescence intensity in the ventral SCN (Suppl. Fig. 1) of WT and Fmr1 KO mice. The intensity peaks of the profile plot of 4 to 5 consecutive coronal sections containing the middle SCN were aligned and then averaged to obtain a single curve per animal. Results are shown as the mean ± standard deviation (SD) for the left (B) and the right (C) SCN of each genotype. (D, E) Light-induction of cFos was greatly reduced in the SCN of the Fmr1 KO mice compared to WT. Mice held in DD, were exposed to a light (300 lx, 4500K) pulse for 15 min at CT 16, and perfused 45 minutes later (CT 17). (D) Representative serial images of light-evoked cFos expression in the SCN. The inset in the lower left panel shows the lack of cFos immunopositive cells in the SCN of mice held in DD but not exposed to the light pulse. (E) The number of immune-positive cells in the left and right SCN from 3-5 consecutive coronal sections per animal were averaged to obtain one number per animal and are presented as the mean ± SD per genotype. One-way ANOVA followed by Bonferroni’s multiple comparisons test, *P=0.0201. See Table 4

A well-established test of the light input to the circadian system is the light-evoked cFos response in the SCN. In line with the results obtained with the Cholera Toxin suggesting an impaired light-pathway to the SCN in the mutants, the number of cFos positive cells in the SCN of the Fmr1 KO was greatly reduced (50%) in comparison to the WT mice (Fig. 5D, E; Table 4). Contrary to other models of neurodevelopmental disabilities (Li et al., 2015; Lee et al., 2018), the Fmr1 KO mice did not display any histomorphometrical alteration of the SCN (Table 4). Based on these findings, we can surmise the presence of an abnormal connectivity between the retina and the SCN in the mutants, which could provide, at least in part, a mechanism for their difficulty in responding to photic cues.

Fmr1 KOs exhibited deficits in social interactions as well as increased repetitive behaviors

Social deficits and stereotypic symptoms are hallmark problems in NDDs and we sought to evaluate these behaviors during the night, between ZT 16-18, in the two genotypes. The active-phase social behavior was first tested with the three-chamber test (Fig. 6A). In the first stage of testing, both genotypes showed more interest toward the stranger mouse when they were given the choice between an inert object and a mouse. The direct comparison of the time spent with the object and the novel mouse did not show significant differences between the genotypes (Table 5). On the other hand, measurement of the social preference index (SPI) indicated that the Fmr1 KO exhibited reduced interest in conspecifics compared to WT mice (Table 5). In the second testing stage, the lifeless object was replaced with a second novel mouse while the stranger mouse from the first stage had become a familiar mouse (Fig. 6B). Under these conditions, the Fmr1 KO spent more time with the familiar mouse than the WT (Table 5). Again, the social novelty preference index (SNPI) indicated that the Fmr1 KO exhibited reduced interest in novel mouse compared to WT mice (Table 5). The reduced time in exploring and staying in the novel-mouse chamber suggested that the Fmr1 KO mutants were not able to distinguish the second novel mouse from the first now familiar mouse. The possibility of impaired social memory was further tested with the 5-trial social test (Fig. 6C). In this test, the first stranger mouse becomes a familiar mouse after 4 exposures to the testing mouse. When the second novel mouse is introduced in the 5th trial, the testing mouse typically shows a boosted interest in investigating the novel mouse. As expected, when the second novel stranger mouse replaced the first familiar mouse and was introduced to the testing animals in the 5th trial, WT showed elevated social behavior. In contrast, the mutants did not display increased interest in exploring the second novel mouse (Table 5).

The deficits in social recognition and repetitive behaviors of the Fmr1 KO mice correlate with altered sleep behavior.

Social behavior was evaluated with the 3-chamber social test. (A) In the first stage, the testing mouse was given a choice between a novel mouse and an inanimate object. The social preference index (SPI) was determined. WT mice preferred to spend time with a novel mouse compared to the Fmr1 KO and had a higher SPI. (B) In the second stage of the 3-chamber test, the testing mouse was given the choice between a chamber with a novel mouse and one with the familiar mouse. WT mice preferred to spend time with the novel mouse compared to the familiar one and had a higher SNPI compared to the mutants. (C) The possibility of impaired social memory was further tested by the 5-trial social test. In this test, the first stranger mouse becomes a familiar mouse after 4 exposures to the testing mouse. When a novel mouse was introduced in the 5th trial, the WT mice showed a higher interest for the novel mouse compared to the Fmr1 KO mice. Test of repetitive behaviors were also performed. The amount of digging in the bedding (D) and the percentage of marbles buried (E) were measured with the marble bury test. Fmr1 KO mice spent longer time digging and buried more marbles compared to WT. (F) Grooming behavior, assessed in a novel arena, was significantly higher in the Fmr1 KO mice as compared to WT. Histograms show the means ± SEM with the values from the individual animals overlaid. Significant differences (P < 0.05) by t-test or Mann-Whitney Rank sum test are indicated with an asterisk. See also Table 4. Sleep duration (G, H) and sleep fragmentation (I, J) were correlated with impaired social recognition and abnormal grooming behaviors (Pearson Correlation test). See Table 5.

Next, when the active-phase repetitive behavior was examined using the marble bury test, the Fmr1 KO spent more time digging and buried more marbles compared to WT (Fig. 6D, E; Table 5). When this 30-min trial was divided into three 10-min intervals, the repetitive digging behavior was significantly higher in the Fmr1 mutants compared with the WT through all three intervals. Two-way ANOVA demonstrated that there was a significant effect of genotype (F(1, 107) = 11.04; P = 0.001) but not of the interval (F(2, 107) = 0.42; P = 0.66). The Fmr1 KO also exhibited significantly more grooming than their WT counterpart (Fig. 6F, Table 5). Finally, we used the data collected to determine if disrupted sleep correlated with NDD-like behavioral deficits. Significant correlations were found in the social memory tested by the 5-trial social interaction test, the percentage of buried marbles, and the grooming time (Table 6). Sleep duration (Fig. 6G, H) and fragmentation (Fig. 6I, J) exhibited a moderate-strong correlation with both social recognition and grooming. In addition, we also found a moderate correlation between the grooming time and the circadian rhythmic power as well as the activity onset variability (Table 6). In short, our work demonstrated that even when tested at their circadian active phase, the Fmr1 KO mice exhibit robust deficits in social memory and repetitive behavior. Moreover, the shorter and more fragmented the daytime sleep, the more severe the social memory impairment and the repetitive behavior in the mutants.

A scheduled feeding paradigm ameliorated sleep disturbances and improved the deficits in social memory and repetitive grooming behavior in the Fmr1 KO

The correlations between the fragmented sleep and the severity of other behavioral phenotypes support the possibility that interventions focused on improving circadian rhythms and sleep may benefit the observed deficits in the Fmr1 KO mice. Scheduled feeding can be a powerful regulator of circadian rhythms (Long and Satchidananda, 2022; Manoogian et al., 2022) and has been shown to be effective in several disease models (Whittaker et al., 2018; Wang et al., 2018; Gao et al., 2022; Gupta et al., 2022; Whittaker et al., 2023). We, therefore, sought to determine if a scheduled feeding paradigm (time restricted feeding, TRF) consisting of 6-h feeding/18-h fast for 2 weeks would benefit the Fmr1 KO mice, and compared the 2 genotypes held on ad libitum feeding (ALF) as well as TRF. We confirmed that food consumption was very similar in the four groups by the second week (S. Fig. 2); both genotypes were able to well adapt to the TRF protocol and ate as much as their ALF controls after 5 days. At the end of the two-week treatment, the TRF groups weighed less than the ALF groups (S. Fig. 2). The TRF treatment benefited a number of aspects of the temporal patterning of activity and sleep (Fig. 7) in both genotypes. The scheduled feeding improved the power of the rhythms while reducing the fragmentation in both WT and Fmr1 KO, and markedly restored the cycle-to-cycle variability in the mutants to WT levels (Fig. 7C-E; Table 7). The total cage activity was not altered by the intervention in either genotype. When the sleep behavior was assessed (Fig. 7F, G), TRF increased the total amount of daytime sleep (Fig. 7H; Table 7), reduced sleep fragmentation (bout number, Fig. 7I) and increased the average duration of each bout (Fig. 7J) in both genotypes. During the light phase, the TRF-treated mutants spent more time sleeping with longer sleep bout length and a smaller number of sleep bouts than the ALF controls. In general, strong genotypic differences were present in the response to TRF with the WT mice showing a more robust sleep response (Table 7).

Amelioration of sleep/wake rhythms in the Fmr1 KO mutants by TRF.

(A, B) Waveforms of daily rhythms in cage activity using IR detection in the WT (circle) and Fmr1 KO (triangle) mice under ad lib feeding (ALF) or TRF. The activity waveforms (1 hr bins) were analyzed using a three-way ANOVA with genotype, treatments, and time as factors followed by Holm-Sidak’s multiple comparisons test. There were significant effects of genotype (F(1, 767) = 13.301; P < 0.001) and time (F(23,767) = 94.188; P < 0.001), as well as significant interactions between genotype and time (P < 0.001) and treatment and time (P < 0.001) on the locomotor activity rhythms of both WT and Fmr1 KO mice. The green area indicates the time-period when food hoppers were opened for the TRF groups, 6 hours between ZT 15 and ZT 21. (C-E) Measures of locomotor activity rhythms. Both genotypes exhibited an increase in the rhythm power under TRF compared to ALF controls. The increase in early day activity and cycle-to-cycle variation seen in the Fmr1 KO mice was corrected by the TRF. Data are shown as the means ± SEM; two-way ANOVA followed by Holm-Sidak’s multiple comparisons test with genotype and diet as factors, *P < 0.05 significant differences between diet regimens; #P < 0.05 significant differences between genotypes. See also Table 6. (F, G) Waveforms of daily rhythms in the immobility-defined sleep. The sleep waveforms (1 hr bins) were analyzed using a two-way ANOVA with time and the feeding treatments as factors followed by the Holm-Sidak’s multiple comparisons test. There were significant effects of time for both WT (F(23, 351) = 9.828, P <0.001) and Fmr1 KO (F(23, 351) = 1.806, P = 0.014) mice. Treatment did not significantly affect either genotype. Missing data points precluded the use of three-way ANOVA for these measures. (H-J) Measures of immobility-defined sleep in the light phase. Both genotypes held on TRF exhibited an increase in sleep duration and in sleep bout duration as well as a reduction in sleep fragmentation compared to ALF controls. Data are shown as the means ± SEM; two-way ANOVA followed by Holm-Sidak’s multiple comparisons test with genotype and diet as factors, *P < 0.05 significant differences between diet regimens; #P < 0.05 significant differences between genotypes. See Table 6.

The benefits of the TRF were also evident on other behaviors in the Fmr1 KO model. Focusing on social memory and the grooming behavior, we found that this feeding regimen improved memory while reducing the inappropriate grooming behavior (Table 8). For these experiments, the social memory was evaluated with the 5-trial social interaction test as described above (Fig. 8A). The interest of the Fmr1 KO mice in interacting with the second novel mouse was significantly increased to WT-like level, suggesting that the treated mutants were able to distinguish the novel mouse from the familiar mouse. Regarding the repetitive grooming behavior, the TRF-treated Fmr1 KO mice showed a significant reduction in the time spent on self-grooming (Fig. 8B). This reduction was not due to changes in their locomotor ability as there was no effects on the distance travelled during the test between the TRF and the ALF groups (Fig. 8C). In summary, TRF not only improved the sleep/wake cycles but also the deficits in social memory as well as the self-grooming behavior in the Fmr1 KO mice.

TRF improved social memory and stereotypic grooming behavior in the Fmr1 KO mice.

(A) Social memory was evaluated by the 5-trial social interaction test as described above. The social memory of the Fmr1 KO was significantly augmented by the TRF intervention, suggesting that the treated mutants were able to distinguish the novel mouse from the familiar mouse. The left panels show the time spent in social interactions when the second novel stranger mouse was introduced to the testing mouse in the 5-trial social interaction test. The significant differences were analyzed by two-way ANOVA followed by the Holm-Sidak’s multiple comparisons test with feeding treatment and genotype as factors. *P < 0.05 indicates the significant time spent with novel mouse compared to familiar mouse. Note that the Fmr1 KO under ALF exhibited a slight preference for the novel mouse while under TRF this preference was higher. (B) Grooming was assessed in a novel arena in mice of each genotype (WT, Fmr1 KO) under each feeding condition and the resulting data analyzed by two-way ANOVA followed by the Holm-Sidak’s multiple comparisons test with feeding treatment and genotype as factors. *P < 0.05 indicates the significant difference between the treatments, and #P < 0.05 those between the genotypes. (C) Distance travelled in the grooming test. TRF did not alter the overall locomotion in the treated mice. See Table 7.

Scheduled feeding treatment reduced the elevated IL-12 and INFƳ pathway in the Fmr1 KO mice

An abnormal immune response has been suggested to play a role in FXS pathophysiology (Reynolds et al., 2021; Dias et al., 2022; Robinson-Agramonte et al., 2022). To investigate whether the scheduled feeding protocol impacted cytokine profiles in the Fmr1 KO mice, plasma samples were collected at the end of the intervention. Our assessment revealed that several cytokine levels were altered in the mutants, and these changes were countered by TRF (Table 9). The comparison within the ALF groups revealed that the mutants had elevated levels of Interleukin-12, (IL-12), Interferon-gamma (IFNƳ), and of the Chemokine Ligand-9 (CXCL-9) compared to WT. The observed genotypic differences in these pro-inflammatory markers were abolished in the TRF-treated groups (Fig. 9A). The scheduled feeding also repristinated the levels of IL-2 in the mutant to WT levels. We next asked whether there was a correlation between the levels of these proinflammatory markers and the expression of the behavioral phenotypes. We focused our attention on IL-12 and IFNƳ axis because their elevated levels in the KO were successfully reduced by TRF. Markers of poor sleep such as higher activity in the light phase, shorter sleep time/sleep bout length, and higher numbers of sleep bouts were significantly associated with higher levels of IL-12 and IFNƳ (Table 10). The levels of IL-12 and IFNƳ were also significantly associated with the social memory impairments and the severity of repetitive grooming behavior (Fig. 9B, C; Table 10). Together, our data demonstrate that TRF treatment effectively reduced the levels of pro-inflammatory cytokines in this model of FXS.

The plasma levels of IL-12 and IFNƳ correlate with altered sleep/wake rhythms and autistic behaviors, and are corrected by TRF in the Fmr1 KO mice.

(A) The levels of selected plasma pro-inflammatory markers are shown. The full list of the assayed makers is reported in Table 8. Data were analyzed with two-way ANOVA followed by the Holm-Sidak’s multiple comparisons test with treatment and genotype as factors. *P < 0.05 indicates the significant difference between the feeding treatments, and #P < 0.05 between the genotypes. (B-G) Correlations between IL-12 or IFN-γ and sleep time, social recognition, and grooming behavior. Data were analyzed using the Pearson Correlation, and the coefficients are reported in Table 9.

Discussion

A significant proportion of individuals with NDDs such as FXS exhibit disruptions to their daily sleep/wake cycles. Particularly common are difficulty falling asleep, staying asleep, and altered sleep patterns (Budimirovic et al., 2022; Minhas et al., 2024), which have a major impact on the quality of life of the patients. Individuals with the worse set of symptoms in other behaviors also experience the most pronounced sleep problems (Schreck et al., 2004; Taylor et al., 2012; Kaufmann et al., 2024) raising the possibility that the disturbances in the sleep/wake cycle contribute to the severity of other behavioral systems. Notably, abnormalities in sleep are detected early in development, even in infants and toddlers carrying the FMR1 mutation (D’Souza et al., 2020). These sleep/wake disruptions may be due to a direct effect of the mutation itself on the circuitry underlying circadian rhythms and sleep homeostasis or indirect effects due to other symptoms including anxiety, sensory hypersensitivity and repetitive behaviors that would themselves disrupt the sleep/wake cycle.

As with other symptoms (Sandoval et al., 2024), animal models are helpful for better understanding the mechanisms underlying circadian and sleep disorders as well as for developing strategies focused on restoring the sleep/wake cycle. In the present study, we used a mouse model to demonstrate that loss of Fmr1 has deleterious effects on the temporal pattern of sleep and locomotor activity with strong deficits in the response to the environment lighting. The disrupted rhythms were significantly associated with social memory impairments and repetitive behaviors. Critically, we found that the circadian-based treatment strategy, scheduled feeding, was effective in countering the behaviors deficits as well as reducing the inflammatory signature in the Fmr1 mutants.

In our behavioral sleep assays, we observed reduced total sleep time and sleep fragmentation in Fmr1 KO mice (Fig. 1). Similarly, Saré and colleagues found evidence for an impact of the Fmr1 KO on sleep duration during the light phase using a comparable assay (Saré et al., 2017). Another study (Westmark et al., 2023) examined sleep architecture in the Fmr1 KO using electroencephalography (EEG). While this work focused on the effects of a ketogenic diet, the control recordings did not find evidence for differences between the Fmr1 KO and WT. We cannot explain the differences in the findings between these studies and would only point out that perhaps the sleep phenotype under baseline conditions is subtle. A gap in this literature is the lack of investigation into sleep homeostatic mechanisms. It is possible that a sleep deprivation protocol could uncover deficits in the Fmr1 KO model. Notably, these mutant mice display neural oscillation abnormalities, including increased resting state gamma power and heightened amplitude of auditory evoked potentials, paralleling the EEG signatures observed in FXS patients (Jonak et al., 2024).

In this study, we assessed diurnal and circadian rhythms in locomotor activity using a wheel-running assay (Fig. 2). The Fmr1 KO mice exhibited weaker rhythms, with increased cycle-to-cycle variation and heightened daytime activity. These disruptions persisted when the mice were housed in DD, where locomotor rhythms are regulated by the endogenous circadian system without direct light influence. Previous research has shown that the Fmr1/Fxr2 double KO and the Fmr1 KO/Fxr2 heterozygous mice display a marked loss of rhythmic activity in a LD cycle (Zhang et al., 2008). While this early study did report effects in the Fmr1 KO alone, it did not examine the single mutation in detail. Other studies measuring home cage activity found that the Fmr1 KO mice exhibited significantly reduced movement during the active phase (the dark cycle) and increased bouts of activity during the light cycle compared to WT mice (Bonasera et al., 2017). Similarly, hypoactivity during the night was a prominent feature in male Fmr1 KO mice, as measured by infrared beam breaks (Angelakos et al., 2019). Collectively, these studies consistently demonstrate disrupted locomotor activity rhythms in the Fmr1 KO mice. The increased daytime activity and greater cycle-to-cycle variability may be attributable to deficits in the light response of the circadian system.

A well-defined circuit mediates the effects of light on the circadian system entrainment. The photic signal is detected by the ipRGCs, which express the photopigment melanopsin (Hattar et al., 2002; Panda et al., 2005). The ipRGCs receive input from rod and cone photoreceptors and project to the central circadian clock in the SCN, as well as to other brain regions involved in cognitive regulation (Lucas et al., 2012; Mahoney and Schmidt, 2024). We conducted a series of experiments to determine if this circuit is compromised in Fmr1 KO mice (Figs. 3 & 4). Behaviorally, we identified deficits in the light-evoked suppression of activity (negative masking), the speed of re-entrainment to a new LD cycle, and the light-induced phase shifts of the circadian system. A particularly intriguing assay we used involved exposing the mice to a skeleton photoperiod, consisting of one hour of light followed by 11 hours of darkness. While the WT mice readily synchronized to this skeleton photoperiod, Fmr1 KO mice exhibited several difficulties. These behavioral findings confirmed impairments in the light input to the circadian system, prompting further examination of this pathway.

Using fluorescently labeled cholera toxin β-subunit, we demonstrated that the ipRGC innervation of the SCN is defective in the Fmr1 KO mice (Fig. 5A). Additionally, the light-evoked increase in cFos expression in the SCN was strikingly reduced in the mutants (Fig. 5B). Together, the behavioral, anatomical, and functional data suggest deficits in the circadian light response in Fmr1 KO mice. Since FMRP expression has been reported in the ipRGCs (Zhang et al., 2020), these findings may be explained by a loss of function in the retina. Interestingly, these results highlight a reduced circadian light response, contrasting with the general sensory hypersensitivity observed in humans with FXS and Fmr1 KO models (Rais et al., 2018). It will be important to investigate whether these prominent defects in the light input seen in this mouse model are also present in FXS patients. If the circadian light response is indeed compromised, it could have practical implications for the management of FXS.

There was no significant shortening of the free-running circadian period (tau) when activity was measured in constant darkness (Fig. 2D; see also Angelakos et al., 2019). The free-running period, as assessed by locomotor behavior in DD, is a highly sensitive marker of the output from the core molecular clock driving circadian oscillations (Takahashi et al., 2008). Furthermore, gene expression measurements in the central clock (SCN) did not reveal any alterations in the rhythmic transcription patterns of the core clock genes (Per1, Per2, Bmal1, or Cry1) in the Fmr1 KO mice (Zhang et al., 2008). Given the absence of changes in tau or clock gene expression in the SCN, it is unlikely that the core molecular clock mechanism underlying circadian oscillation is impaired in these mutants. Still our data show that this FXS model exhibits deficits in the photic regulation of the circadian clock and its rhythmic outputs. Although numerous molecular and cellular pathways are impacted by loss of FMRP, several studies suggest that changes in excitatory/inhibitory (E/I) balance and circuit hyperexcitability may contribute to FXS symptoms. Within the SCN, there is evidence that E/I balance plays a crucial role in network synchronization (Olde Engberink et al., 2023). Therefore, future research should investigate the impact of the Fmr1 mutation on the E/I balance within the SCN. Regardless of the underlying mechanism, the present study demonstrates that the Fmr1 KO mice show disruptions in the temporal patterning of activity and sleep, which aligns with a broader pattern of deficits observed in mouse models of monogenic neurological disorders (Li et al., 2015; Angelakos et al., 2017, 2019; Brown et al., 2019; Shi and Johnson, 2019; Takumi et al., 2019; Wang et al., 2020, 2023).

The Fmr1 KO mouse model replicates many of the neurological and behavioral deficits observed in FXS (Melancia and Trezza, 2018; Kat et al., 2022). In the present study, we report clear deficits in social recognition and repetitive behaviors, such as excessive grooming. By collecting data on sleep, activity rhythms, and other behaviors from the same animals, we were able to investigate correlations between these behaviors (Fig. 6). Based on human data, we hypothesized that individual mice exhibiting the most disrupted sleep would also show the most severe behavioral deficits (Schreck et al., 2004; Taylor et al., 2012; Kaufmann et al., 2024). For instance, one study found that objective sleep and circadian disturbances accounted for approximately 30% of the variance in stereotypic behaviors among ASD patients (Yavuz-Kodat et al., 2020). Consistent with this, we found that mutant mice with the shortest sleep duration, the highest sleep fragmentation, and the weakest rhythm power performed the worst in tests of social recognition and showed the highest levels of grooming. While these correlations do not establish causation, they align with findings from human studies, which consistently report that individuals with poorer sleep perform worse behaviorally the following day (Robinson-Shelton and Malow, 2016; Budimirovic et al., 2022; Schwichtenberg et al., 2022; Minhas et al., 2024).

A wide range of growing studies have demonstrated that disrupted circadian rhythms and poor sleep lead to a cluster of symptoms, including metabolic deficits, cardiovascular problems, abnormal immune activities, and cognitive deficits (Abdul et al., 2021; Cheng et al., 2021; Johnson et al., 2021; Bisdounis et al., 2022). Many of these same symptoms are seen in FXS and ASD patients (Parente et al., 2022; Protic et al., 2022), indicating a potentially bidirectional link between the circadian clocks and the core NDD symptoms. No medications are available to treat these core symptoms (i.e., deficits in social communication, sensory aberrations, stereotypic behaviors, and restricted interests). Hence, a bidirectional link is exciting as it would raise the possibility of interactions that, by improving the daily rhythms, could have therapeutic utility in a variety of NDD conditions. A key test of this hypothesis is to determine if an intervention that improves the sleep/wake cycle in a disease model also improves other phenotypes.

Other research has demonstrated sleep benefits with various dietary interventions. Recent studies have shown that a ketogenic diet (KD) increases non-rapid eye movement (NREM) sleep and reduces sleep fragmentation while flattening day/night rhythms in Fmr1 KO mice (Westmark et al., 2024). This same group also found that the KD reduced seizures but did not observe significant KD-responsive effects in several other behavioral tests, such as marble burying, tail suspension, and fear conditioning (Westmark et al., 2020). In our own hands, we found that the KD improved sleep in a mouse model of Huntington’s disease (HD) (Whittaker et al., 2022). In this model, improvements in sleep and circadian rhythms paralleled enhanced motor performance. Interestingly, one consequence of placing mice on a KD is the induction of a robust daily rhythm in ketone body levels (Whittaker et al., 2022), suggesting that the KD may function as a type of circadian intervention.

The most common intervention to improve sleep in NDD patients would be to give a sedative before bed. There is a surprising lack of clinical evidence to support this practice or even to suggest which hypnotic drug to use. Both clinical and pre-clinical work is needed to address these issues. In the Frm1 model, three different classes of hypnotics (zolpidem, ramelteon, DORA-22) were all found to improve sleep (Saré et al., 2022). The rationale behind the use of these different drugs is related to their different pharmacological profiles and mechanisms involved in sleep i.e. zolpidem is an allosteric modulator of the GABAA receptor, ramelteon is a melatonin receptor agonist, and DORA-22 is a dual orexin receptor antagonist. Pharmacologically increasing sleep was found to reduce anxiety as measured by the open field but had negligible effects on social behavior in the Fmr1 KO. In addition, a recent study used EEG measurements to demonstrate that a novel hypnotic (ML297, an activator of G-protein-activated inward-rectifying potassium channels) increased NREM sleep duration and reduced fragmentation in Fmr1−/y mice (Martinez et al., 2024). Importantly, this study was also able to show that such treatment following contextual fear or spatial learning improved memory consolidation in the mutant mice.

We tested the hypothesis that improving sleep and circadian rhythms would alleviate behavioral deficits in Fmr1 KO mutants using a time-restricted feeding (TRF) protocol. TRF was selected over light therapy due to the deficient responses to photic cues of the mutants. Even after just a two-week treatment, TRF produced significant improvements in the Fmr1 KO mice. Specifically, TRF enhanced rhythmic power, reduced variability, decreased daytime activity, lengthened and consolidated sleep bouts during the light phase (Fig. 7). Crucially, TRF led to improvements in social memory and reduced repetitive grooming behavior (Fig. 8). Previous studies have demonstrated that TRF can improve circadian behavior in aging (Acosta-Rodríguez et al., 2022; Roth et al., 2023) and neurodegenerative models (Wang et al., 2018; Whittaker et al., 2018; 2023). Prior work has shown benefits of intermittent fasting in uncovering and rescuing cognitive phenotypes in the PTEN model of NDD (Cabral-Costa et al., 2018). However, to our knowledge, this is the first demonstration of TRF benefiting a model of NDD.

Our findings on the levels of peripheral cytokines (Fig. 9A) suggest that inflammation may play a mechanistic role linking improvements in sleep/wake rhythms to reductions in NDD-like behaviors. Altered immune profiles and exaggerated responses to immune challenges have been reported in both FXS patients (Ashwood et al., 2010; Careaga et al., 2014; Van Dijck et al., 2020) and Fmr1 KO mice (Pacey et al., 2015; Hodges et al., 2020; Bertrand et al., 2021), indicating that the absence of FMRP may contribute to immune system imbalance. In our study, most pro-inflammatory cytokines (>80%) did not vary by genotype (Table 9). However, IL-12 (p40) and IFN-γ were elevated in Fmr1 KO mice and significantly reduced by the TRF protocol (Fig. 9A, Table 9). Notably, we found significant correlations between plasma levels of these two inflammatory markers and the extent of sleep disturbances and behavioral deficits (Fig. 9B, Table 10). Similar associations between dysregulated cytokine levels and impaired behavior have been documented in both patients and preclinical models (Siniscalco et al., 2018; Matta et al., 2019). For instance, elevated levels of IL-6 and IL-12 have been associated with increased stereotypy, cognitive impairments, anxiety, and reduced social interactions in autism spectrum disorders (Konsman et al., 2008; Ashwood et al., 2011; Wei et al., 2012; Fallah et al., 2020). Moreover, counteracting immune dysregulation has been proposed as a strategy to improve behavioral deficits in the Fmr1 KO mice (Goo et al., 2020). At least one previous study found that TRF reduced the levels of inflammatory cytokines in mice (Chen et al., 2023). Whether TRF directly counteracts neuroinflammation in Fmr1 KO mice remains to be explored, as does the specific role of IL-12 (p40) and IFN-γ in mediating the observed behavioral disruptions. Future studies will be needed to clarify these mechanisms.

Methods

Animals

The experimental protocols were approved by the UCLA Animal Research Committee and adhered to the UCLA Division of Laboratory Animal Medicine (DLAM) and National Institute of Health (NIH) guidelines. Fmr1 KO mice (Fmr1tm4Cgr) on the C57BL/6J background (JAX ID: 003025) and wild-type (WT) (JAX ID: 000664) were acquired from the Jackson laboratory (Bar Harbor, ME). The studies were carried out in male adult mice (3 to 5 months old) housed in light-tight ventilated cabinets in temperature- and humidity-controlled conditions, with free access to food and water ad libitum except for the experimental groups held on the timed-restricted feeding (TRF) regimen. Mice were individually housed throughout experimental protocols.

Animals tested for social and/or stereotypical behaviors were habituated to the testing room at least 30 mins before the experiments started. The behavioral assays were performed under dim red light (< 2 lx at the level of the testing arena) during the animal active phase, between ZT 16 & 18.

Immobility-Based Sleep Behavior

The sleep behavior under 12h:12h LD cycles was recorded using Anymaze automatic mouse tracking software (Stoelting Co., Wood Dale, IL) that tracked the (40 sec or greater as previously described (Wang et al., 2016; Whittaker et al., 2018). This threshold was previously determined by Fisher and his colleagues to have 99% correlation with EEG-defined sleep (Fisher et al., 2016). From 5 days of recordings, 2 days with the best recording quality were picked for the analysis. A sleep bout was defined as a time period in which activity stayed above the bout threshold (3 counts of sleep per minute for longer than one minute at a time). Acquiring data were exported in 1-min bins and the total sleep during the light phase and the dark phase was obtained.

Locomotor cage activity rhythms

To characterize the sleep/wake cycles, the WT and the Fmr1 KO mice were single housed under a scheduled 12h:12h LD cycles. The mice were habituated to this cycle for 2 weeks to ensure entrainment to the appointed LD schedule before activity data was used for analysis. The animals were then released into constant darkness (12h:12h dark-dark, DD) for the assessment of the endogenous rhythms. They were kept in DD for at least 2 weeks with free access to food and water. The locomotor activity under LD cycles and DD was monitored using running wheel sensors that tracked wheel-running activity. In the set of TRF experiments, the locomotor activity rhythms were monitored using a top-mounted passive infra-red (IR) motion detector. Both the wheel-running recordings and the IR recordings were reported to a VitalView data recording system (Mini Mitter, Bend, OR) as previously described. 10 days of recordings were used for analysis and waveform presentations. The analysis was conducted by using the El Temps (A. Diez-Nogura, Barcelona, Spain) and ClockLab (Actimetrics, Lafayette Instruments, Lafayette, IN) programs as previously described (Loh et al., 2014; Lee et al 2018). The periodogram generated by the El Temps provided measures on the period, or the length of one activity cycle, and the power of the rhythmicity. The power, or percent variance (%V), assessed the strength of the mouse’s periodicity corrected for activity amount and normalized to variance percent derived from peak significance (P=0.05). The fragmentation and the imprecision of daily onset of sleep/wake cycles were determined by the ClockLab. Fragmentation was determined by the number of activity bouts per day (maximum gap: 1 min; threshold 3 counts/min), and the imprecision was determined by calculating the variability of the time of activity onset between the best-fit line of the 10 analyzed days.

Light-regulating circadian behavior

A cohort of age-matched WT and Fmr1 KO mice were single housed in cages with running wheels under a 12:12 LD cycle (300 lx, 4500K) first and then underwent through four behavioral assays of the photic regulation of the circadian system. (1) Photic suppression on nocturnal activity (negative light masking): mice were exposed to 1 hr of light (300 lx, 4500K) at ZT 14. The number of wheel revolutions during this pulse of white light was compared to the number of wheel revolutions during the equivalent hour collected from the previous day. The fold changes were reported: [(rev during the 1-hr light exposure) – (rev during the dark baseline)]/(rev during the dark baseline) %. (2) Re-entrainment to a 6hr advance of the LD cycle: After stable entrainment, the LD cycles were advanced by 6-hr. The activity was continuously monitored by the wheel-running system and the activity onset determined (VitalView). The re-entrainment was quantified by the difference between the running activity onset and the new ZT 12 in each recording day. A mouse was determined to be full entrained when the activity onset was aligned to lights-off and there was no phase shifting for the consecutive 5 days. (3) Skeleton Photoperiod: After stable entrainment to the LD 12:12 cycle, mice were placed in a skeleton photoperiod consisting of 1:11: 1:11 LD. The 1-hr light treatments at 300 lx were given at the beginning (ZT 0) and the end (ZT 11) of the original light phase. The mice were kept in this photic condition for at least 2 weeks. Basic activity rhythm parameters were determined (ClockLab). (4) Light-induced phase shifts to a single exposure to light: Mice were released into DD for 10 days. The circadian time (CT) of their free-running activity rhythms was determined using VitalView and El Temps software. The time of activity onset under DD was defined as CT 12. On day 11, the mice were exposed to light (300 lx, 4500K, 15 min) at CT 16. After the light pulse, the mice remained in DD for an additional 10 days. The best-fit lines of the activity onsets of sleep/wake cycles before and after the light exposure were measured.

Stereotypic Behavior

The marble burying test was used to evaluate the repetitive digging behavior (Yrigollen et al., 2019). The testing mice were habituated to the testing room for at least 30 mins before the test. An array of 4 marbles by 6 marbles was placed in the testing arena with a layer of 4-cm deep shavings. The testing mouse was introduced to the arena from the corner and allowed to freely behave during the trial. Their behavior was recorded for 30 mins and the testing mouse was carefully returned to their home cages. After the test, the number of buried marbles was counted. Mable buried more than ⅔ area were counted, and all marbles were cleaned with 70% ethanol before the next use. Time spent on digging was manually scored and the distance travelled was derived from the automatic mouse-tracking system (Anymaze software, Stoelting Co., Wood Dale, IL).

The grooming test was conducted as in our previous work (Wang et al 2020). The behavior of self-grooming was defined as the cleaning, licking, or washing of the limbs, tail, and body surface areas, typically from a head to tail direction, but excluded bouts of scratching. Time spent on self-grooming was manually scored and the distance travelled was derived from the automatic mouse-tracking system (Anymaze software).

Social Behavior

The three-chamber test was performed as previously described (Wang et al 2020; 2022). The testing mice were allowed to freely explore an arena with three chambers where the central chamber remained empty. When being habituated to the three-chamber arena, both the mutants and the WT explored the arena evenly with no preference toward the left or the right chamber (left/right ratio: WT: 0.95 ± 0.14; P = 0.43 by paired t-test. Fmr1 KO: 0.82 ± 0.1; P = 0.085 by paired t-test). The three-chamber test consisted of two parts: the first stage assessed the preference of social approach toward the stranger mouse, and the second stage assessed the ability of social discrimination of the testing mouse. In the first stage (social approach), an up-turned metal-grid pencil cup was placed in the side chambers: one remained empty as the novel object (the object chamber), and a never-met stranger mouse that matched the sex, age, and genotype of the testing mouse was placed in the second up-turned cup (the social chamber). Therefore, the testing mouse was tested for the preference between the object chamber and the social chamber in this first testing stage. In the second stage (social discrimination), the first stranger mouse and the cup remained the same while a second never-met stranger mouse that mached the sex, age, and genotype of the testing mouse was placed in the second up-turned cup. In other words, the social chamber in the first stage became a familiar chamber in the second stage, and the object chamber in the first stage became a novel chamber in the second stage. Thus, the testing mouse was tested for the preference between the familiar chamber and the novel chamber in this second testing stage. Time spent in each chamber and the distance travelled were derived from the automatic mouse-tracking system (Anymaze software).

The 5-trial social test was conducted as previously described (Mineur et al., 2006). The testing mouse was first habituated to a testing arena for 30 mins and then introduced to a never-met stranger mouse for 4 trials. The testing mouse was allowed to explore and interact with the stranger mouse for 2 mins in each trial. A second stranger mouse was introduced to the testing mouse during the 5th trial for 2 mins. The resting interval between trials was 5 mins. Active social behavior such as physical contacts (e.g. crawling over, social grooming), nose-to-nose sniffing, nose-to-anus from the testing mouse to the stranger mice were scored manually. Testing mice that showed aggressive behavior were withdrawn from the experiment, and the stranger mice showed aggressive behavior were removed from the testing trial and the pool of stranger mice.

Time-restricted feeding (TRF) paradigm

The TRF was conducted as previously described (Wang et al 2018; Whittaker et al 2018; 2023). WT and Fmr1 KO mice were first entrained to a 12:12 LD (300 lx vs 0 lx respectively) for a minimum of 2 weeks, then were exposed to a feeding paradigm of food (standard chow) available for 6 hr during the middle of the active phase from ZT 15 to ZT 21. Separate cohorts of WT and mutant mice were held on ad libitum feeding (ALF). To monitor the sleep/wake cycles, mice were singly housed in cages with an infrared motion sensor. Scheduled feeding was achieved by manually adding and removing food from the mouse cages, and a careful examination was carried out to ensure no small food fragments were dropped and remained in the cages. Food consumption was manually measured by weighing food at the beginning and the end of the feeding cycles (control (ALF): 24 hr vs TRF: 6 hr). The experimental groups and their ALF controls were held in these conditions for a total of 2 weeks.

Injection, Visualization and Analyses of the Neuroanatomical tracer Cholera Toxin

WT and Fmr1 KO mice (4 months-old) received a bilateral injection of CholeraToxin (β subunit) conjugated to Alexa Fluor™ 555 Conjugate (catalog number: C34776; Invitrogen™, Carlsbad, CA). Prior to the injections, the animals received a drop of a local ophthalmic anesthetic (Proparacaine HCl, 0.5%, Sandoz, Holzkirchen Germany) and an intraperitoneal (i.p.) injection of a non-steroidal anti-inflammatory drug (Carprofen, Zoetis, Parsippany-Troy Hills, NJ). The mice were then anesthetized with isoflurane and a 30G needle (BD PrecisionGlideTM Needle; Becton Dickinson, Franklin Lakes NJ) was inserted at a 45° angle into the sclera and the vitreous chamber, towards the base of the retina to allow leakage of vitreous humor. The Cholera Toxin (2 μg in 2μl of sterile PBS) was injected into the vitreous chamber using a 32G Hamilton syringe (Hamilton, Reno NV). The needle was left in place for ten seconds before being retracted. Seventy-two hours after the injection, the mice were euthanized with isoflurane (30%–32%) and transcardially perfused with phosphate-buffered saline (PBS, 0.1 M, pH 7.4) containing 4% (w/v) paraformaldehyde (PFA, Sigma). The brains were rapidly dissected out, post-fixed overnight in 4% PFA at 4°C, and cryoprotected in 15% sucrose. Sequential coronal sections (40-50 μm) containing the left and right SCN were mounted, and the cover-slips applied with a drop of Vectashield-containing DAPI (4’,6-diamidino-2-phenylinodole; catalog number: H-1200; Vector Laboratories, Burlingame, CA). Sections were visualized on a Zeiss AxioImager M2 microscope equipped with an AxioCam MRm and the ApoTome imaging system, and images acquired with the Zeiss Zen software and a 10x objective to include both left and right SCN. Two methods of analyses were carried out on the images of 5 consecutive sections per animal containing the middle SCN. First, the relative intensity of the Cholera Toxin fluorescent processes was quantified in the whole SCN, both left and right separately, by scanning densitometry using the Fiji image processing package of the NIH ImageJ software (https://imagej.net). A single ROI of fixed size (575.99 μm x 399.9 μm, width x height) was used to measure the relative integrated density (mean gray values x area of the ROI) in all the images. The values from the left and right SCN were averaged per section and 5 sections per animal were averaged to obtain one value per animal. Second, the distribution of the Cholera Toxin fluorescent signal was obtained for each left and right SCN separately in the same 4-5 consecutive sections per animal using the Profile Plot Analysis feature of ImageJ. Briefly, a rectangular box of fixed size (415.38μm x 110.94 μm, width x height, Suppl Fig.1) to include the ventral part of the SCN was set for each side, and a column plot profile was generated whereby the x-axis represents the horizontal distance through the SCN (lateral to medial for the left and medial to lateral for the right; Suppl Fig.1) and the y-axis represents the average pixel intensity per vertical line within the rectangular box. Subsequent processing of the resulting profiles was performed for left and right SCN images separately. To average the profiles of the 5 sections and obtain a single curve per animal, fifth-order polynomial curves were fit to best estimate the position of the intensity peak on the x-axis and, using this position, the original y-axis values were aligned and averaged arithmetically [1 profile per section (either left or right), 5 sections per animal]. Data are shown as the average profile ± SD of 3 animals per genotype.

Photic induction of cFos in the SCN and cFos-positive Cell Counting

A separate cohort of male WT and Fmr1 KO mice (3–4 months-old) were housed in DD conditions and exposed to light (300 lx, 4500K, 15 min) at CT 16. Forty-five mins later, the mice were euthanized with isoflurane (30%–32%) and transcardially perfused with phosphate-buffered saline (PBS, 0.1 M, pH 7.4) containing 4% (w/v) PFA. The brains were rapidly dissected out, post-fixed overnight in 4% PFA at 4°C, and cryoprotected in 15% sucrose until further processing. Sequential coronal sections (40-50 μm), containing the middle SCN, were collected on a cryostat (Leica, Buffalo Grove, IL) and further processed for cFos immunofluorescence as previously described (Wang et al., 2017; 2022; Longcore et al., 2024). Briefly, free-floating coronal sections, paired on the rostral-caudal axis and containing both the left and right SCN, were blocked for 1 h at room temperature (1% Bovine Serum Albumin, 0.3% Triton X-100, 10% normal donkey serum in 1xPBS) and then incubated overnight at 4°C with a rabbit polyclonal antiserum against cFos (1:1000, RRID: AB_2247211; Cell Signaling) followed by a Cy3-conjugated donkey-anti-rabbit secondary antibody (Jackson ImmunoResearch Laboratories, Bar Harbor, ME). Sections were mounted and coverslips applied with Vectashield mounting medium containing DAPI, and visualized on a Zeiss AxioImager M2 microscope (Zeiss, Thornwood NY) equipped with a motorized stage, an AxioCam MRm and the ApoTome imaging system. Z-Stack Images (35 images; 34mm, 1.029mm interval) of both the left and right middle SCN were acquired with a 20X objective using the Zeiss Zen digital imaging software, and two observers masked to the experimental groups performed the cell counting. The boundaries of the SCN were visualized using the DAPI nuclear staining and the cells immuno-positive for cFos counted with the aid of the Zen software tool ‘marker’ in three to five consecutive sections. The numbers obtained from the left and right SCN were averaged to obtain one value per section, and those from three-five sections averaged to obtain one value per animal and are presented as the mean ± standard deviation (SD) of 4 animals per genotype.

Histomorphometrical analyses of the SCN

Photographs of DAPI-stained sections generated from the WT and Fmr1 KO mice as described above were used to estimate the area, the perimeter, height and width of the SCN as previously reported (Li et al., 2015; Lee et al., 2018). For each animal, the four measurements were performed in three consecutive sections containing the middle SCN and acquired with a 10X objective and the Zen software. Measurements (in μm) of both the left and right SCN were obtained with the auxilium of the AxioVision software (Zeiss, Pleasanton, CA, USA). Because the borders of the DAPI-defined SCN are somewhat arbitrary, measurements were performed independently by two observers masked to the genotype of the animals. The area of the SCN in the three sections was summed, whilst the perimeter, height and width were averaged to obtain one value per side. No significant differences were found between the left and right SCN, therefore the values of the left and right SCN were averaged to obtain one value per animal. Data are shown as the mean ± standard deviation (SD) of 6 animals per genotype.

Blood sampling and measurements of plasma immune molecules

In the beginning of the light phase, blood was collected (∼0.5 mL per animal) via cheek puncture into microvette tubes coated with EDTA (Sarstedt, Numbrecht, Germany). Tubes were gently inverted a few times and placed immediately on wet ice. Within 1 h following collection, samples were centrifuged at 2500 rpm for 15 min at 4° C. The plasma was then collected into prelabeled Eppendorf tubes (Fisher Scientific, Hampton, NH), the lids sealed with parafilm, and immediately stored at -80° C until further processing using the Luminex Multiplexed Assay at UCLA (https://www.uclahealth.org/pathology/services-immunoassays).

Statistics

Data analyses were performed using SigmaPlot 14.5 (Grafiti LLC, Palo Alto, CA) or Prism 10 (GraphPad Software, La Jolla, CA). The impact of loss of Fmrp on the waveforms of sleep/wake cycles was analyzed using repeated measures two-way analysis of variance (ANOVA) with time and genotype factors. While a two-way ANOVA with genotype and treatment as factors was used for the differences in sleep bouts between light phase and dark phase. The Holm-Sidak’s multiple comparisons test was applied after the two-way ANOVA. The datasets were examined for normality (Shapiro–Wilk test) and equal variance (Brown– Forsythe test); genotypic differences in the behavioral tests were determined by Student t-test. Correlations between circadian/sleep parameters and other behaviors were examined by applying the Pearson correlation analysis. Normal distribution of the histomorphological datasets and the relative intensity of the Cholera Toxin fluorescent processes in the whole SCN was assessed using the Shapiro-Wilk test. Since the samples did not pass the normality test a two-tailed Mann Whitney test was employed to identify significant differences between groups. The effect of the loss of FMRP on the photic induction of cFos cells was assessed by one-way ANOVA followed by Bonferroni’s multiple comparisons test. Values are reported as the mean ± standard error of the mean (SEM) or mean ± standard deviation (SD). Differences were determined significant if P < 0.05.

eLife Assessment

This study presents valuable findings by using Fmr1 knockout mice as a model to investigate the role of Fmr1 in sleep regulation. These mice exhibited clear evidence of sleep and circadian disturbances, including abnormal retinal innervation of the SCN, which may provide a potential mechanistic explanation for the observed behavioral deficits. Interestingly, the results suggest that a scheduled feeding approach could improve sleep and circadian rhythms while enhancing social interactions and reducing repetitive behaviors in a mouse model of Fragile X syndrome. The topic is both intriguing and highly significant; however, while the evidence supporting the authors' claims is solid, several issues hinder the manuscript's clarity and impact.

Reviewer #1 (Public review):

Summary:

The authors investigated sleep and circadian rhythm disturbances in Fmr1 KO mice. Initially, they monitored daily home cage behaviors to assess sleep and circadian disruptions. Next, they examined the adaptability of circadian rhythms in response to photic suppression and skeleton photic periods. To explore the underlying mechanisms, they traced retino-suprachiasmatic connectivity. The authors further analyzed the social behaviors of Fmr1 KO mice and tested whether a scheduled feeding strategy could mitigate sleep, circadian, and social behavior deficits. Finally, they demonstrated that scheduled feeding corrected cytokine levels in the plasma of mutant mice.

Strengths:

(1) The manuscript addresses an important topic-investigating sleep deficits in an FXS mouse model and proposing a potential therapeutic strategy.

(2) The study includes a comprehensive experimental design with multiple methodologies, which adds depth to the investigation.

Weaknesses:

(1) The first serious issue in the manuscript is the lack of a clear description of how they performed the experiments and the missing definitions of various parameters in the results. Given that monitoring and analyzing sleep behaviors are the key experiments of this manuscript, I use the "Immobility-Based Sleep Behavior" section of Methods as an example to elaborate:

Incomplete or Incorrect Description of Tracking Threshold:

o The phrase "tracked the (40 sec or greater as previously described" is incomplete and does not clarify what is being tracked. This appears to be an error in writing or editing.

Unclear Relationship Between Threshold and EEG Validation:

o The threshold "40 sec or greater" is mentioned without context or explanation of what it represents (e.g., sleep bout duration, inactivity, or another parameter). The reference to Fisher et al. (2016) and "99% correlation with EEG-defined sleep" seems misaligned with the paragraph's content.

Confusing Definition of Sleep Bout:

o The definition of a sleep bout is unclear. Sleep bouts should logically be based on periods of inactivity, not activity. The sentence suggesting sleep is measured by "activity staying above the threshold" is confusing. The phrase "3 counts of sleep per minute for longer than one minute" requires clarification.

Unclear Data Selection for Analysis:

o The phrase "2 days with the best recording quality" is vague and does not specify how "best" was determined or why only two days out of five were analyzed.

Awkward Grammar and Structure:

o Phrases like "Acquiring data were exported in 1-min bins" are grammatically awkward. "Acquiring" should be "Acquired." Some sentences are overly long and lack clarity, making the text harder to follow.

In addition to this section, the authors should review all paragraphs in the Methods section to improve readability.

(2) Although the manuscript has a relatively long Methods section, some essential information is missing. For instance, the definition of sleep bout, as described above, is unclear. Additional missing information includes:

Figure 2: "Rhythmic strength (%)" and "Cycle-to-cycle variability (min)."

Figure 3: "Activity suppression."

Figure 4: "Rhythmic power (V%)" (is this different from rhythmic strength (%)?) and "Subjective day activity (%)."

Figure 5: Clear labeling of the SCN's anatomical features and an explanation for quantifying only the ventral part instead of the entire SCN. Alternatively, the authors should consider quantifying the whole SCN.

Figure 6: Inconsistencies in terms like "Sleep frag. (bout #)" and "Sleep bouts (#)." Consistent terminology throughout the manuscript is essential.

(3) Figure 1A shows higher mouse activity during ZT13-16. It is unclear why the authors scheduled feeding during ZT15-21, as this seems to disturb the rhythm. Consistent with this, the body weights of WT and Fmr1 KO mice decreased after scheduled feeding. The authors should explain the rationale for this design clearly.

(4) The interpretation of social behavior results in Figure 6 is questionable. The authors claim that Fmr1 KO mice cannot remember the first stranger in a three-chamber test, writing, "The reduced time in exploring and staying in the novel-mouse chamber suggested that the Fmr1 KO mutants were not able to distinguish the second novel mouse from the first now-familiar mouse." However, an alternative explanation is that Fmr1 KO mice do remember the first stranger but prefer to interact with it due to autistic-like tendencies. Data in Table 5 show that Fmr1 KO mice spent more time interacting with the first stranger in the 3-chamber social recognition test, which support this possibility. Similarly, in the five-trial social test, Fmr1 KO mice's preference for familiar mice might explain the reduced interaction with the second stranger.

In Figure 6C (five-trial social test results), only the fifth trial results are shown. Data for trials 1-4 should be provided and compared with the fifth trial. The behavioral features of mice in the 5-trial test can then be shown completely. In addition, the total interaction times for trials 1-4 (154 {plus minus} 15.3 for WT and 150 {plus minus} 20.9 for Fmr1 KO) suggest normal sociability in Fmr1 KO mice (it is different from the results of 3-chamber). Thus, individual data for trials 1-4 are required to draw reliable conclusions.

In Table 6 and Figure 6G-6J, the authors claim that "Sleep duration (Figures 6G, H) and fragmentation (Figures 6I, J) exhibited a moderate-strong correlation with both social recognition and grooming." However, Figure 6I shows a p-value of 0.077, which is not significant. Moreover, Table 6 shows no significant correlation between SNPI of the three-chamber social test and any sleep parameters. These data do not support the authors' conclusions.

(5) Figure 7 demonstrates the effect of scheduled feeding on circadian activity and sleep behaviors, representing another critical set of results in the manuscript. Notably, the WT+ALF and Fmr1 KO+ALF groups in Figure 7 underwent the same handling as the WT and Fmr1 KO groups in Figures 1 and 2, as no special treatments were applied to these mice. However, the daily patterns observed in Figures 7A, 7B, 7F, and 7G differ substantially from those shown in Figures 2B and 1A, respectively. Additionally, it is unclear why the WT+ALF and Fmr1 KO+ALF groups did not exhibit differences in Figures 7I and 7J, especially considering that Fmr1 KO mice displayed more sleep bouts but shorter bout lengths in Figures 1C and 1D.

Furthermore, it is not specified whether the results in Figure 7 were collected after two weeks of scheduled feeding (for how many days?) or if they represent the average data from the two-week treatment period.

The rationale behind analyzing "ZT 0-3 activity" in Figure 7D instead of the parameters shown in Figures 2C and 2D is also unclear.

In Figure 7F, some data points appear to be incorrectly plotted. For instance, the dark blue circle at ZT13 connects to the light blue circle at ZT14 and the dark blue circle at ZT17. This is inconsistent, as the dark blue circle at ZT13 should link to the dark blue circle at ZT14. Similarly, it is perplexing that the dark blue circle at ZT16 connects to both the light blue and dark blue circles at ZT17. Such errors undermine confidence in the data. The authors need to provide a clear explanation of how these data were processed.

Lastly, in the Figure 7 legend, Table 6 is cited; however, this appears to be incorrect. It seems the authors intended to refer to Table 7.

(6) Similar to the issue in Figure 7F, the data for day 12 in Supplemental Figure 2 includes two yellow triangles but lacks a green triangle. It is unclear how the authors constructed this chart, and clarification is needed.

(7) In Figure 8, a 5-trial test was used to assess the effect of scheduled feeding on social behaviors. It is essential to present the results for all trials (1 to 4). Additionally, it is unclear whether the results for familial mice in Figure 8A correspond to trials 1, 2, 3, or 4.

The legend for Figure 8 also appears to be incorrect: "The left panels show the time spent in social interactions when the second novel stranger mouse was introduced to the testing mouse in the 5-trial social interaction test. The significant differences were analyzed by two-way ANOVA followed by Holm-Sidak's multiple comparisons test with feeding treatment and genotype as factors." This description does not align with the content of the left panels. Moreover, two-way ANOVA is not the appropriate statistical analysis for Figure 8A. The authors need to provide accurate details about the analysis and revise the figure legend accordingly.

(8) The circadian activity and sleep behaviors of Fmr1 KO mice have been reported previously, with some findings consistent with the current manuscript, while others contradict it. Although the authors acknowledge this discrepancy, it seems insufficiently thorough to simply state that the reasons for the conflicts are unknown. Did the studies use the same equipment for behavior recording? Were the same parameters used to define locomotor activity and sleep behaviors? The authors are encouraged to investigate these details further, as doing so may uncover something interesting or significant.

(9) Some subtitles in the Results section and the figure legends do not align well with the presented data. For example, in the section titled "Reduced rhythmic strength and nocturnality in the Fmr1 KOs," it is unclear how the authors justify the claim of altered nocturnality in Fmr1 KO mice. How do the authors define changes in nocturnality? Additionally, the tense used in the subtitles and figure legends is incorrect. The authors are encouraged to carefully review all subtitles and figure legends to correct these errors and enhance readability.

Reviewer #2 (Public review):

Summary:

In the present study, the authors, using a mouse model of Fragile X syndrome, explore the very interesting hypothesis that restricting food access over a daily schedule will improve sleep patterns and, subsequently, behavioral capacities. By restricting food access from 12h to 6h over the nocturnal period (active period for mice), they show, in these KO mice, an improvement of the sleep pattern accompanied by reduced systemic levels of inflammatory markers and improved behavior. Using a classical mouse model of neurodevelopmental disorder (NDD), these data suggest that eating patterns might improve sleep quality, reduce inflammation and improve cognitive/behavioral capacities in children with NDD.

Strengths:

Overall, the paper is very well-written and easy to follow. The rationale of the study is generally well-introduced. The data are globally sound. The provided data support the interpretation overall.

Weaknesses:

(1) The introduction part is quite long in the Abstract, leaving limited space for the data provided by the present study.

(2) A couple of points are not totally clear for a non-expert reader:

- The Fmr1/Fxr2 double KO mice are not well described.

- What is the rationale for performing both LD and DD measures?

(3) The data on cytokines and chemokines are interesting. However, the rationale for the selection of these molecules is not given. In addition, these measures have been performed in the systemic blood. Measures in the brain could be very informative.

(4) An important question is the potential impact of fasting vs the impact of the food availability restriction. Indeed fasting has several effects on brain functioning including cognitive functions.

(5) How do the authors envision the potential translation of the present study to human patients? How to translate the 12 to 6 hours of food access in mice to children with Fragile X syndrome?

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Summary:

The authors investigated sleep and circadian rhythm disturbances in Fmr1 KO mice. Initially, they monitored daily home cage behaviors to assess sleep and circadian disruptions. Next, they examined the adaptability of circadian rhythms in response to photic suppression and skeleton photic periods. To explore the underlying mechanisms, they traced retino-suprachiasmatic connectivity. The authors further analyzed the social behaviors of Fmr1 KO mice and tested whether a scheduled feeding strategy could mitigate sleep, circadian, and social behavior deficits. Finally, they demonstrated that scheduled feeding corrected cytokine levels in the plasma of mutant mice.

Strengths:

(1) The manuscript addresses an important topic-investigating sleep deficits in an FXS mouse model and proposing a potential therapeutic strategy.

(2) The study includes a comprehensive experimental design with multiple methodologies, which adds depth to the investigation.

We thank the reviewer for the positive comments.

Weaknesses:

(1) The first serious issue in the manuscript is the lack of a clear description of how they performed the experiments and the missing definitions of various parameters in the results.

We thank the reviewer for pointing out lapses in the editing of the manuscript. We were trying to keep the descriptions of previously published methods brief but must have gone too far, the manuscript has been carefully checked for grammar and readability. Description of the experimental design has been refined and a graphical presentation has been added as Suppl Fig 3. The sleep and circadian parameters have been thoroughly explained in the methods and briefly in the figure legnds.

(2) Although the manuscript has a relatively long Methods section, some essential information is missing. For instance, the definition of sleep bout, as described above, is unclear. Additional missing information includes

Figure 2: "Rhythmic strength (%)" and "Cycle-to-cycle variability (min)."

Figure 3: "Activity suppression."

Figure 4: "Rhythmic power (V%)" (is this different from rhythmic strength (%)?) and "Subjective day activity (%)."

We have provided definitions for the general audience of the terms used in the field of circadian rhythms, such as sleep bout, rhythm power, cycle-to-cycle, masking, and % of activity during the day in the methods and Fig legends. Most of the techniques used in this study, for example, the behavioral measurement of sleep or locomotor activity, are well established and have been used in multiple published works, including our own. We have made sure to include citations for interested readers.

Figure 5: Clear labeling of the SCN's anatomical features and an explanation for quantifying only the ventral part instead of the entire SCN.

We have added more landmarks (position of the third ventricle and optic chiasm) to Fig 5, and have outlined the shell and core of the SCN in two additional images of the ventral hypothalamus in Suppl fig 4.

We had actually quantified the fluorescence in the whole SCN as well as in the ventral part.This was/is described in the methods as well as reported in the results section and Table 4 “Likewise, a subtle decrease in the intensity of the labelled fibers was found in the whole SCN (Table 4) of the Fmr1 KO mice as compared to WT.“

Methods: ” Two methods of analyses were carried out on the images of 5 consecutive sections per animal containing the middle SCN. First, the relative intensity of the Cholera Toxin fluorescent processes was quantified in the whole SCN, both left and right separately, by scanning densitometry using the Fiji image processing package of the NIH ImageJ software (https://imagej.net). A single ROI of fixed size (575.99 μm x 399.9 μm, width x height) was used to measure the relative integrated density (mean gray values x area of the ROI) in all the images. The values from the left and right SCN were averaged per section and 5 sections per animal were averaged to obtain one value per animal………..”

Since the retinal innervation of the SCN is strongest in the ventral aspect, where the retino-hypothalamic fibers reach the SCN and our goal was to identify differences in the input to the SCN, e.g. defects in the retino-SCN connectivity as suggested by some deficits in circadian behaviour; we also looked at intensity of Cholera Toxin in the fibers arriving to the ventral SCN from the retina.

We have added a sentence in the methods about the rationale for measuring the intensity of the cholera toxin labelled fiber in the whole SCN and also just in the ventral part: “Second, the retinal innervation of the SCN is strongest in the ventral aspect, where the retino-hypothalamic fibers reach the SCN, hence, the distribution….”

Figure 6: Inconsistencies in terms like "Sleep frag. (bout #)" and "Sleep bouts (#)." Consistent terminology throughout the manuscript is essential.

We have now clearly explained that sleep bouts are a measure of sleep fragmentation throughout the manuscript and in the fig legends; in addition, we have corrected the figures, reconciled the terminology, which is now consistent throughout the results and methods.

Methods: “Sleep fragmentation was determined by the number of sleep bouts, which were operationally defined as episodes of continuous immobility with a sleep count greater than 3 per minute, persisting for at least 60 secs.”

(3) Figure 1A shows higher mouse activity during ZT13-16. It is unclear why the authors scheduled feeding during ZT15- 21, as this seems to disturb the rhythm. Consistent with this, the body weights of WT and Fmr1 KO mice decreased after scheduled feeding. The authors should explain the rationale for this design clearly.

We have added to the rationale for the feeding schedule. This protocol was initially used by the Panda group to counter metabolic dysfunction (Hatori et al., 2012). We have used it for many years now (see citations below) in various mouse models presenting with circadian disruption to reset the clock and improve sleep. This study represents our first application/intervention in a mouse model of a neurodevelopmental disease.

Hatori M, Vollmers C, Zarrinpar A, DiTacchio L, Bushong EA, Gill S, Leblanc M, Chaix A, Joens M, Fitzpatrick JA, Ellisman MH, Panda S. Time-restricted feeding without reducing caloric intake prevents metabolic diseases in mice fed a high-fat diet. Cell Metab. 2012 Jun 6;15(6):848-60. doi: 10.1016/j.cmet.2012.04.019. Epub 2012 May 17. PMID: 22608008; PMCID: PMC3491655.

Chiem E, Zhao K, Dell'Angelica D, Ghiani CA, Paul KN, Colwell CS. Scheduled feeding improves sleep in a mouse model of Huntington's disease. Front Neurosci. 2024 18:1427125. doi: 10.3389/fnins.2024.1427125. PMID: 39161652.

Whittaker DS, Akhmetova L, Carlin D, Romero H, Welsh DK, Colwell CS, Desplats P. Circadian modulation by time-restricted feeding rescues brain pathology and improves memory in mouse models of Alzheimer's disease. Cell Metab. 2023 35(10):1704- 1721.e6. doi: 10.1016/j.cmet.2023.07.014. PMID: 37607543

Brown MR, Sen SK, Mazzone A, Her TK, Xiong Y, Lee JH, Javeed N, Colwell CS, Rakshit K, LeBrasseur NK, Gaspar-Maia A, Ordog T, Matveyenko AV. Time-restricted feeding prevents deleterious metabolic effects of circadian disruption through epigenetic control of β cell function. Sci Adv. 2021 7(51):eabg6856. doi: 10.1126/sciadv.abg6856. PMID: 34910509

Whittaker DS, Loh DH, Wang HB, Tahara Y, Kuljis D, Cutler T, Ghiani CA, Shibata S, Block GD, Colwell CS. Circadian-based Treatment Strategy Effective in the BACHD Mouse Model of Huntington's Disease. J Biol Rhythms. 2018 33(5):535-554. doi: 10.1177/0748730418790401. PMID: 30084274.

Wang HB, Loh DH, Whittaker DS, Cutler T, Howland D, Colwell CS. Time-Restricted Feeding Improves Circadian Dysfunction as well as Motor Symptoms in the Q175 Mouse Model of Huntington's Disease. eNeuro. 2018 Jan 3;5(1):ENEURO.0431-17.2017. doi: 10.1523/ENEURO.0431-17.2017.

Loh DH, Jami SA, Flores RE, Truong D, Ghiani CA, O'Dell TJ, Colwell CS. Misaligned feeding impairs memories. Elife. 2015 4:e09460. doi: 10.7554/eLife.09460.

(4) The interpretation of social behavior results in Figure 6 is questionable. The authors claim that Fmr1 KO mice cannot remember the first stranger in a three-chamber test, writing, "The reduced time in exploring and staying in the novelmouse chamber suggested that the Fmr1 KO mutants were not able to distinguish the second novel mouse from the first now-familiar mouse." However, an alternative explanation is that Fmr1 KO mice do remember the first stranger but prefer to interact with it due to autistic-like tendencies. Data in Table 5 show that Fmr1 KO mice spent more time interacting with the first stranger in the 3-chamber social recognition test, which support this possibility. Similarly, in the five-trial social test, Fmr1 KO mice's preference for familiar mice might explain the reduced interaction with the second stranger.

Thank you for this interesting interpretation of the social behavior experiments. We used the common interpretations for both the three-chamber test and the 5-trial social interaction test, but have now modified the text leaving space for alternative interpretations, have soften the language, and mentioned decreased sociability in the Fmr1 KO mice. “The reduced time spent exploring the novel-mouse chamber suggest that the mutants were, perhaps, unable to distinguish the second novel mouse from the first, now familiar, mouse, along with decreased sociability.”

In Figure 6C (five-trial social test results), only the fifth trial results are shown. Data for trials 1-4 should be provided and compared with the fifth trial. The behavioral features of mice in the 5-trial test can then be shown completely. In addition, the total interaction times for trials 1-4 (154 {plus minus} 15.3 for WT and 150 {plus minus} 20.9 for Fmr1 KO) suggest normal sociability in Fmr1 KO mice (it is different from the results of 3-chamber). Thus, individual data for trials 1-4 are required to draw reliable conclusions.

We have added a suppl figure showing the individual trial results for both WT and Fmr1 KO mice as requested (Suppl. Fig. 2).

In Table 6 and Figure 6G-6J, the authors claim that "Sleep duration (Figures 6G, H) and fragmentation (Figures 6I, J) exhibited a moderate-strong correlation with both social recognition and grooming." However, Figure 6I shows a p-value of 0.077, which is not significant. Moreover, Table 6 shows no significant correlation between SNPI of the three-chamber social test and any sleep parameters. These data do not support the authors' conclusions.

Thanks for pointing out the error with statement about Fig. 6I.

“…. Sleep duration (Fig. 6G, H; Table 6) exhibited a moderate to strong correlation with both social recognition and grooming time, while sleep fragmentation (measured by sleep bouts number) only correlated with the latter (Fig. 6J); the length of sleep bouts (Table 6) showed moderate correlation with both social recognition and repetitive behavior. In addition, a moderate correlation was seen between grooming time and the circadian parameters, rhythmic power and activity onset variability (Table 6). In short, our work suggests that even when tested during their circadian active phase, the Fmr1 KO mice exhibit robust repetitive and social behavioral deficits. Moreover, the shorter and more fragmented the daytime sleep, the more severe the behavioral impairment in the mutants.”

(5) Figure 7 demonstrates the effect of scheduled feeding on circadian activity and sleep behaviors, representing another critical set of results in the manuscript. Notably, the WT+ALF and Fmr1 KO+ALF groups in Figure 7 underwent the same handling as the WT and Fmr1 KO groups in Figures 1 and 2, as no special treatments were applied to these mice. However, the daily patterns observed in Figures 7A, 7B, 7F, and 7G differ substantially from those shown in Figures 2B and 1A, respectively. Additionally, it is unclear why the WT+ALF and Fmr1 KO+ALF groups did not exhibit differences in Figures 7I and 7J, especially considering that Fmr1 KO mice displayed more sleep bouts but shorter bout lengths in Figures 1C and 1D.

We appreciate the reviewer’s attention to the subtle details of the behavioral measurement of sleep and believe the reviewer to be referring to differences in the behavioral measurements of sleep with data shown in Table 1 and Table 7. The first set of experiments described in this study was carried out between 2016 and 2017 and involves the comparison between WT and Fmr1 KO mice. The WT and mutants were obtained from JAX. In this initial set of experiments (Table 1), the total amount of sleep in 24 hrs was reduced in the KO, albeit not significantly, and these also exhibited sleep bouts of significantly reduced duration. The pandemic forced us to greatly slow down the research and reduce our mouse colonies. Post-pandemic, we used new cohorts of Fmr1 KO ordered again from JAX for the TRF experiment presented in this study. In these cohorts, the KO mice exhibited a significant reduction in total sleep (Table 7) and the sleep bouts were still shorter but not significantly. We have added to our text to explain that the description of the mutants and TRF interventions were carried out at different times (2017 vs 2022). We would like to emphasize that we always run contemporaneously controls and experimental groups to be used for the statistical analyses. We believe that the data are remarkably consistent over these years, even with different students doing the measurements.

Furthermore, it is not specified whether the results in Figure 7 were collected after two weeks of scheduled feeding (for how many days?) or if they represent the average data from the two-week treatment period.

This is another good point raised by the reviewer. The activity measurements are collected during the 2 weeks (14 days) then the TRF was extended for a 3 more days to allow the behavioral sleep measurements.

We have added a supplementary figure (Supp Fig 3) depicting the different experimental designs.

The rationale behind analyzing "ZT 0-3 activity" in Figure 7D instead of the parameters shown in Figures 2C and 2D is also unclear.

We have added to our explanation. In prior work, we found that the TRF protocol has a big impact on the beginning of the sleep time, hence, we specifically targeted this 3-hours interval in the analysis.

In Figure 7F, some data points appear to be incorrectly plotted. For instance, the dark blue circle at ZT13 connects to the light blue circle at ZT14 and the dark blue circle at ZT17. This is inconsistent, as the dark blue circle at ZT13 should link to the dark blue circle at ZT14. Similarly, it is perplexing that the dark blue circle at ZT16 connects to both the light blue and dark blue circles at ZT17. Such errors undermine confidence in the data. The authors need to provide a clear explanation of how these data were processed.

Thank you for bringing this to our attention. The data were plotted correctly, however, those data points completely overlapped with those behind, masking them. We have now offset a bit them for clarity.

Lastly, in the Figure 7 legend, Table 6 is cited; however, this appears to be incorrect. It seems the authors intended to refer to Table 7.

We have corrected this error, thank you.

(6) Similar to the issue in Figure 7F, the data for day 12 in Supplemental Figure 2 includes two yellow triangles but lacks a green triangle. It is unclear how the authors constructed this chart, and clarification is needed.

We have corrected this error. As the reviewer pointed out, we filled the triangle on day 12 with yellow instead of green.

(7) In Figure 8, a 5-trial test was used to assess the effect of scheduled feeding on social behaviors. It is essential to present the results for all trials (1 to 4). Additionally, it is unclear whether the results for familial mice in Figure 8A correspond to trials 1, 2, 3, or 4.

The legend for Figure 8 also appears to be incorrect: "The left panels show the time spent in social interactions when the second novel stranger mouse was introduced to the testing mouse in the 5-trial social interaction test. The significant differences were analyzed by two-way ANOVA followed by Holm-Sidak's multiple comparisons test with feeding treatment and genotype as factors." This description does not align with the content of the left panels. Moreover, two-way ANOVA is not the appropriate statistical analysis for Figure 8A. The authors need to provide accurate details about the analysis and revise the figure legend accordingly.

We apologies for the confusing Figure legend which has been revised:

“Fig. 8: TRF improved social memory and stereotypic grooming behavior in the Fmr1 KO mice. (A) Social memory was evaluated with the 5-trial social interaction test as described above. The social memory recognition was significantly augmented in the Fmr1 KO by the intervention, suggesting that the treated mutants were able to distinguish the novel mouse from the familiar mouse. The time spent in social interactions with the novel mouse in the 5th-trial was increased to WT-like levels in the mutants on TRF. Paired t-tests were used to evaluate significant differences in the time spent interacting with the test mouse in the 4th (familiar mouse) and 5th (novel mouse) trials. *P < 0.05 indicates the significant time spent with the novel mouse compared to the familiar mouse. (B) Grooming was assessed in a novel arena in mice of each genotype (WT, Fmr1 KO) under each feeding condition and the resulting data analyzed by two-way ANOVA followed by the Holm-Sidak’s multiple comparisons test with feeding regimen and genotype as factors. *P < 0.05 indicates the significant difference within genotype - between diet regimens , and #P < 0.05 those between genotypes - same feeding regimen. (C) TRF did not alter the overall locomotion in the treated mice. See Table 8.”

To assess social recognition memory, mice underwent a five-trial social interaction paradigm in a neutral open-field arena. Each trial lasted 5 minutes and was separated by a 1-minute inter-trial interval. During trials 1–4, the test mouse was exposed to the same conspecific (Stimulus A) enclosed within a wire cup to permit olfactory and limited tactile interaction. In trial 5, a novel conspecific (Stimulus B) was introduced. Time spent investigating the stimulus B mouse (defined as sniffing or directing the nose toward the enclosure within close proximity) was scored using AnyMaze software. A progressive decrease in investigation time across trials 1–4 reflects habituation, while a significant increase in trial 5 indicates dishabituation and intact social recognition memory. In our data, there was not a lot of habituation in both genotypes, but clear differences can be appreciated between trial 4 with the now familiar mouse and trial 5 with novel mouse. Fig. 8A plots the results from individual animals in Trial 4 with a familiar mouse and in Trial 5 with a novel mouse, we have well specified this in the legends. As such, these data were analyzed with a pair t-test.

We used Tow-Way ANOVA to analyse the data reported in Panel 8B and as well as the results in Table 8. This has been clarified in the legend.

(8) The circadian activity and sleep behaviors of Fmr1 KO mice have been reported previously, with some findings consistent with the current manuscript, while others contradict it. Although the authors acknowledge this discrepancy, it seems insufficiently thorough to simply state that the reasons for the conflicts are unknown. Did the studies use the same equipment for behavior recording? Were the same parameters used to define locomotor activity and sleep behaviors? The authors are encouraged to investigate these details further, as doing so may uncover something interesting or significant.

We agree with the reviewers, and believe that the main differences were likely in the experimental design and possibly interpretation.

(9) Some subtitles in the Results section and the figure legends do not align well with the presented data. For example, in the section titled "Reduced rhythmic strength and nocturnality in the Fmr1 KOs," it is unclear how the authors justify the claim of altered nocturnality in Fmr1 KO mice. How do the authors define changes in nocturnality? Additionally, the tense used in the subtitles and figure legends is incorrect. The authors are encouraged to carefully review all subtitles and figure legends to correct these errors and enhance readability.

Nocturnality is defined as the % of total activity within a 24-h cycle that occurred in the night, since this can be confusing and we agree that it was not well explained we have removed it from the subtitle/figure legends.

We have adjusted the subtitles as recommended; however, the tense of the verbs might be a matter of writing style.

Reviewer #2 (Public review):

Summary:

In the present study, the authors, using a mouse model of Fragile X syndrome, explore the very interesting hypothesis that restricting food access over a daily schedule will improve sleep patterns and, subsequently, behavioral capacities. By restricting food access from 12h to 6h over the nocturnal period (active period for mice), they show, in these KO mice, an improvement of the sleep pattern accompanied by reduced systemic levels of inflammatory markers and improved behavior. Using a classical mouse model of neurodevelopmental disorder (NDD), these data suggest that eating patterns might improve sleep quality, reduce inflammation and improve cognitive/behavioral capacities in children with NDD.

Strengths:

Overall, the paper is very well-written and easy to follow. The rationale of the study is generally well-introduced. The data are globally sound. The provided data support the interpretation overall.

Thank you for the positive comments.

Weaknesses:

(1) The introduction part is quite long in the Abstract, leaving limited space for the data provided by the present study.

We have revised the Abstract to better focus on the most impactful findings as suggested.

(2) A couple of points are not totally clear for a non-expert reader: - The Fmr1/Fxr2 double KO mice are not well described. What is the rationale for performing both LD and DD measures?

We did not use the Fmr1/Fxr2 double KO mice in this study.

While measurement of day/night differences in activity rhythms are standardly done in a light/dark (LD) cycle, the organisms must be under constant conditions (DD) to measure their endogenous circadian rhythms (free running activity); this is often needed to uncover a compromised clock as entrainment to the LD cycle can mask deficits in the endogenous circadian rhythms.

(3) The data on cytokines and chemokines are interesting. However, the rationale for the selection of these molecules is not given. In addition, these measures have been performed in the systemic blood. Measures in the brain could be very informative.

The panel that we used had 16 cytokines/chemokines which are reported in Table 9. The experiment included WT and mutants held under 2 different feeding conditions with an n=8 per group. If we are able to obtain more resources, we would like to also carry out a comprehensive investigation of immunomediator levels as well as RNA-seq or Nanostring in selected brain regions associated with ASD aberrant behavioural phenotypes, for instance the prefrontal cortex.

(4) An important question is the potential impact of fasting vs the impact of the food availability restriction. Indeed, fasting has several effects on brain functioning including cognitive functions.

We did not address this issue in the present study. Briefly, the distinction between caloric restriction (CR) and TRF, in which no calories are restricted, has important mechanistic implications in mouse models. While both interventions can impact metabolism, circadian rhythms, and aging, they operate via overlapping but distinct molecular pathways. These have been the topic of recent reviews and investigations. Importantly, the fast-feed cycle can also act as a circadian entrainer (Zeitgeber)

Ribas-Latre A, Fernández-Veledo S, Vendrell J. Time-restricted eating, the clock ticking behind the scenes. Front Pharmacol. 2024 Aug 8;15:1428601. doi: 10.3389/fphar.2024.1428601. PMID: 39175542; PMCID: PMC11338815.

Wang R, Liao Y, Deng Y, Shuang R. Unraveling the Health Benefits and Mechanisms of Time-Restricted Feeding: Beyond Caloric Restriction. Nutr Rev. 2025 Mar 1;83(3):e1209-e1224. doi: 10.1093/nutrit/nuae074.

(5) How do the authors envision the potential translation of the present study to human patients? How to translate the 12 to 6 hours of food access in mice to children with Fragile X syndrome?

Time-restricted feeding (TRF) is a type of intermittent fasting that limits food intake to a specific window of time each day (usually 8–12 hours in humans), is being actively studied in adults for benefits on metabolic health, sleep, and circadian rhythms. However, applying TRF to children is not currently recommended as a general intervention, and there are important developmental, medical, and ethical considerations to take into account.

On the other hand, we believe that the Fmr1 KO mouse is a good preclinical model for FXS because it closely recapitulates key molecular, cellular, and behavioral phenotypes observed in humans with the disorder. A number of the behavioral phenotypes seen in the mouse mirror those seen in patients including increased anxiety-like behavior, sensory hypersensitivity, social interaction deficits and repetitive behaviors so there is strong face validity.

As we show in this study, Fmr1 KO mice present with disrupted sleep/wake cycles and reduced amplitude of circadian rhythms, consistent with findings in individuals with FXS. This makes the Fmr1 KO an excellent model to test out circadian based interventions such as scheduled feeding.

We believe that pre-clinical research in Fmr1 KO mice bridges the gap between basic discovery and human clinical application. It provides a controlled, cost-effective, and biologically relevant platform for understanding disease mechanisms and testing interventions. These types of experiments need to be done before jumping to humans to ensure that the human trials are scientifically justified and ethically sound.

Reviewer #1 (Recommendations for the authors):

The authors should:

(1) Revise the Methods section for clarity and completeness.

We have re-worked the methods for clarity and completeness.

(2) Provide consistent and precise definitions for all parameters and terms.

We believe that we have provided definitions for all terms.

(3) Clarify the rationale for experimental designs, such as the feeding schedule.

We have added to the rationale for the feeding schedule. This feeding schedule has been used in a number of prior studies including our own. All this work is cited in the manuscript.

(4) Reanalyze and transparently present data, including individual trial results.

We have added to the figure showing the individual trail results for the 5-trial tests as requested (Supplementary Fig. 2).

(5) Conduct appropriate statistical tests and correct figure legends.

We believe that we have carried out appropriate statistical tests and have carefully rechecked the figure legends.

(6) Investigate discrepancies with prior studies to enhance the discussion.

We have added to our discussion of prior work.

(7) Improve language quality and ensure consistency in terminology and grammar.

We have edited the manuscript to improve language quality.

Reviewer #2 (Recommendations for the authors):

(1) The Abstract should be rewritten to provide more room for the obtained data.

We have re-written the Abstract to focus on the most impactful findings.

(2) An additional sentence describing the double KO mice should be added.

We did not use double KO mice in this study.

(3) The rationale for studying LD and DD should be provided.

Measurement of day/night differences are standardly done in a light/dark cycle. To measure the endogenous circadian rhythms, the organisms must be under constant conditions (Dark/Dark).

(4) The data on cytokines/chemokines should be strengthened by performing a larger panel of measures both in blood and the brain.

The panel that we used had 16 cytokines/chemokines which we report in Table 9. This was a large experiment with 2 genotypes being held under 2 feeding conditions with n=8 mice per group. If we are able to obtain more resources, we would like to also carry out RNA-seq in different brain regions.

(5) The authors should discuss in more detail the potential role of fastening vs restriction of food access.

We did not address this issue in the present study. Briefly, the distinction between caloric restriction (CR) and TRF when no calories are restricted has important mechanistic implications in mouse models. While both interventions can impact metabolism, circadian rhythms, and aging, they operate via overlapping but distinct molecular pathways.

(6) The authors should also provide some insight into their view on the potential translation of their experimental studies.

We believe that the Fmr1 KO mouse is considered a good preclinical model for FXS because it closely recapitulates key molecular, cellular, and behavioral phenotypes observed in humans with the disorder. A number of the behavioral phenotypes seen in the mouse mirror those seen in patients including increased anxiety-like behavior, sensory hypersensitivity, social interaction deficits and repetitive behaviors so there is strong face validity. As we demonstrate in this study, Fmr1 KO mice exibit disrupted sleep/wake cycles and reduced amplitude of circadian rhythms, consistent with findings in individuals with FXS. This makes the Fmr1 KO an excellent model to test out circadian based interventions such as scheduled feeding.

Still we are mindful that the translation of therapeutic findings from mouse to human has proven challenging e.g., mGluR5 antagonists failed in clinical trials despite strong preclinical data (Berry-Kravis et al., 2016). Therefore, we are cautious in overreaching in our translational interpretations.

Berry-Kravis, E., Des Portes, V., Hagerman, R., Jacquemont, S., Charles, P., Visootsak, J., Brinkman, M., Rerat, K., Koumaras, B., Zhu, L., Barth, G. M., Jaecklin, T., Apostol, G., & von Raison, F. (2016). Mavoglurant in fragile X syndrome: Results of two randomized, double-blind, placebo-controlled trials. Science translational medicine, 8(321), 321ra5. https://doi.org/10.1126/scitranslmed.aab4109).

Version 2

Abstract

Abstract

Fragile X syndrome (FXS), a leading inherited cause of intellectual disability and autism, is frequently accompanied by sleep and circadian rhythm disturbances. In this study, we comprehensively characterized these disruptions and evaluated the therapeutic potential of a circadian-based intervention in the fragile X mental retardation 1 (FMR1) knockout (KO) mouse. The Fmr1 KO mice exhibited fragmented sleep, impaired locomotor rhythmicity, and attenuated behavioral responses to light, linked to an abnormal retinal innervation and reduction of light-evoked neuronal activation in the suprachiasmatic nucleus. Behavioral testing revealed significant deficits in social memory and increased repetitive behaviors in the mutants, which correlated with sleep fragmentation. Remarkably, a scheduled feeding paradigm (6-hour feeding/18-hour fasting) significantly enhanced circadian rhythmicity, consolidated sleep, and improved social deficits and repetitive behaviors in the Fmr1 KO mice. This intervention also normalized the elevated levels of some pro-inflammatory cytokines, including IL-12 and IFN-γ, in the mutants’ blood, suggesting that its benefits extend to inflammatory pathways. These findings highlight the interplay between circadian disruption, behavior, and an inflammatory response in FXS, and provide compelling evidence that time-restricted feeding may serve as a promising non- pharmacological approach for improving core symptoms in neurodevelopmental disorders.

Introduction

Fragile X syndrome (FXS) is a relatively common inherited cause of intellectual disability, with a prevalence of approximately 1 in 4,000 males and 1 in 8,000 females (Hersh & Saul, 2011). Males are typically more severely affected due to their single X chromosome. FXS is caused by an abnormal expansion of CGG trinucleotide repeats in the fragile X mental retardation 1 (FMR1) gene, which results in the transcriptional silencing of FMR1 gene and consequent reduction in the levels of the fragile X mental retardation protein (FMRP), an RNA-binding protein essential for synaptic plasticity and neuronal cytoarchitecture (Zalfa et al., 2006; Soden et al., 2010; Kute et al., 2019). There is currently no cure for FXS. Existing treatments are primarily symptomatic, targeting common manifestations such as social deficits, repetitive behaviors, and sleep disturbances (Budimirovic et al., 2022; Mendez & Mendez, 2024). Given the essential role of sleep and circadian rhythms in neural recovery and plasticity, a compelling question is whether improving sleep/wake cycles might alleviate other core symptoms of this disease.

Animal models are invaluable for elucidating the mechanisms underlying circadian and sleep disturbances as well as developing interventions aimed at restoring healthy sleep/wake cycles. Among the available models for FXS, the Fmr1 knockout (KO) mouse has been extensively validated for investigating disease mechanisms and is widely used in preclinical drug development, including for sleep-related outcomes (Thomas et al., 2012; Kazdoba et al., 2014; Kat et al., 2022; Saré et al., 2022; Martinez et al., 2024). Early studies have shown disrupted rhythmic activity in the light/dark (LD) cycle in the Fmr1/Fxr2 double KO and Fmr1 KO/Fxr2 heterozygous mice (Zhang et al., 2008), whilst more recent findings indicate impairments in both sleep (Saré et al., 2017) and activity rhythms (Bonasera et al., 2017; Angelakos et al., 2019) in Fmr1 KO mice, all consistent with the sleep/wake disturbances reported in FXS individuals (Kronk et al., 2009, 2010; Budimirovic et al., 2022).

Despite these findings, significant gaps remain in our understanding of the precise nature of the sleep and circadian deficits in Fmr1 KO mice, and whether targeted correction of these rhythms might mitigate behavioral symptoms. Furthermore, much of the behavioral research in Fmr1 KO mice has been conducted during the light phase—when mice are typically asleep—potentially confounding results due to sleep disruption.

In this study, we aimed at comprehensively characterizing sleep and circadian disturbances in the Fmr1 KO mice and examine their association with behavioral alterations. To avoid disrupting natural sleep patterns, we employed non-invasive home-cage monitoring systems to assess sleep and activity rhythms and conducted behavioral testing during the active (dark) phase. Importantly, the same animals were evaluated for both sleep/wake cycles and behavioral phenotypes, enabling direct correlations. Finally, because of the above mentioned evidence of circadian dysfunction, we tested the efficacy of a circadian-based intervention— scheduled feeding (6-hour feeding/18-hour fasting)—on improving sleep/wake rhythms, social behavior, and repetitive behaviors in these mutants.

Results

Fragmented Light-Phase Sleep in the Fmr1 KO Mice

To determine whether the Fmr1 KO exhibit deficits in immobility-defined sleep behavior, the animals were examined using a combination of video recording and mouse tracking system under a 12h:12h light/dark (LD) condition. Both the WT and mutant mice exhibited robust circadian rhythms in sleep with higher levels during the day and minimal sleep during the early dark phase (Fig. 1A). The amount of total sleep within the 24 hr cycle and the number of sleep bouts did not differ between genotypes, although the Fmr1 KO did exhibit a reduction in the average duration of each sleep bout (Table 1). A two-way ANOVA used to analyze the temporal pattern of sleep (1-hr bins) in WT and mutants revealed significant effects of time (F(23, 287) = 31.94; P < 0.001) and genotype (F(23, 287) = 11.95; P < 0.001), along with differences in day and night sleep parameters (Table 1). During the resting phase (lights on), the Fmr1 KO mutants slept significantly less (Fig. 1B) with more frequent sleep bouts (Fig. 1C) of shorter duration (Fig. 1D). The maximum sleep bout duration was about 20 minutes shorter in the mutant mice. The reduced sleep time, shorter bout length, and greater number of sleep bouts seen during the day are all evidence of a fragmented sleep pattern in the Fmr1 KO mutants. The negative findings should be viewed with caution as the sample size of n=6 per group is below ideal.

The Fmr1 KO mice exhibit reduced and fragmented sleep during the light phase.

(A) Waveforms of daily rhythms in sleep behavior under standard 12:12 h light-dark (LD) cycles in WT (blue circle) and Fmr1 KO (yellow triangle) mice. Zeitgeber time (ZT) 0 corresponds to lights on, and ZT 12 to lights off. Sleep was defined by immobility and binned in 1-hr intervals. Both genotypes exhibited clear diurnal rhythms, but the Fmr1 KO mice showed significantly less sleep during the light phase and at several times in the dark phase. Asterisks indicate significant differences between genotypes at individual time points (two-way ANOVA with genotype and time as factors, followed by the Holm-Sidak’s multiple comparisons test, *P < 0.05). The white/black bar on the top indicates the LD cycle; gray shading indicates the dark phase time-period. (B–D) Immobility-defined sleep metrics during the light-phase. Compared to WT, Fmr1 KO mice showed a greater number of sleep bouts, indicating sleep fragmentation, and of shorter duration. Histograms show the means ± SEM with the values from individual animals overlaid. Genotypic differences were analyzed by t-test (*P < 0.05). See Table 1.

Disrupted diurnal and circadian activity rhythms in the Fmr1 KO Mice

Next, we investigated the presence of deficits in locomotor activity rhythms in the Fmr1 KO mice. Age- matched WT and Fmr1 KO were housed in cages equipped with running-wheels under 12h:12h LD cycles for 2 weeks and then released into constant darkness (DD) to measure their endogenous rhythms (Suppl. Fig. 1). Both genotypes exhibited robust daily and circadian rhythms in wheel running activity (Fig. 2A; Table 2). Some indication of the commonly reported hyper-activity in the Fmr1 KO mice (Sørensen et al., 2015) emerged in the LD cycle (Fig. 2B), although the activity levels over 24 hours did not differ significantly (Fig. 2C) between genotypes. Analysis of the diurnal activity patterns with two-way ANOVA confirmed significant effects of time (F(23, 287) = 8.84; P = 0.003) and genotype (F(23, 287) = 39.75; P < 0.001) on the locomotor activity. In LD (Fig. 2C), the mutants exhibited lower rhythm power with increased activity during lights-on and more variability in the activity onset. When released in DD (Fig. 2D), WT and mutants displayed very similar free-running period (tau), but again, the latter presented with weaker rhythm power and a higher cycle-to-cycle variability in activity onset (Table 2). Together, our data suggest that the Fmr1 KO mice exhibit deficits in the circadian regulation of locomotor activity and raise the possibility of impaired ability to synchronize to the light cycle.

The Fmr1 KO mice show unstable locomotor rhythms.

(A) Representative actograms showing daily rhythms in wheel-running activity under LD followed by constant darkness (DD) in WT (left) and Fmr1 KO (right) mice. Activity levels were normalized to 85% of the most active mouse. Each row represents two consecutive days, and the second day is repeated at the beginning of the next row. (B) Waveforms of daily rhythms in cage activity in WT (blue circle) and Fmr1 KO (yellow triangle) mice under the LD cycles. Activity under LD (1-hr bins) was analyzed by two-way ANOVA with genotype and time as factors followed by the Holm-Sidak’s multiple comparisons test (*P < 0.05). There were significant effects of both time (F = 8.84; P = 0.003) and genotype (F = 39.75; P < 0.001) on the temporal pattern of the locomotor activity rhythms. Note that significant genotypic differences were found before and after dawn. Activity metrics in LD (C) and DD (D). Rhythmic power under both LD and DD conditions was significantly reduced in the mutants, which also presented higher imprecision in the activity onset in DD. Histograms show the means ± SEM with the values from individual animals overlaid. Genotypic differences were analyzed by t-test (*P < 0.05). See Table 2.

Impaired Behavioral Response to Photic Cues in the Fmr1 KO Mice

The likelihood of the Fmr1 KO mice presenting deficits in their response to light was tested using four behavioral assays of the retinal input to the circadian system. First, the ability of exposure to light (4500K, 50lx) during the dark phase to suppress or “mask” the locomotor activity of the nocturnal animals was evaluated. For this assay, the level of locomotor activity between Zeitgeber Time (ZT) 14 and 15 (two hours after lights off) was first measured under baseline LD conditions (i.e. the day before), and no differences were observed between genotypes (WT: 2229 ± 162 revolutions per hour (rev/h); KO: 2639 ± 328 rev/h; t(18) = -1.123, P = 0.277). The next day the mice were exposed to light for 1 hour at this same phase and the levels of activity measured. The mutants exhibited a significantly reduced light-driven suppression of the locomotor activity as compared to WT, (Fig. 3A, B; Table 3), with significant effects of both genotype (F(1, 39) = 12.002; P = 0.001, Two-way ANOVA) and light exposure (F(1, 39) = 24.253; P < 0.001, Two-way ANOVA). Next, we determined the number of cycles needed by the mice to re-entrain to a 6-hours phase advance of the LD cycle. While the WT mice re-synchronized in 5.9 ± 0.6 days to the new phase, the Fmr1 KO mice required 11.3 ± 0.4 days to adjust to the new ZT 12 (Fig. 3C, D; Table 3). Third, we evaluated the ability of the mice to entrain to a skeleton photoperiod (SPP) in which the full 12 hours of light or dark are replaced by two 1-hour light exposures (ZT 0-1 & ZT 11-12) separated by 11 hours of dark (Fig. 4A). Mice were first entrained to the standard LD cycle and then released into the SPP for 2 weeks (Fig. 4A). The activity recordings showed that the WT stably entrained to this challenging environment and exhibited robust circadian rhythms with a period (tau) of 24.0 ± 0.0 hr (Table 3). Conversely, the Fmr1 KO exhibited a tau of 23.7 ± 0.2 hr with the shorter period driven by three mutants that failed to entrain to the SPP (Table 3). Compared to the WT, the mutants also showed reduced power of the rhythms, increased light-phase activity, and a larger onset variability (Fig. 4B). Finally, the direct light-induced phase shift of the circadian system of the mice was measured by exposing them to light (300 lx) for 15min while in constant darkness at circadian time (CT) 16. The WT showed a phase delay of 135.6 ± 26.9 min, whilst, in the Fmr1 KO mutants, the delay in the activity onset on the next day was about half the magnitude (64.0 ± 0.1 min; Fig. 4C, D; Table 3). These findings provide evidence for the presence of a defective circadian response to light in the Fmr1 KO mice.

The Fmr1 KO mice display deficits in light-regulated circadian behaviors.

(A, B) Photic- suppression (masking) of activity in response to a 1-hr light (300 lx, 4500K) pulse at ZT 14 (lights off; n=10/genotype). The activity level during the light exposure was compared to the activity level during the equivalent hour (ZT 14-15) on the day before the treatment (baseline activity). (A) The genotypic difference in the fold change was determined by t-test, with the mutants showing a significantly reduced suppression of activity as compared to the WT (*P = 0.05). (B) Changes in the activity levels of each individual mice during the baseline window and the light masking were analyzed using a paired t-test. Within-genotype comparisons showed significant suppression in WT (P < 0.001), but not in KO mice (P = 0.12). (C, D) Entrainment induced by a 6-hr-phase advanced LD cycle. Representative actograms of light-induced phase shifts of wheel-running activity rhythms (C). The white/black bars on the top of actograms indicate the LD cycle before (upper) and after (lower) the 6-hr-phase advance. The gray shading in the waveforms indicates the dark phase time-period. The arrows next to the actograms indicate the day when the 6 hours-phase advance was applied. (D) Quantification of the days to re-entrain show that the KO mice required more time to adjust (two-way ANOVA: genotype effect F(1,285) = 130.157, P < 0.001; followed by the Holm-Sidak’s multiple comparisons test *P < 0.001). The entrainment shifting in the WT (blue circle) and the Fmr1 KO (yellow triangle) was quantified by the difference between the activity onset and the new ZT12 on each day. The yellow and blue arrow heads in the graph indicate the day when the activity rhythms are considered well entrained. See Table 3.

The Fmr1 KO mice exhibit difficulty in adapting to the skeleton photic period (SPP).

(A) Representative actograms of daily rhythms in cage activity under standard LD cycles (2 weeks) followed by the SPP (1h light:11h dark:1h light:11h dark) in WT (left) and Fmr1 KO (right) mice. The white/black bars on the top of actograms indicate the baseline LD cycle (upper) and the SPP LD cycles (lower). The gray shading in the waveforms indicates the time of the dark phases. (B) Measures of locomotor activity rhythms under SPP. Many of the parameters measured were significantly different between the genotypes with the mutants being more impacted and showing lower rhythmic power and increased variability of the activity onset. In addition, the mutants displayed higher activity during their day. Histograms show the means ± SEM with the values from each individual animal overlaid. Significant differences (P < 0.05), determined by t-test or Mann-Whitney test, are indicated with an asterisk. (C, D) Light-induced phase delay of free-running activity rhythms in mice exposed to light (300 lx, 4500K) for 15 min at circadian time (CT) 16. Mice were held in constant darkness, by definition, CT 12 is the beginning of the activity cycle in DD for a nocturnal organism. Examples of light-induced phase shifts of wheel-running activity rhythms (C) of WT (left) and Fmr1 KO (right) and quantified phase delay (D). In the representative actograms, the yellow lines indicate the best-fit line of the activity onset across the 10 days before and after the light pulse. The amount of phase delay is determined by the difference between the two lines on the day after the light pulse. The sunny-shape symbols indicate when the mice were exposed to light (CT16). Compared to WT, the Fmr1 KO showed reduced phase shift of their activity rhythms (Mann Whitney U *P = 0.011). See Table 3.

Subtle deficits in the retinal afferent innervation to the SCN in the Fmr1 KO Mice

The results described above suggest that the Fmr1 KO mice present with an anomalous retinal input to the suprachiasmatic nucleus (SCN), the master circadian clock located in the anterior hypothalamus. Hence, to assess the integrity of this pathway, the WT and Fmr1 KO mice received a bilateral intravitreal injection of the fluorescence-conjugated neurotracer Cholera Toxin (β subunit), previously used to map the projections of the melanopsin-expressing intrinsically photoreceptive retinal ganglion cells (ipRGCs) from the retina to the SCN (Muscat et al., 2003; Hattar et al., 2006). At variance with the WT, a lower fluorescent signal could be observed in the mutant mice both laterally and medially to the ventral SCN (Fig. 5A), where the retino-hypothalamic fibers reach the nuclei, as well as beneath in the optic chiasm. Analysis of the intensity and distribution (Suppl. Fig. 2) of the labelled retino-hypothalamic processes reaching and entering the ventral SCN showed a reduction, particularly evident in the lateral side of both the left and right mutant SCN (Fig. 5B, C). Likewise, a subtle decrease in the intensity of the labelled fibers was found in the whole SCN (Table 4) of the Fmr1 KO mice as compared to WT.

Abnormal Retinal-Suprachiasmatic Nucleus connectivity in the Fmr1 KO mice.

To trace the projections from the retina to the suprachiasmatic nucleus (SCN) via the Retino-Hypothalamic tract (RHT), WT and Fmr1 KO mice received a bilateral intravitreal injection of Cholera Toxin (β-subunit) conjugated to Alexa Fluor555 and were perfused 72 hours later. (A) Lower intensity of the fluorescently labelled RHT projections can be observed both laterally and medially to the ventral part of the SCN in the Fmr1 KO mice (white arrows) as compared to WT, suggesting a loss of afferent projections to the SCN. (B, C) Densitometric analysis of the distribution of the cholera toxin fluorescence intensity in the ventral SCN (Suppl. Fig. 2) of WT and Fmr1 KO mice. The intensity peaks of the profile plot of 4 to 5 consecutive coronal sections containing the middle SCN were aligned and then averaged to obtain a single curve per animal. Results are shown as the mean ± standard deviation (SD) for the left (B) and the right (C) SCN of each genotype. (D, E) Light-induction of cFos was greatly reduced in the SCN of the Fmr1 KO mice compared to WT. Mice held in DD, were exposed to light (300 lx, 4500K) for 15 min at CT 16, and perfused 45 minutes later (CT 17). (D) Representative serial images of light-evoked cFos expression in the SCN. The inset in the lower left panel shows the lack of cFos immunopositive cells in the SCN of mice held in DD but not exposed to light. Dotted magenta lines delineate the SCN. OC=Optic Chiasm, V=third Ventricle, (E) The number of immune-positive cells in the left and right SCN from 3-5 consecutive coronal sections per animal were averaged to obtain one number per animal and are presented as the mean ± SD per genotype. One-way ANOVA followed by Bonferroni’s multiple comparisons test, *P=0.0201. See Table 4

A well-established test of the light input to the circadian system is the light-evoked cFos response in the SCN. In line with the results obtained with the Cholera Toxin suggesting an impaired light-pathway to the SCN in the mutants, the number of cFos positive cells in the SCN of the Fmr1 KO was greatly reduced (50%) in comparison to the WT mice (Fig. 5D, E; Table 4). Contrary to other models of neurodevelopmental disabilities (NDDs; Li et al., 2015; Lee et al., 2018), the Fmr1 KO mice did not display any histomorphometrical alteration of the SCN (Table 4). Based on these findings, we can surmise the presence of a compromised connectivity between the retina and the SCN in the mutants, which could provide, at least in part, a mechanism for their difficulty in responding to photic cues.

Social and Repetitive Behavior Deficits in Fmr1 KO Mice

Social deficits and stereotypic symptoms are hallmark problems in autism spectrum disorders, which we sought to evaluate during the active (dark) phase, between ZT 16-18, in the two genotypes. Social behavior was tested with the three-chamber and the five-trial social interaction tests (Fig. 6). In the first stage of the three- chambers test, both genotypes showed more interest toward the stranger mouse when given the choice between an inert object and a mouse. The direct comparison of the time spent with the object and the mouse did not show significant differences between the genotypes (Table 5). On the other hand, measurement of the social preference index (SPI) indicated that the Fmr1 KO mice exhibited a reduced interest in the conspecific compared to WT (Fig. 6A; Table 5). In the second testing stage, when the lifeless object was replaced with a second novel mouse and the first stranger became the familiar mouse (Fig. 6B), the mutants spent more time with the familiar mouse than the WT (Table 5). Again, the social novelty preference index (SNPI) confirmed that the Fmr1 KO exhibited reduced interest in novel mouse compared to WT mice (Fig. 6B; Table 5). The reduced time spent exploring the novel-mouse chamber suggest that the mutants were, perhaps, unable to distinguish the second novel mouse from the first, now familiar, mouse, along with decreased sociability. The possibility of impaired social recognition was further tested with the five-trial social test (Fig. 6C). In this test, the first stranger mouse becomes the familiar mouse after 4 exposures to the testing mouse. When the second mouse is introduced in the 5th trial, the testing mouse typically shows a boosted interest in investigating the novel mouse. As expected, when the second stranger mouse was introduced to the testing animals in the 5th trial, the WT showed elevated social behavior; whilst the mutants did not display increased interest in exploring the second novel mouse (Fig. 6C; Table 5). No genotypic differences in social interaction in the first four trials were observed, indicating that habituation was unaltered in the mutants (Suppl. Fig. 3).

The deficits in social recognition and repetitive behaviors of the Fmr1 KO mice correlate with altered sleep behavior.

Tests of Social behavior: (A) In the first stage of the 3-chambers social test, when the testing mouse is given a choice between a stranger mouse and an inanimate object, the Fmr1 KO mice spent less time with the stranger mouse and had a lower social preference index (SPI) than the WT. (B) In the second stage, the testing mouse is given the choice between a chamber with a novel mouse and the one with the familiar mouse. The mutants spent less time with the novel mouse compared to the familiar one, and also in this phase exhibited lower social novelty preference (SNPI) as compared to the WT. (C) The possibility of reduced social recognition was further tested with the 5-trial social test. In this test, the stranger mouse becomes a familiar mouse after 4 exposures to the testing mouse, then a novel mouse is introduced in the 5th trial. The WT mice showed a higher interest for the novel mouse compared to the Fmr1 KO mice. Tests of repetitive behaviors: The amount of digging in the bedding (D) and the percentage of marbles buried (E) were measured with the marble bury test. The Fmr1 KO mice spent longer time digging and buried more marbles compared to WT. (F) Grooming behavior, assessed in a novel arena, was significantly higher in the Fmr1 KO mice as compared to WT. Histograms show the means ± SEM with the values from the individual animals overlaid. Significant differences (*P < 0.05) were determined by t-test or Mann-Whitney test. See also Table 5. Sleep duration (min, G, H) correlated with impaired social recognition and abnormal grooming behaviors. Sleep fragmentation, measured by number of sleep bouts, correlated only with grooming (I,J) (Pearson Correlation test). See Table 6.

Next, when the active-phase repetitive behavior was examined using the marble bury test, the Fmr1 KO spent more time digging and buried more marbles compared to WT (Fig. 6D, E; Table 5). When the 30-min trial was divided into three 10-min intervals, the repetitive digging behavior was significantly higher in the Fmr1 mutants compared with the WT in all three intervals. Two-way ANOVA demonstrated that there was a significant effect of genotype (F(1, 107) = 11.04; P = 0.001) but not of the interval (F(2, 107) = 0.42; P = 0.66). The Fmr1 KO mice also exhibited significantly more grooming behavior than their WT counterpart (Fig. 6F, Table 5).

Finally, we used these data to determine if disrupted sleep correlated with autistic-like behavioral deficits. Sleep duration (Fig. 6G, H; Table 6) exhibited a moderate to strong correlation with both social recognition and grooming time, while sleep fragmentation (measured by sleep bouts number) only correlated with the latter (Fig. 6J); the length of sleep bouts (Table 6) showed moderate correlation with both social recognition and repetitive behavior. In addition, a moderate correlation was seen between grooming time and the circadian parameters, rhythmic power and activity onset variability (Table 6). In short, our work suggests that even when tested during their circadian active phase, the Fmr1 KO mice exhibit robust repetitive and social behavioral deficits. Moreover, the shorter and more fragmented the daytime sleep, the more severe the behavioral impairment in the mutants.

Amelioration of the Behavioral Disruptions by Scheduled Feeding in the Fmr1 KO Mice

The correlations between fragmented sleep and the severity of other behavioral phenotypes support the possibility that interventions focused on improving circadian rhythms and sleep may benefit the observed deficits in the Fmr1 KO mice. Scheduled feeding can be a powerful regulator of circadian rhythms (Long and Satchidananda, 2022; Manoogian et al., 2022) and has been shown to be effective in several disease models (Whittaker et al., 2018; Wang et al., 2018; Gao et al., 2022; Gupta et al., 2022; Whittaker et al., 2023).To test whether circadian reinforcement could ameliorate the observed deficits in the mutants, a new cohort of WT and Fmr1 KO mice was subjected to a time-restricted feeding (TRF) protocol (6-h feeding/18-h fasting, Suppl. Fig. 1) for two weeks and compared to counterpart animals held on ad libitum feeding (ALF). Both genotypes were able to well adapt to the TRF regimen and ate as much as their ALF counterpart after 5 days; at the end of the two-weeks, the TRF groups weighed less than the ALF groups (Suppl. Fig. 4).

The TRF treatment benefited several aspects of the temporal patterning of activity and sleep (Fig. 7) in both genotypes. The scheduled feeding improved the power of the rhythms and markedly restored the cycle-to- cycle variability in the activity onset of the mutants to WT levels (Fig. 7C-E), while reducing fragmentation (less activity bouts) in both WT and KO (Table 7). Total cage activity was not altered by the intervention in either genotype. By the second week, TRF also improved sleep behavior in both genotypes (Fig. 7F, G), with the Fmr1 KO mice showing increased total daytime sleep (Fig. 7H; Table 7), reduced sleep fragmentation, i.e. fewer bouts (Fig. 7I) of longer bout durations (Fig. 7J) as compared to their ALF counterpart. In general, strong genotypic differences were present in the response to TRF with the WT mice showing a more robust sleep response (Table 7).

Amelioration of sleep/wake rhythms in the Fmr1 KO mutants by Time Restricted Feeding (TRF).

(A, B) Waveforms of daily rhythms in cage activity using IR detection in the WT (circle) and Fmr1 KO (triangle) mice under ad libitum feeding (ALF) or TRF. The activity waveforms (1-hr bins) were analyzed using a three- way ANOVA with genotype, feeding regimen, and time as factors followed by Holm-Sidak’s multiple comparisons test. There were significant effects of genotype (F(1, 767) = 13.301; P < 0.001) and time (F(23,767) = 94.188; P < 0.001), as well as significant interactions between genotype and time (P < 0.001) and feeding regimen and time (P < 0.001) on the locomotor activity rhythms of both WT and Fmr1 KO mice. The green area indicates the time-period when the food hoppers were opened for 6 hours between ZT 15 and ZT 21. (C- E) Measures of locomotor activity rhythms: Both genotypes exhibited an increase in the power of the rhythms under TRF compared to ALF controls. The increase in early-day and late-night activity as well as the onset variability seen in the Fmr1 KO mice was corrected by the TRF. Data are shown as the means ± SEM; two- way ANOVA followed by Holm-Sidak’s multiple comparisons test with genotype and feeding regimen as factors, *P < 0.05 significant differences within genotypes (different feeding regimen); #P < 0.05 significant differences between genotypes (same feeding regimen). (F, G) Waveforms of daily rhythms of immobility- defined sleep. The sleep waveforms (1-hr bins) were analyzed by two-way ANOVA with time and feeding regimen as factors followed by the Holm-Sidak’s multiple comparisons test. There were significant effects of time for both WT (F(23, 351) = 9.828, P <0.001) and Fmr1 KO (F(23, 351) = 1.806, P = 0.014) mice, but not of feeding regimens. Missing data points precluded the use of three-way ANOVA for these measures. (H-J) Measures of immobility-defined sleep in the light phase. Both genotypes held on TRF exhibited an increase in sleep duration and in sleep bout length as well as a reduction in sleep fragmentation, measured by the number of sleep bouts, compared to their ALF counterparts. Data are shown as the means ± SEM; two-way ANOVA followed by Holm-Sidak’s multiple comparisons test with genotype and diet as factors, *P < 0.05 significant differences within genotype - between diet regimens; #P < 0.05 significant differences between genotypes - same feeding regimen. See Table 7.

The benefits of TRF were also evident on other aberrant behaviors in the Fmr1 KO model, and after two weeks of treatment, this feeding regimen improved social memory recognition while reducing the inappropriate grooming behavior (Table 8). Scheduled feeding improved social memory in the KO mice, by restoring the preference for the novel over the familiar mice to WT-like level (Fig. 8A; Table 8), suggesting that, perhaps, the treated mutants were able to distinguish the novel mouse from the familiar mouse. Grooming behavior was also significantly reduced by TRF in the Fmr1 KO mice, which showed a significant reduction in the time spent self- grooming (Fig. 8B), without affecting their locomotor activity (Fig. 8C). These data suggest that TRF has ameliorating effects on sleep/wake rhythms as well as autistic-like behaviors in the Fmr1 KO model.

Improved social memory and stereotypic grooming behavior in the Fmr1 KO mice after 2 weeks of TRF.

(A) Social memory was evaluated with the 5-trial social interaction test as described above. The social memory recognition was significantly augmented in the Fmr1 KO by the intervention, suggesting that the treated mutants were, perhaps, able to distinguish the novel mouse from the familiar mouse. The time spent in social interactions with the novel mouse in the 5th-trial was increased to WT-like levels in the mutants on TRF. Paired t-tests were used to evaluate significant differences in the time spent interacting with the test mouse in the 4th (familiar mouse) and 5th (novel mouse) trials. *P < 0.05 indicates the significant time spent with the novel mouse compared to the familiar mouse. (B) Grooming was assessed in a novel arena in mice of each genotype (WT, Fmr1 KO) under each feeding condition and the resulting data analyzed by two-way ANOVA followed by the Holm-Sidak’s multiple comparisons test with feeding regimen and genotype as factors. *P < 0.05 indicates the significant difference within genotype - between diet regimens, and #P < 0.05 those between genotypes - same feeding regimen. (C) TRF did not alter the overall locomotion in the treated mice. See Table 8.

Attenuation of the Elevated Levels of Pro-inflammatory Markers in the Fmr1 KO Mice by TRF

An abnormal immune response has been suggested to play a role in FXS pathophysiology (Reynolds et al., 2021; Dias et al., 2023; Robinson-Agramonte et al., 2022). Hence, blood was collected at the end of the two weeks of the scheduled feeding regimen to investigate whether TRF would impact the profile of pro-inflammatory markers in the Fmr1 KO mice. The mutants on ALF displayed elevated levels of several cytokine in comparison with their WT counterpart (Table 9), in particular, Interleukin-12, (IL-12), Interferon-gamma (IFNƳ), and the Chemokine Ligand-9 (CXCL-9). These genotypic differences were abolished in the TRF-treated groups (Fig. 9A). Scheduled feeding also repristinated the Fmr1 KO levels of IL-2 to WT levels. Furthermore, elevated IL-12 and IFN-γ levels were significantly associated with markers of poor sleep, such as higher activity in the light phase, shorter sleep time/sleep bout length, and higher numbers of sleep bouts (Table 10), as well as with social memory deficits and excessive grooming (Fig. 9B, C; Table 10). These data suggest that TRF-dampening of the pro-inflammatory signaling in Fmr1 KO mice provides a potential mechanism to explain, at in least part, the observed behavioral improvements.

TRF repristinates the levels IL-12 and IFNƳ in the plasma of Fmr1 KO mice to WT levels.

(A) The levels of selected plasma pro-inflammatory markers are shown. The increase in IL-12 and IFNƳ in the mutants is lowered by TRF to WT levels. The full list of the assayed makers is reported in Table 9. Data were analyzed with two-way ANOVA followed by the Holm-Sidak’s multiple comparisons test with feeding regimen and genotype as factors. *P < 0.05 indicates the significant difference within genotype - between feeding regimen, and #P < 0.05 between genotypes – same feeding regimen. (B-G) Correlations between IL-12 or IFN-γ levels and social recognition or grooming behavior. IL-12 levels correlated with both behaviors. Data were analyzed using the Pearson Correlation, and the coefficients are reported in Table 10.

Discussion

Sleep and circadian rhythm disruptions are prevalent in individuals with NDDs, including FXS. Common issues include difficulty initiating and maintaining sleep and irregular sleep patterns, which significantly impair the quality of life of FXS individuals and their families/caretakers (Budimirovic et al., 2022; Minhas et al., 2024).

Notably, individuals exhibiting more severe behavioral symptoms often experience more pronounced sleep problems (Schreck et al., 2004; Taylor et al., 2012; Kaufmann et al., 2024), suggesting that circadian and sleep dysfunction exacerbate core NDD features. These disturbances are detectable quite early, even in infants and toddlers carrying the FMR1 mutation (D’Souza et al., 2020), raising the possibility that disrupted sleep/wake cycles may be an initial and influential feature of disease expression. Mechanistically, such disruptions could stem from direct effects of FMR1 loss on sleep and circadian circuits, or indirectly from secondary symptoms such as anxiety, sensory hypersensitivity, and repetitive behaviors.

Animal models of FXS, such as the Fmr1 KO mouse, are essential for dissecting the underlying mechanisms and testing interventions (Sandoval et al., 2024). In the present study, we demonstrate that loss of Fmr1 results in disrupted sleep and circadian rhythms, including higher fragmentation of the rest-phase sleep and abnormal diurnal activity patterns. These disruptions were associated with deficits in social memory and increased repetitive behaviors. Importantly, we found that a circadian-based intervention (TRF) not only improved rhythmicity and sleep consolidation but also ameliorated behavioral deficits and reduced peripheral inflammatory markers.

Circadian and Sleep Disturbances in Fmr1 KO Mice

Our behavioral sleep assay revealed reduced sleep duration and increased fragmentation during the light/rest phase in Fmr1 KO mice (Fig. 1), consistent with a previous report by Saré and colleagues (2017). In contrast, Westmark et al. (2023), using EEG, did not observe baseline sleep differences between genotypes, possibly reflecting methodological differences or subtle phenotypic variability. A key gap in the field is the lack of studies probing sleep homeostasis in Fmr1 KO mice; future studies using sleep deprivation protocols may reveal underlying vulnerabilities. Notably, the Fmr1 KO mice exhibit neural oscillation abnormalities, including increased resting gamma power and enhanced auditory-evoked responses, paralleling the EEG signatures in FXS patients (Jonak et al., 2024).

Assessment of circadian locomotor activity revealed weakened rhythmicity, elevated daytime activity, and a greater onset variability in Fmr1 KO mice (Fig. 2), consistent with prior studies (Bonasera et al., 2017; Angelakos et al., 2019). These deficits persisted in constant darkness (DD), indicating a compromised endogenous circadian clock. Differences in activity rhythms measured in LD conditions could be driven by the circadian clock or the direct effects of light and dark. Although earlier studies using Fmr1/Fxr2 double mutants reported severe rhythm loss (Zhang et al., 2008), the present study demonstrates that the loss of Fmr1 alone is sufficient to impair circadian regulation.

Impaired Light Response and SCN Connectivity in the Fmr1 KO Mice

To explore the basis of the observed dysregulations, we evaluated the photic input to the circadian system. Fmr1 KO mice displayed impaired negative masking, delayed re-entrainment to phase shifts, blunted light- induced phase shifts, and poor entrainment to a skeleton photoperiod (Figs. 3–4). These behavioral deficits suggest a compromised signal transmission from the retina to the SCN via the ipRGCs. Such hypothesis was confirmed by anatomical and functional assays; indeed, the ipRGC innervation of the SCN was lessened in the Fmr1 KO mice (Fig. 5A), and the light-evoked c-Fos induction substantially reduced (Fig. 5B). Given that the FMRP is expressed in the ipRGCs (Zhang et al., 2020), its loss may underline the deficient light response observed in the mutants. Interestingly, this reduced sensitivity to circadian light input contrasts with the general sensory hypersensitivity typical of FXS, highlighting a unique dissociation in sensory processing pathways (Rais et al., 2018). Whether these circadian-specific photic impairments are present in FXS patients warrants further investigation.

Despite these light input abnormalities, the Fmr1 KO mice did not exhibit a shortened free-running period (tau) in DD (Fig. 2D; Angelakos et al., 2019), nor were the rhythmic expressions of the clock genes altered (Zhang et al., 2008). These findings suggest that the core molecular clock remains intact, and the circadian dysfunction may arise from an altered excitatory/inhibitory (E/I) balance or impaired network synchronization within the SCN (Olde Engberink et al., 2023). Hence, our present results reinforce that loss of Fmr1 impairs the circadian system at the level of input and output regulation rather than at the core molecular oscillation.

Sleep and Behavior Correlate in the Fmr1 KO Mice

The Fmr1 KO model reproduces core behavioral deficits of FXS, including impaired social recognition and increased repetitive behaviors (Melancia & Trezza, 2018; Kat et al., 2022). By analyzing sleep, circadian activity, and behavior in the same animals, we identified significant correlations: mice with more disrupted sleep and weaker circadian rhythms exhibited more severe social and repetitive behavior impairments (Fig. 6). These findings parallel clinical studies showing that disrupted sleep predicts next-day behavioral difficulties in FXS and autism spectrum disorders (Cohen et al., 2018; Robinson-Shelton & Malow, 2016; Schwichtenberg et al., 2022; Minhas et al., 2024). Although correlational, these results suggest that targeting circadian and sleep dysfunction may offer a novel route for ameliorating some behavioral phenotypes present in certain NDDs.

Time-Restricted Feeding as a Therapeutic Strategy

Given the link between circadian disruption and behavior, we tested whether restoring rhythmicity could improve outcomes. TRF, a circadian-aligned feeding schedule, significantly improved the power of the rhythms, reduced cycle-to-cycle (onset) variability, and consolidated daytime sleep (Fig. 7). Importantly, TRF reduced repetitive behavior and improved social memory and interactions in the Fmr1 KO mice (Fig. 8), establishing proof-of-principle that circadian-based behavioral interventions can ameliorate core phenotypes of FXS.

Ketogenic diets have previously been shown to improve both sleep and seizure control in the Fmr1 KO mice, albeit with limited effects on other behaviors (Westmark et al., 2020, 2023, 2024). TRF, by contrast, appears to directly impact both behavioral and physiological endpoints, potentially through its entraining effects on peripheral and central clocks. Prior studies have shown that TRF improves behavior in aging and animal models of neurodegenerative disorders (Wang et al., 2018; Whittaker et al., 2018, 2023), but this is the first report to demonstrate benefits in a model of NDDs.

TRF Reduces Inflammatory Signatures in Fmr1 KO Mice

An emerging literature suggests that immune dysregulation accompanies FXS pathology (Ashwood et al., 2010; Careaga et al., 2014; Bertrand et al., 2021). In the present study, TRF reduced the abnormal levels of IL- 12 and IFN-γ in the Fmr1 KO mice (Fig. 9A; Table 9), and the increased levels of these cytokines correlated with both sleep disturbances and behavioral impairments (Fig. 9B, C; Table 10). While most cytokines did not differ between genotypes, the normalization of IL-12 and IFN-γ levels following TRF supports a mechanistic link between inflammation and behavioral symptoms in FXS. These results are consistent with prior evidence implicating IL-12 in social and repetitive behavior in autism spectrum disorders (Ashwood et al., 2011; Fallah et al., 2020) and suggest that circadian-based interventions may exert part of their benefit by dampening low- grade inflammation.

Whether TRF exerts direct anti-inflammatory effects in the central nervous system remains unknown. Future studies should assess microglia activation, blood–brain barrier integrity, and central cytokine levels following TRF in the Fmr1 KO and related models. Moreover, defining the causal role of specific cytokines such as IL-12 and IFN-γ in behavioral regulation may offer novel therapeutic targets for FXS and related disorders.

Methods

Animals and Time-restricted feeding (TRF) paradigm

All experimental procedures were approved by the UCLA Animal Research Committee and conformed to guidelines from the UCLA Division of Laboratory Animal Medicine (DLAM) and the National Institutes of Health (NIH). Wild-type (WT) (JAX ID: 000664) and Fmr1 KO mice (Fmr1tm4Cgr) on the C57BL/6J background (JAX ID: 003025) were acquired from the Jackson laboratory (Bar Harbor, ME). These mutants have a neomycin resistance cassette replacing exon 5 of the fragile X mental retardation syndrome 1 (Fmr1) gene. The studies were carried out in male adult mice (3 to 5 months old) singly housed in light-tight ventilated cabinets in temperature- and humidity-controlled conditions, with ad libitum access to food and water unless otherwise stated. Cages were equipped with either running wheel or infrared motion sensors to monitor the sleep/wake behavior.

TRF was conducted as previously described (Wang et al 2018; Whittaker et al 2018; 2023). Briefly, WT and Fmr1 KO mice were singly housed and entrained to a 12h:12h light/dark (LD) cycle (300 lx vs 0 lx respectively) for a minimum of 2 weeks, then randomly assigned to one of two feeding conditions: food (standard chow) available ad libitum (ad lib) or available for 6 hr during the middle of the active phase from Zeitgeber Time (ZT) 15 to ZT 21. By definition ZT0 and ZT12 are, respectively, the onsets of lights turning on and off. Scheduled feeding was achieved by manually adding and removing food from the mouse cages, and a careful examination was carried out to ensure no small food fragments were dropped and remained in the cages.

Food consumption was measured by weighing food at the beginning and the end of the feeding cycles (ALF: 24 hr vs TRF: 6 hr). The ALF and TRF mice groups were held in these conditions for at least two weeks (14 days) plus 3 additional days to measure immobility-based sleep behavior (Suppl. Fig. 1).

This study used two different cohorts of mice for a total of 135 mice. The experiments shown in Fig. 1 through 6 were designed to compare the circadian phenotype of the Fmr1 KO mice to the WT and were carried out between 2017-2020 before the Covid pandemic research shutdown. After the restart of research, new cohorts of mice were obtained from Jackson Laboratory, and used to test the effects of scheduled feeding on the mutants’ behavioral and circadian deficits (Fig. 7 through 9).

Immobility-Based Sleep Behavior

Sleep Behavior was assessed under a 12h:12h LD cycle by video recording in combination with an automated mouse tracking analysis software system (Anymaze, Stoelting Co., Wood Dale, IL) as previously described (Lee et al., 2018; Whittaker et al., 2018). Sleep was defined as the periods when 95% of the animal remained immobile for at least 40 seconds or longer, a threshold shown to correlate with >99% accuracy to EEG-defined sleep (Fisher et al., 2016). For each animal, continuous recordings were performed over 5 days and data from two consecutive days with the fewest artifact-free epochs were averaged and used for further analysis. Data were exported in 1-min bins and analyzed separately for the light and dark phases. Total sleep time was determined by summing the immobility durations in the rest phase (ZT 0-12) or active phase (ZT 12-24). Sleep fragmentation was determined by the number of sleep bouts, which were operationally defined as episodes of continuous immobility with a sleep count greater than 3 per minute, persisting for at least 60 secs.

Locomotor cage activity rhythms

Methods used to characterize the sleep/wake cycles have been previously described (Lee et al., 2018; Wang et al., 2018; Whittaker et al., 2018). Single housed WT and the Fmr1 KO mice were habituated to a 12h:12h LD cycle for two weeks to ensure entrainment to the appointed LD schedule before data collection.

For the phenotypical comparisons between WT and Fmr1 KO, cage activity was recorded with running wheel sensors for at least 14 days in LD, then, after the assessment of sleep behavior (3 days), the animals were released for two additional weeks into constant darkness (12h:12h dark-dark, DD) to obtain measures of the endogenous rhythms (free-running activity, Suppl. Fig. 1). To investigate the effects of TRF, cage activity rhythms were monitored using overhead passive infrared (IR) motion sensors for two weeks (Wang et al., 2018; Whittaker et al., 2018).

Activity data were collected via the VitalView data recording system (Mini Mitter, Bend, OR) and ten days of recordings were used to obtain the period (tau), the rhythmic power as well as the waveform presentations. The analysis was conducted using the El Temps (A. Diez-Nogura, Barcelona, Spain) and ClockLab (Actimetrics, Lafayette Instruments, Lafayette, IN) programs as previously described (Lee et al 2018; Wang et al., 2018; Whittaker et al., 2018). The periodogram generated by the El Temps uses χ2 test with a threshold of 0.001 significance, from these the amplitude of the periodicities is calculated at the circadian harmonic to obtain the power of the rhythmicity. The rhythmic power, or percentage of variation (%V), provides a measure of the strength of the mouse’s periodicity corrected for the activity amount and normalized to the percentage of variance derived from peak significance (P=0.05). The amount of cage activity over a 24hr period was averaged over 10 days and reported here as the arbitrary units (a.u.)/h. Fragmentation and imprecision of the daily onset of sleep/wake cycles were determined using ClockLab. Fragmentation was determined by the number of activity bouts per day, with one bout counted when activity was separated by a gap of 21 min or more (maximum gap: 21 min; threshold 3 counts/min). Imprecision, a measure of cycle-to- cycle variability or onset variability, was determined by calculating the daily variability in the time of activity onset from a best-fit regression line drawn through 10 days of activity in both LD and DD conditions using the ClockLab program (Actimetrics, Wilmette, IL, United States).

Photic Regulation of Circadian Behavior

Cohorts of age-matched WT and Fmr1 KO mice were single housed in cages with running wheels under a 12h:12h LD cycle (300 lx, 4500K) for two weeks and, after stable entrainment, exposed to four behavioral assays to test the photic regulation of their circadian system.

Photic suppression on nocturnal activity (negative light masking): mice were exposed to 1 hr of white light (300 lx, 4500K) at ZT 14 to determine their response to light exposure during the dark period (masking of nocturnal activity). The animals’ activity level, measured as the number of wheel revolutions, during the light pulse was compared to the level at the same hour (ZT 14 to 15) on the day before the exposure. Data are reported as percent of suppression= % [(rev during the 1-hr light exposure) – (rev during the dark baseline)]/(rev during the dark baseline).

Re-entrainment to a 6hr advance of the LD cycle: the mice underwent a 6-hr phase advance. The activity levels were continuously monitored by the wheel-running system and the activity onset determined with VitalView. The ability of the animals to re-entrain was quantified by the difference between the new ZT 12 (lights-off) and the onset of the running activity for each subsequent recording day. A mouse was considered fully re-entrained when the activity onset aligned to the new lights-off time for the consecutive 5 days.

Skeleton Photoperiod (SPP): the mice were transferred from the 12h:12h LD cycle to an SPP consisting of 1hL:11hD:1L:11hD cycles. Mice were exposed to the 1-hr light (300 lx) at the beginning (ZT 0) and the end (ZT 11) of the original light phase. The mice were kept in this photic condition for at least 2 weeks and activity was recorded and analyzed to determine rhythm power, period (tau) and other basic activity rhythm parameters using ClockLab.

Light-induced phase shift of activity rhythms: Mice were released into DD for 10 days. The circadian time (CT) of their free-running activity rhythms was determined using VitalView and El Temps software. The time of activity onset under DD was defined as CT 12. On day 11, the mice were exposed to light (300 lx, 4500K) at CT 16 for 15 min, and after the light pulse, they were held in DD for an additional 10 days. The best- fit lines of the activity onsets of sleep/wake cycles before and after light exposure were measured and compared.

Behavioral Assays

All these tests were conducted under dim red light (<2 lx) during the dark (active) phase between 2 and 4 h after lights- off (ZT 14–16) to avoid sleep disruptions after the sleep and activity recordings had been completed (Suppl. Fig. 1). Animals were habituated to the testing environment for at least 30 minutes prior to the behavioral assays.

The Stereotypic Behavior was assessed using two tests:

The marble burying test was used to evaluate repetitive digging behavior (Yrigollen et al., 2019). An array of 4 by 6 marbles was placed in the testing arena over a layer of 4-cm deep shavings. The testing mouse was introduced to the arena from the corner and its behavior recorded for 30 mins. After that, the testing mouse was carefully removed from the arena with special precaution to not move the marbles and returned to the home cages. The number of marbles buried more than ⅔ of their area was counted, and these were cleaned with 70% ethanol before the next use. Time spent on digging was manually scored by an observer masked to the experimental groups, whilst the distance travelled was derived from the automatic mouse- tracking system (Anymaze software, Stoelting Co., Wood Dale, IL).

The grooming test was conducted as in our previous work (Wang et al 2020). The mouse was introduced to the testing arena and its behavior recorded by a camcorder for 30 mins. The time spent on self- grooming, defined as the cleaning, licking, or washing of the limbs, tail, and body surface areas, typically from a head to tail direction, but excluding bouts of scratching, was manually scored by an observer masked to the experimental groups. The distance travelled was derived from the automatic mouse-tracking system (Anymaze software).

Social Behavior was assessed using two tests:

The three-chamber test was performed as previously described (Wang et al 2020). The testing mice were allowed to freely explore an arena with three chambers where the central chamber remained empty. When being habituated to the three-chamber arena, both the WT and the mutants explored the arena evenly with no preference toward the left or the right chamber (left/right ratio: WT: 0.95 ± 0.14; P = 0.43 by paired t- test. Fmr1 KO: 0.82 ± 0.1; P = 0.085 by paired t-test). The three-chamber test consisted of two parts: the first stage assessed the preference of social approach toward the stranger mouse, and the second stage assessed the ability of social discrimination of the testing mouse. In the first stage (social approach), an up-turned metal- grid pencil cup was placed in the side chambers: one was left empty as the novel object (the object chamber), while a never-met stranger mouse matched by sex, age, and genotype with the testing mouse was placed in the second up-turned cup (the social chamber). Thus, the testing mouse was tested for its preference between the object chamber and the social chamber in this first testing stage. In the second stage (social discrimination), the first stranger mouse and the cup remained the same, while a second stranger mouse matched by sex, age, and genotype with the testing mouse was placed in the second up-turned cup. In other words, the social chamber in the first stage became a familiar chamber in the second stage, and the object chamber of the first stage a novel chamber in the second stage. Thus, the preference between the familiar chamber and the novel chamber of the testing mouse was assessed in this second stage. Time spent in each chamber and the distance travelled were derived from the automatic mouse-tracking system (Anymaze software).

The 5-trial social test was conducted as previously described (Mineur et al., 2006). The testing mouse was first habituated to the arena for 30 mins and then introduced to a never-met stranger mouse for 4 trials. The testing mouse was allowed to explore and interact with the stranger mouse for 2 mins in each trial, then it was introduced to a second stranger mouse during the 5th trial for 2 mins. The resting interval between trials was 5 mins. Active social behaviors such as physical contacts (e.g. crawling over, social grooming), nose-to-nose sniffing, nose-to-anus from the testing mouse to the stranger mice were scored manually. Testing mice as well as stranger mice that showed aggressive behavior were withdrawn from the experiment.

Retinal-SCN Connectivity: Injection, Visualization and Analyses of the Neuroanatomical tracer Cholera Toxin

WT and Fmr1 KO mice (4 months-old) received a bilateral injection of CholeraToxin (β subunit) conjugated to Alexa Fluor™ 555 Conjugate (catalog number: C34776; Invitrogen™, Carlsbad, CA). Prior to the injections, the animals received a drop of a local ophthalmic anesthetic (Proparacaine HCl, 0.5%, Sandoz, Holzkirchen Germany) and an intraperitoneal (i.p.) injection of a non-steroidal anti-inflammatory drug (Carprofen, Zoetis, Parsippany-Troy Hills, NJ). The mice were then anesthetized with isoflurane and a 30G needle (BD PrecisionGlideTM Needle; Becton Dickinson, Franklin Lakes NJ) was inserted at a 45° angle into the sclera and the vitreous chamber, towards the base of the retina to allow leakage of vitreous humor. The Cholera Toxin (2 μg in 2μl of sterile PBS) was injected into the vitreous chamber using a 32G Hamilton syringe (Hamilton, Reno NV). The needle was left in place for ten seconds before being retracted. Seventy-two hours after the injection, the mice were euthanized with isoflurane (30%–32%) and transcardially perfused with phosphate-buffered saline (PBS, 0.1 M, pH 7.4) containing 4% (w/v) paraformaldehyde (PFA, Sigma). The brains were rapidly dissected out, post-fixed overnight in 4% PFA at 4°C, and cryoprotected in 15% sucrose.

Sequential coronal sections (40-50 μm) containing the left and right SCN were mounted, and the cover-slips applied with a drop of Vectashield-containing DAPI (4’,6-diamidino-2-phenylinodole; catalog number: H-1200; Vector Laboratories, Burlingame, CA). Sections were visualized on a Zeiss AxioImager M2 microscope equipped with an AxioCam MRm and the ApoTome imaging system, and images acquired with the Zeiss Zen software and a 10x objective to include both left and right SCN. Two methods of analyses were carried out on the images of 5 consecutive sections per animal containing the middle SCN (Lee et al., 2018). First, the relative intensity of the Cholera Toxin fluorescent processes was quantified in the whole SCN, both left and right separately, by scanning densitometry using the Fiji image processing package of the NIH ImageJ software (https://imagej.net). A single ROI of fixed size (575.99 μm x 399.9 μm, width x height) was used to measure the relative integrated density (mean gray values x area of the ROI) in all the images. The values from the left and right SCN were averaged per section and 5 sections per animal were averaged to obtain one value per animal. Second, the retinal innervation of the SCN is strongest in the ventral aspect, where the retino-hypothalamic fibers reach the nuclei, hence, the distribution of the Cholera Toxin fluorescent signal was also obtained for each left and right ventral SCN separately in the same 4-5 consecutive sections per animal using the Profile Plot Analysis feature of ImageJ (Lee et al., 2018). Briefly, a rectangular box of fixed size (415.38μm x 110.94 μm, width x height, Suppl. Fig. 2) to include the ventral part of the SCN was set for each side, and a column plot profile was generated whereby the x-axis represents the horizontal distance through the SCN (lateral to medial for the left and medial to lateral for the right; Suppl. Fig. 2) and the y-axis represents the average pixel intensity per vertical line within the rectangular box. Subsequent processing of the resulting profiles was performed for left and right SCN images separately. To average the profiles of the 5 sections and obtain a single curve per animal, fifth-order polynomial curves were fit to best estimate the position of the intensity peak on the x-axis and, using this position, the original y-axis values were aligned and averaged arithmetically [1 profile per section (either left or right), 5 sections per animal]. Analyses were performed by two observers masked to the genotype of the animals. Data are shown as the average profile ± SD of 3 animals per genotype.

Photic induction of cFos in the SCN and cFos-positive Cell Counting

A separate cohort of male WT and Fmr1 KO mice (3–4 months-old) was housed in DD conditions and exposed to light (300 lx, 4500K, 15 min) at CT 16. Forty-five mins later, the mice were euthanized with isoflurane (30%–32%) and transcardially perfused with phosphate-buffered saline (PBS, 0.1 M, pH 7.4) containing 4% (w/v) PFA. The brains were rapidly dissected out, post-fixed overnight in 4% PFA at 4°C, and cryoprotected in 15% sucrose until further processing. Sequential coronal sections (40-50 μm), containing the middle SCN, were collected on a cryostat (Leica, Buffalo Grove, IL) and further processed for cFos immunofluorescence as previously described (Wang et al., 2020; 2023; Longcore et al., 2024). Briefly, free- floating coronal sections, paired on the rostral-caudal axis and containing both the left and right SCN, were blocked for 1 h at room temperature (1% Bovine Serum Albumin, 0.3% Triton X-100, 10% normal donkey serum in 1xPBS) and then incubated overnight at 4°C with a rabbit polyclonal antiserum against cFos (1:1000, RRID: AB_2247211; Cell Signaling) followed by a Cy3-conjugated donkey-anti-rabbit secondary antibody (Jackson ImmunoResearch Laboratories, Bar Harbor, ME). Sections were mounted and coverslips applied with Vectashield mounting medium containing DAPI, and visualized on a Zeiss AxioImager M2 microscope (Zeiss, Thornwood NY) equipped with a motorized stage, an AxioCam MRm and the ApoTome imaging system. Z- Stack Images (35 images; 34mm, 1.029mm interval) of both the left and right middle SCN were acquired with a 20X objective using the Zeiss Zen digital imaging software, and two observers masked to the experimental groups performed the cell counting. The boundaries of the SCN were visualized using the DAPI nuclear staining and the cells immuno-positive for cFos counted with the aid of the Zen software tool ‘marker’ in three to five consecutive sections. The numbers obtained from the left and right SCN were averaged to obtain one value per section, and those from three-five sections averaged to obtain one value per animal and are presented as the mean ± standard deviation (SD) of 4 animals per genotype.

Histomorphometrical analyses of the SCN

Photographs of DAPI-stained sections generated from the WT and Fmr1 KO mice as described above were used to estimate the area, the perimeter, height and width of the SCN as previously reported (Li et al., 2015; Lee et al., 2018). For each animal, the four measurements were performed in three consecutive sections containing the middle SCN and acquired with a 10X objective and the Zen software. Measurements (in μm) of both the left and right SCN were obtained with the auxilium of the AxioVision software (Zeiss, Pleasanton, CA, USA). Because the borders of the DAPI-defined SCN are somewhat arbitrary, measurements were performed independently by two observers masked to the genotype of the animals. The area of the SCN in the three sections was summed, whilst the perimeter, height and width were averaged to obtain one value per side. No significant differences were found between the left and right SCN, therefore the values of the left and right SCN were averaged to obtain one value per animal. Data are shown as the mean ± standard deviation (SD) of 6 animals per genotype.

Blood sampling and measurements of plasma immune molecules

Blood was collected (∼0.5 mL per animal) in the beginning of the light phase via cheek puncture into microvette tubes coated with EDTA (Sarstedt, Numbrecht, Germany; Suppl. Fig. 1). Tubes were gently inverted a few times and placed immediately on wet ice. Within 1 h following collection, samples were centrifuged at 2500 rpm for 15 min at 4° C and the plasma collected into prelabeled Eppendorf tubes (Fisher Scientific, Hampton, NH), and immediately stored at -80° C until further processing using the Luminex Multiplex Assay at the UCLA Immune Assessment Core (https://www.uclahealth.org/pathology/services-immunoassays).

Statistical Analyses

Data analyses were performed using SigmaPlot 14.5 (Grafiti LLC, Palo Alto, CA) or Prism 10 (GraphPad Software, La Jolla, CA). The impact of the loss of Fmr1 on the waveforms of sleep/wake cycles was analyzed using repeated measures two-way analysis of variance (ANOVA) with time and genotype factors. While a two-way ANOVA with genotype and treatment as factors was used for the differences in sleep bouts between light phase and dark phase. The Holm-Sidak’s multiple comparisons test was applied to determine significant differences among the groups. The effect of TRF on the activity waveforms was analyzed using a three-way ANOVA with genotype, feeding regimen, and time as factors followed by Holm-Sidak’s multiple comparisons test. The datasets were examined for normality (Shapiro–Wilk test) and equal variance (Brown–Forsythe test); genotypic differences in the behavioral tests were determined by Student t-test or the Mann-Whitney test. Correlations between circadian/sleep parameters and other behaviors were examined by applying the Pearson correlation analysis. Normal distribution of the histomorphological datasets and the relative intensity of the Cholera Toxin fluorescent processes in the whole SCN were assessed using the Shapiro-Wilk test, and since these did not pass the normality test, a two-tailed Mann-Whitney test was employed to identify significant differences between groups. The effect of the loss of FMRP on the photic induction of cFos cells was assessed by one-way ANOVA followed by Bonferroni’s multiple comparisons test. Values are reported as the mean ± standard error of the mean (SEM) or mean ± standard deviation (SD).

Differences were determined significant if P < 0.05.

eLife Assessment

This manuscript presents solid experimental data using Fmr1 knockout mice to explore the fundamental role of Fmr1 in sleep regulation. The study supports the hypothesis that scheduled feeding can improve circadian rhythm and behavior in a mouse model of Fragile X syndrome. These findings may offer new insights into neurodevelopmental disorders and their potential treatment strategies.

Reviewer #1 (Public review):

The authors conducted a comprehensive investigation into sleep and circadian rhythm disturbances in Fmr1 knockout (KO) mice, a model for Fragile X Syndrome (FXS). They began by monitoring daily home cage behaviors to identify disruptions in sleep and circadian patterns, then assessed the mice's adaptability to altered light conditions through photic suppression and skeleton photoperiod experiments. To uncover potential mechanisms, they examined the connectivity between the retina and the suprachiasmatic nucleus. The study also included an analysis of social behavior deficits in the mutant mice and tested whether scheduled feeding could alleviate these issues. Notably, scheduled feeding not only improved sleep, circadian, and social behaviors but also normalized plasma cytokine levels. The manuscript is strengthened by its focus on a significant and underexplored area-sleep deficits in an FXS model-and by its robust experimental design, which integrates a variety of methodological approaches to provide a thorough understanding of the observed phenomena and potential therapeutic avenues.

Reviewer #2 (Public review):

Summary:

In the present study, the authors, using a mouse model of Fragile X syndrome, explore the intriguing hypothesis that restricting food access over the daily schedule will improve sleep patterns and subsequently enhanced behavioral capacities. By restricting food access from 12h to 6h over the nocturnal period (the active period for mice), they show, in these KO mice, an improvement in the sleep pattern accompanied by reduced systemic levels of inflammatory markers and improved behavior. These data, using a classical mouse model of neurodevelopmental disorder (NDD), suggest that modifying eating patterns might improve sleep quality, leading to reduced inflammation and enhanced cognitive/behavioral capacities in children with NDD.

Overall, the paper is well-written and easy to follow. The rationale of the study is generally well introduced. Data are globally sound. The interpretation is overall supported by the provided data.

Author response:

The following is the authors’ response to the previous reviews

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Thank you for the extensive response to my comments and questions.

Reviewer #2 (Recommendations for the authors):

(1) The Fmr1/Fxr2 double KO mice are not well described in the Introduction.

We have changed the sentence in the introduction to clarify that in Zhang et al ., 2008 they used a mouse lacking both the Fmr1 gene and its paralog Fxr2.

(3) The Authors decided not to discuss the potential translation of the present study to human patients, despite their final conclusion statement.

The paragraph below has been added to the end of the discussion:

“Translational Implications”

The present findings support the view that circadian disruption is not merely a downstream consequence of disease processes but actively contributes to symptom expression. Hence, the possibility that interventions designed to reinforce circadian rhythms can hold therapeutic value for individuals with FXS and related neurodevelopmental conditions. Given that sleep and circadian dysfunction are detectable early in development and are predictive of more severe clinical phenotypes, circadian-based interventions may be particularly beneficial if applied during periods of heightened neural plasticity. Importantly, time-restricted feeding represents a relatively low-cost, non-invasive strategy that could be feasibly implemented in realworld settings. Further translational work is needed to evaluate whether the mechanistic links identified here—between circadian misalignment, immune dysregulation, and behavioral impairments—are conserved in humans, and similar approaches can be implemented for clinical use.

Version 3

Abstract

Abstract

Fragile X syndrome (FXS), a leading inherited cause of intellectual disability and autism, is frequently accompanied by sleep and circadian rhythm disturbances. In this study, we comprehensively characterized these disruptions and evaluated the therapeutic potential of a circadian-based intervention in the fragile X mental retardation 1 (FMR1) knockout (KO) mouse. The Fmr1 KO mice exhibited fragmented sleep, impaired locomotor rhythmicity, and attenuated behavioral responses to light, linked to an abnormal retinal innervation and reduction of light-evoked neuronal activation in the suprachiasmatic nucleus. Behavioral testing revealed significant deficits in social memory and increased repetitive behaviors in the mutants, which correlated with sleep fragmentation. Remarkably, a scheduled feeding paradigm (6-hour feeding/18-hour fasting) significantly enhanced circadian rhythmicity, consolidated sleep, and improved social deficits and repetitive behaviors in the Fmr1 KO mice. This intervention also normalized the elevated levels of some pro-inflammatory cytokines, including IL-12 and IFN-γ, in the mutants’ blood, suggesting that its benefits extend to inflammatory pathways. These findings highlight the interplay between circadian disruption, behavior, and an inflammatory response in FXS, and provide compelling evidence that time-restricted feeding may serve as a promising non-pharmacological approach for improving core symptoms in neurodevelopmental disorders.

Introduction

Fragile X syndrome (FXS) is a relatively common inherited cause of intellectual disability, with a prevalence of approximately 1 in 4,000 males and 1 in 8,000 females (Hersh & Saul, 2011). Males are typically more severely affected due to their single X chromosome. FXS is caused by an abnormal expansion of CGG trinucleotide repeats in the fragile X mental retardation 1 (FMR1) gene, which results in the transcriptional silencing of FMR1 gene and consequent reduction in the levels of the fragile X mental retardation protein (FMRP), an RNA-binding protein essential for synaptic plasticity and neuronal cytoarchitecture (Zalfa et al., 2006; Soden et al., 2010; Kute et al., 2019). There is currently no cure for FXS. Existing treatments are primarily symptomatic, targeting common manifestations such as social deficits, repetitive behaviors, and sleep disturbances (Budimirovic et al., 2022; Mendez & Mendez, 2024). Given the essential role of sleep and circadian rhythms in neural recovery and plasticity, a compelling question is whether improving sleep/wake cycles might alleviate other core symptoms of this disease.

Animal models are invaluable for elucidating the mechanisms underlying circadian and sleep disturbances as well as developing interventions aimed at restoring healthy sleep/wake cycles. Among the available models for FXS, the Fmr1 knockout (KO) mouse has been extensively validated for investigating disease mechanisms and is widely used in preclinical drug development, including for sleep-related outcomes (Thomas et al., 2012; Kazdoba et al., 2014; Kat et al., 2022; Saré et al., 2022; Martinez et al., 2024). Early studies have shown disrupted rhythms in activity during the light/dark (LD) cycle in mice missing both the Fmr1 gene and the paralog Fxr2 (Zhang et al., 2008). More recent findings indicate impairments in both sleep (Saré et al., 2017) and activity rhythms (Bonasera et al., 2017; Angelakos et al., 2019) in Fmr1 KO mice, all consistent with the sleep/wake disturbances reported in FXS individuals (Kronk et al., 2009, 2010; Budimirovic et al., 2022).

Despite these findings, significant gaps remain in our understanding of the precise nature of the sleep and circadian deficits in Fmr1 KO mice, and whether targeted correction of these rhythms might mitigate behavioral symptoms. Furthermore, much of the behavioral research in Fmr1 KO mice has been conducted during the light phase—when mice are typically asleep—potentially confounding results due to sleep disruption.

In this study, we aimed at comprehensively characterizing sleep and circadian disturbances in the Fmr1 KO mice and examine their association with behavioral alterations. To avoid disrupting natural sleep patterns, we employed non-invasive home-cage monitoring systems to assess sleep and activity rhythms and conducted behavioral testing during the active (dark) phase. Importantly, the same animals were evaluated for both sleep/wake cycles and behavioral phenotypes, enabling direct correlations. Finally, because of the above mentioned evidence of circadian dysfunction, we tested the efficacy of a circadian-based intervention— scheduled feeding (6-hour feeding/18-hour fasting)—on improving sleep/wake rhythms, social behavior, and repetitive behaviors in these mutants.

Results

Fragmented Light-Phase Sleep in the Fmr1 KO Mice

To determine whether the Fmr1 KO exhibit deficits in immobility-defined sleep behavior, the animals were examined using a combination of video recording and mouse tracking system under a 12h:12h light/dark (LD) condition. Both the WT and mutant mice exhibited robust circadian rhythms in sleep with higher levels during the day and minimal sleep during the early dark phase (Fig. 1A). The amount of total sleep within the 24 hr cycle and the number of sleep bouts did not differ between genotypes, although the Fmr1 KO did exhibit a reduction in the average duration of each sleep bout (Table 1). A two-way ANOVA used to analyze the temporal pattern of sleep (1-hr bins) in WT and mutants revealed significant effects of time (F(23, 287) = 31.94; P < 0.001) and genotype (F(23, 287) = 11.95; P < 0.001), along with differences in day and night sleep parameters (Table 1). During the resting phase (lights on), the Fmr1 KO mutants slept significantly less (Fig. 1B) with more frequent sleep bouts (Fig. 1C) of shorter duration (Fig. 1D). The maximum sleep bout duration was about 20 minutes shorter in the mutant mice. The reduced sleep time, shorter bout length, and greater number of sleep bouts seen during the day are all evidence of a fragmented sleep pattern in the Fmr1 KO mutants. The negative findings should be viewed with caution as the sample size of n=6 per group is below ideal.

The Fmr1 KO mice exhibit reduced and fragmented sleep during the light phase.

(A) Waveforms of daily rhythms in sleep behavior under standard 12:12 h light-dark (LD) cycles in WT (blue circle) and Fmr1 KO (yellow triangle) mice. Zeitgeber time (ZT) 0 corresponds to lights on, and ZT 12 to lights off. Sleep was defined by immobility and binned in 1-hr intervals. Both genotypes exhibited clear diurnal rhythms, but the Fmr1 KO mice showed significantly less sleep during the light phase and at several times in the dark phase. Asterisks indicate significant differences between genotypes at individual time points (two-way ANOVA with genotype and time as factors, followed by the Holm-Sidak’s multiple comparisons test, *P < 0.05). The white/black bar on the top indicates the LD cycle; gray shading indicates the dark phase time-period. (B–D) Immobility-defined sleep metrics during the light-phase. Compared to WT, Fmr1 KO mice showed a greater number of sleep bouts, indicating sleep fragmentation, and of shorter duration. Histograms show the means ± SEM with the values from individual animals overlaid. Genotypic differences were analyzed by t-test (*P < 0.05). See Table 1.

Disrupted diurnal and circadian activity rhythms in the Fmr1 KO Mice

Next, we investigated the presence of deficits in locomotor activity rhythms in the Fmr1 KO mice. Age-matched WT and Fmr1 KO were housed in cages equipped with running-wheels under 12h:12h LD cycles for 2 weeks and then released into constant darkness (DD) to measure their endogenous rhythms (Suppl. Fig. 1). Both genotypes exhibited robust daily and circadian rhythms in wheel running activity (Fig. 2A; Table 2). Some indication of the commonly reported hyper-activity in the Fmr1 KO mice (Sørensen et al., 2015) emerged in the LD cycle (Fig. 2B), although the activity levels over 24 hours did not differ significantly (Fig. 2C) between genotypes. Analysis of the diurnal activity patterns with two-way ANOVA confirmed significant effects of time (F(23, 287) = 8.84; P = 0.003) and genotype (F(23, 287) = 39.75; P < 0.001) on the locomotor activity. In LD (Fig. 2C), the mutants exhibited lower rhythm power with increased activity during lights-on and more variability in the activity onset. When released in DD (Fig. 2D), WT and mutants displayed very similar free-running period (tau), but again, the latter presented with weaker rhythm power and a higher cycle-to-cycle variability in activity onset (Table 2). Together, our data suggest that the Fmr1 KO mice exhibit deficits in the circadian regulation of locomotor activity and raise the possibility of impaired ability to synchronize to the light cycle.

The Fmr1 KO mice show unstable locomotor rhythms.

(A) Representative actograms showing daily rhythms in wheel-running activity under LD followed by constant darkness (DD) in WT (left) and Fmr1 KO (right) mice. Activity levels were normalized to 85% of the most active mouse. Each row represents two consecutive days, and the second day is repeated at the beginning of the next row. (B) Waveforms of daily rhythms in cage activity in WT (blue circle) and Fmr1 KO (yellow triangle) mice under the LD cycles. Activity under LD (1-hr bins) was analyzed by two-way ANOVA with genotype and time as factors followed by the Holm-Sidak’s multiple comparisons test (*P < 0.05). There were significant effects of both time (F = 8.84; P = 0.003) and genotype (F = 39.75; P < 0.001) on the temporal pattern of the locomotor activity rhythms. Note that significant genotypic differences were found before and after dawn. Activity metrics in LD (C) and DD (D). Rhythmic power under both LD and DD conditions was significantly reduced in the mutants, which also presented higher imprecision in the activity onset in DD. Histograms show the means ± SEM with the values from individual animals overlaid. Genotypic differences were analyzed by t-test (*P < 0.05). See Table 2.

Impaired Behavioral Response to Photic Cues in the Fmr1 KO Mice

The likelihood of the Fmr1 KO mice presenting deficits in their response to light was tested using four behavioral assays of the retinal input to the circadian system. First, the ability of exposure to light (4500K, 50lx) during the dark phase to suppress or “mask” the locomotor activity of the nocturnal animals was evaluated. For this assay, the level of locomotor activity between Zeitgeber Time (ZT) 14 and 15 (two hours after lights off) was first measured under baseline LD conditions (i.e. the day before), and no differences were observed between genotypes (WT: 2229 ± 162 revolutions per hour (rev/h); KO: 2639 ± 328 rev/h; t(18) = −1.123, P = 0.277). The next day the mice were exposed to light for 1 hour at this same phase and the levels of activity measured. The mutants exhibited a significantly reduced light-driven suppression of the locomotor activity as compared to WT, (Fig. 3A, B; Table 3), with significant effects of both genotype (F(1, 39) = 12.002; P = 0.001, Two-way ANOVA) and light exposure (F(1, 39) = 24.253; P < 0.001, Two-way ANOVA). Next, we determined the number of cycles needed by the mice to re-entrain to a 6-hours phase advance of the LD cycle. While the WT mice re-synchronized in 5.9 ± 0.6 days to the new phase, the Fmr1 KO mice required 11.3 ± 0.4 days to adjust to the new ZT 12 (Fig. 3C, D; Table 3). Third, we evaluated the ability of the mice to entrain to a skeleton photoperiod (SPP) in which the full 12 hours of light or dark are replaced by two 1-hour light exposures (ZT 0-1 & ZT 11-12) separated by 11 hours of dark (Fig. 4A). Mice were first entrained to the standard LD cycle and then released into the SPP for 2 weeks (Fig. 4A). The activity recordings showed that the WT stably entrained to this challenging environment and exhibited robust circadian rhythms with a period (tau) of 24.0 ± 0.0 hr (Table 3). Conversely, the Fmr1 KO exhibited a tau of 23.7 ± 0.2 hr with the shorter period driven by three mutants that failed to entrain to the SPP (Table 3). Compared to the WT, the mutants also showed reduced power of the rhythms, increased light-phase activity, and a larger onset variability (Fig. 4B). Finally, the direct light-induced phase shift of the circadian system of the mice was measured by exposing them to light (300 lx) for 15min while in constant darkness at circadian time (CT) 16. The WT showed a phase delay of 135.6 ± 26.9 min, whilst, in the Fmr1 KO mutants, the delay in the activity onset on the next day was about half the magnitude (64.0 ± 0.1 min; Fig. 4C, D; Table 3). These findings provide evidence for the presence of a defective circadian response to light in the Fmr1 KO mice.

The Fmr1 KO mice display deficits in light-regulated circadian behaviors.

(A, B) Photic-suppression (masking) of activity in response to a 1-hr light (300 lx, 4500K) pulse at ZT 14 (lights off; n=10/genotype). The activity level during the light exposure was compared to the activity level during the equivalent hour (ZT 14-15) on the day before the treatment (baseline activity). (A) The genotypic difference in the fold change was determined by t-test, with the mutants showing a significantly reduced suppression of activity as compared to the WT (*P = 0.05). (B) Changes in the activity levels of each individual mice during the baseline window and the light masking were analyzed using a paired t-test. Within-genotype comparisons showed significant suppression in WT (P < 0.001), but not in KO mice (P = 0.12). (C, D) Entrainment induced by a 6-hr-phase advanced LD cycle. Representative actograms of light-induced phase shifts of wheel-running activity rhythms (C). The white/black bars on the top of actograms indicate the LD cycle before (upper) and after (lower) the 6-hr-phase advance. The gray shading in the waveforms indicates the dark phase time-period. The arrows next to the actograms indicate the day when the 6 hours-phase advance was applied. (D) Quantification of the days to re-entrain show that the KO mice required more time to adjust (two-way ANOVA: genotype effect F(1,285) = 130.157, P < 0.001; followed by the Holm-Sidak’s multiple comparisons test *P < 0.001). The entrainment shifting in the WT (blue circle) and the Fmr1 KO (yellow triangle) was quantified by the difference between the activity onset and the new ZT12 on each day. The yellow and blue arrow heads in the graph indicate the day when the activity rhythms are considered well entrained. See Table 3.

The Fmr1 KO mice exhibit difficulty in adapting to the skeleton photic period (SPP).

(A) Representative actograms of daily rhythms in cage activity under standard LD cycles (2 weeks) followed by the SPP (1h light:11h dark:1h light:11h dark) in WT (left) and Fmr1 KO (right) mice. The white/black bars on the top of actograms indicate the baseline LD cycle (upper) and the SPP LD cycles (lower). The gray shading in the waveforms indicates the time of the dark phases. (B) Measures of locomotor activity rhythms under SPP. Many of the parameters measured were significantly different between the genotypes with the mutants being more impacted and showing lower rhythmic power and increased variability of the activity onset. In addition, the mutants displayed higher activity during their day. Histograms show the means ± SEM with the values from each individual animal overlaid. Significant differences (P < 0.05), determined by t-test or Mann-Whitney test, are indicated with an asterisk. (C, D) Light-induced phase delay of free-running activity rhythms in mice exposed to light (300 lx, 4500K) for 15 min at circadian time (CT) 16. Mice were held in constant darkness, by definition, CT 12 is the beginning of the activity cycle in DD for a nocturnal organism. Examples of light-induced phase shifts of wheel-running activity rhythms (C) of WT (left) and Fmr1 KO (right) and quantified phase delay (D). In the representative actograms, the yellow lines indicate the best-fit line of the activity onset across the 10 days before and after the light pulse. The amount of phase delay is determined by the difference between the two lines on the day after the light pulse. The sunny-shape symbols indicate when the mice were exposed to light (CT16). Compared to WT, the Fmr1 KO showed reduced phase shift of their activity rhythms (Mann Whitney U *P = 0.011). See Table 3.

Subtle deficits in the retinal afferent innervation to the SCN in the Fmr1 KO Mice

The results described above suggest that the Fmr1 KO mice present with an anomalous retinal input to the suprachiasmatic nucleus (SCN), the master circadian clock located in the anterior hypothalamus. Hence, to assess the integrity of this pathway, the WT and Fmr1 KO mice received a bilateral intravitreal injection of the fluorescence-conjugated neurotracer Cholera Toxin (β subunit), previously used to map the projections of the melanopsin-expressing intrinsically photoreceptive retinal ganglion cells (ipRGCs) from the retina to the SCN (Muscat et al., 2003; Hattar et al., 2006). At variance with the WT, a lower fluorescent signal could be observed in the mutant mice both laterally and medially to the ventral SCN (Fig. 5A), where the retino-hypothalamic fibers reach the nuclei, as well as beneath in the optic chiasm. Analysis of the intensity and distribution (Suppl. Fig. 2) of the labelled retino-hypothalamic processes reaching and entering the ventral SCN showed a reduction, particularly evident in the lateral side of both the left and right mutant SCN (Fig. 5B, C). Likewise, a subtle decrease in the intensity of the labelled fibers was found in the whole SCN (Table 4) of the Fmr1 KO mice as compared to WT.

Abnormal Retinal-Suprachiasmatic Nucleus connectivity in the Fmr1 KO mice.

To trace the projections from the retina to the suprachiasmatic nucleus (SCN) via the Retino-Hypothalamic tract (RHT), WT and Fmr1 KO mice received a bilateral intravitreal injection of Cholera Toxin (β-subunit) conjugated to Alexa Fluor555 and were perfused 72 hours later. (A) Lower intensity of the fluorescently labelled RHT projections can be observed both laterally and medially to the ventral part of the SCN in the Fmr1 KO mice (white arrows) as compared to WT, suggesting a loss of afferent projections to the SCN. (B, C) Densitometric analysis of the distribution of the cholera toxin fluorescence intensity in the ventral SCN (Suppl. Fig. 2) of WT and Fmr1 KO mice. The intensity peaks of the profile plot of 4 to 5 consecutive coronal sections containing the middle SCN were aligned and then averaged to obtain a single curve per animal. Results are shown as the mean ± standard deviation (SD) for the left (B) and the right (C) SCN of each genotype. (D, E) Light-induction of cFos was greatly reduced in the SCN of the Fmr1 KO mice compared to WT. Mice held in DD, were exposed to light (300 lx, 4500K) for 15 min at CT 16, and perfused 45 minutes later (CT 17). (D) Representative serial images of light-evoked cFos expression in the SCN. The inset in the lower left panel shows the lack of cFos immunopositive cells in the SCN of mice held in DD but not exposed to light. Dotted magenta lines delineate the SCN. OC=Optic Chiasm, V=third Ventricle, (E) The number of immune-positive cells in the left and right SCN from 3-5 consecutive coronal sections per animal were averaged to obtain one number per animal and are presented as the mean ± SD per genotype. One-way ANOVA followed by Bonferroni’s multiple comparisons test, *P=0.0201. See Table 4

A well-established test of the light input to the circadian system is the light-evoked cFos response in the SCN. In line with the results obtained with the Cholera Toxin suggesting an impaired light-pathway to the SCN in the mutants, the number of cFos positive cells in the SCN of the Fmr1 KO was greatly reduced (50%) in comparison to the WT mice (Fig. 5D, E; Table 4). Contrary to other models of neurodevelopmental disabilities (NDDs; Li et al., 2015; Lee et al., 2018), the Fmr1 KO mice did not display any histomorphometrical alteration of the SCN (Table 4). Based on these findings, we can surmise the presence of a compromised connectivity between the retina and the SCN in the mutants, which could provide, at least in part, a mechanism for their difficulty in responding to photic cues.

Social and Repetitive Behavior Deficits in Fmr1 KO Mice

Social deficits and stereotypic symptoms are hallmark problems in autism spectrum disorders, which we sought to evaluate during the active (dark) phase, between ZT 16-18, in the two genotypes. Social behavior was tested with the three-chamber and the five-trial social interaction tests (Fig. 6). In the first stage of the three-chambers test, both genotypes showed more interest toward the stranger mouse when given the choice between an inert object and a mouse. The direct comparison of the time spent with the object and the mouse did not show significant differences between the genotypes (Table 5). On the other hand, measurement of the social preference index (SPI) indicated that the Fmr1 KO mice exhibited a reduced interest in the conspecific compared to WT (Fig. 6A; Table 5). In the second testing stage, when the lifeless object was replaced with a second novel mouse and the first stranger became the familiar mouse (Fig. 6B), the mutants spent more time with the familiar mouse than the WT (Table 5). Again, the social novelty preference index (SNPI) confirmed that the Fmr1 KO exhibited reduced interest in novel mouse compared to WT mice (Fig. 6B; Table 5). The reduced time spent exploring the novel-mouse chamber suggest that the mutants were, perhaps, unable to distinguish the second novel mouse from the first, now familiar, mouse, along with decreased sociability. The possibility of impaired social recognition was further tested with the five-trial social test (Fig. 6C). In this test, the first stranger mouse becomes the familiar mouse after 4 exposures to the testing mouse. When the second mouse is introduced in the 5th trial, the testing mouse typically shows a boosted interest in investigating the novel mouse. As expected, when the second stranger mouse was introduced to the testing animals in the 5th trial, the WT showed elevated social behavior; whilst the mutants did not display increased interest in exploring the second novel mouse (Fig. 6C; Table 5). No genotypic differences in social interaction in the first four trials were observed, indicating that habituation was unaltered in the mutants (Suppl. Fig. 3).

The deficits in social recognition and repetitive behaviors of the Fmr1 KO mice correlate with altered sleep behavior.

Tests of Social behavior: (A) In the first stage of the 3-chambers social test, when the testing mouse is given a choice between a stranger mouse and an inanimate object, the Fmr1 KO mice spent less time with the stranger mouse and had a lower social preference index (SPI) than the WT. (B) In the second stage, the testing mouse is given the choice between a chamber with a novel mouse and the one with the familiar mouse. The mutants spent less time with the novel mouse compared to the familiar one, and also in this phase exhibited lower social novelty preference (SNPI) as compared to the WT. (C) The possibility of reduced social recognition was further tested with the 5-trial social test. In this test, the stranger mouse becomes a familiar mouse after 4 exposures to the testing mouse, then a novel mouse is introduced in the 5th trial. The WT mice showed a higher interest for the novel mouse compared to the Fmr1 KO mice. Tests of repetitive behaviors: The amount of digging in the bedding (D) and the percentage of marbles buried (E) were measured with the marble bury test. The Fmr1 KO mice spent longer time digging and buried more marbles compared to WT. (F) Grooming behavior, assessed in a novel arena, was significantly higher in the Fmr1 KO mice as compared to WT. Histograms show the means ± SEM with the values from the individual animals overlaid. Significant differences (*P < 0.05) were determined by t-test or Mann-Whitney test. See also Table 5. Sleep duration (min, G, H) correlated with impaired social recognition and abnormal grooming behaviors. Sleep fragmentation, measured by number of sleep bouts, correlated only with grooming (I,J) (Pearson Correlation test). See Table 6

Next, when the active-phase repetitive behavior was examined using the marble bury test, the Fmr1 KO spent more time digging and buried more marbles compared to WT (Fig. 6D, E; Table 5). When the 30-min trial was divided into three 10-min intervals, the repetitive digging behavior was significantly higher in the Fmr1 mutants compared with the WT in all three intervals. Two-way ANOVA demonstrated that there was a significant effect of genotype (F(1, 107) = 11.04; P = 0.001) but not of the interval (F(2, 107) = 0.42; P = 0.66). The Fmr1 KO mice also exhibited significantly more grooming behavior than their WT counterpart (Fig. 6F, Table 5).

Finally, we used these data to determine if disrupted sleep correlated with autistic-like behavioral deficits. Sleep duration (Fig. 6G, H; Table 6) exhibited a moderate to strong correlation with both social recognition and grooming time, while sleep fragmentation (measured by sleep bouts number) only correlated with the latter (Fig. 6J); the length of sleep bouts (Table 6) showed moderate correlation with both social recognition and repetitive behavior. In addition, a moderate correlation was seen between grooming time and the circadian parameters, rhythmic power and activity onset variability (Table 6). In short, our work suggests that even when tested during their circadian active phase, the Fmr1 KO mice exhibit robust repetitive and social behavioral deficits. Moreover, the shorter and more fragmented the daytime sleep, the more severe the behavioral impairment in the mutants.

Amelioration of the Behavioral Disruptions by Scheduled Feeding in the Fmr1 KO Mice

The correlations between fragmented sleep and the severity of other behavioral phenotypes support the possibility that interventions focused on improving circadian rhythms and sleep may benefit the observed deficits in the Fmr1 KO mice. Scheduled feeding can be a powerful regulator of circadian rhythms (Long and Satchidananda, 2022; Manoogian et al., 2022) and has been shown to be effective in several disease models (Whittaker et al., 2018; Wang et al., 2018; Gao et al., 2022; Gupta et al., 2022; Whittaker et al., 2023).To test whether circadian reinforcement could ameliorate the observed deficits in the mutants, a new cohort of WT and Fmr1 KO mice was subjected to a time-restricted feeding (TRF) protocol (6-h feeding/18-h fasting, Suppl. Fig. 1) for two weeks and compared to counterpart animals held on ad libitum feeding (ALF). Both genotypes were able to well adapt to the TRF regimen and ate as much as their ALF counterpart after 5 days; at the end of the two-weeks, the TRF groups weighed less than the ALF groups (Suppl. Fig. 4).

The TRF treatment benefited several aspects of the temporal patterning of activity and sleep (Fig. 7) in both genotypes. The scheduled feeding improved the power of the rhythms and markedly restored the cycle-to-cycle variability in the activity onset of the mutants to WT levels (Fig. 7C-E), while reducing fragmentation (less activity bouts) in both WT and KO (Table 7). Total cage activity was not altered by the intervention in either genotype. By the second week, TRF also improved sleep behavior in both genotypes (Fig. 7F, G), with the Fmr1 KO mice showing increased total daytime sleep (Fig. 7H; Table 7), reduced sleep fragmentation, i.e. fewer bouts (Fig. 7I) of longer bout durations (Fig. 7J) as compared to their ALF counterpart. In general, strong genotypic differences were present in the response to TRF with the WT mice showing a more robust sleep response (Table 7).

Amelioration of sleep/wake rhythms in the Fmr1 KO mutants by Time Restricted Feeding (TRF).

(A, B) Waveforms of daily rhythms in cage activity using IR detection in the WT (circle) and Fmr1 KO (triangle) mice under ad libitum feeding (ALF) or TRF. The activity waveforms (1-hr bins) were analyzed using a three-way ANOVA with genotype, feeding regimen, and time as factors followed by Holm-Sidak’s multiple comparisons test. There were significant effects of genotype (F(1, 767) = 13.301; P < 0.001) and time (F(23,767) = 94.188; P < 0.001), as well as significant interactions between genotype and time (P < 0.001) and feeding regimen and time (P < 0.001) on the locomotor activity rhythms of both WT and Fmr1 KO mice. The green area indicates the time-period when the food hoppers were opened for 6 hours between ZT 15 and ZT 21. (C-E) Measures of locomotor activity rhythms: Both genotypes exhibited an increase in the power of the rhythms under TRF compared to ALF controls. The increase in early-day and late-night activity as well as the onset variability seen in the Fmr1 KO mice was corrected by the TRF. Data are shown as the means ± SEM; two-way ANOVA followed by Holm-Sidak’s multiple comparisons test with genotype and feeding regimen as factors, *P < 0.05 significant differences within genotypes (different feeding regimen); #P < 0.05 significant differences between genotypes (same feeding regimen). (F, G) Waveforms of daily rhythms of immobility-defined sleep. The sleep waveforms (1-hr bins) were analyzed by two-way ANOVA with time and feeding regimen as factors followed by the Holm-Sidak’s multiple comparisons test. There were significant effects of time for both WT (F(23, 351) = 9.828, P <0.001) and Fmr1 KO (F(23, 351) = 1.806, P = 0.014) mice, but not of feeding regimens. Missing data points precluded the use of three-way ANOVA for these measures. (H-J) Measures of immobility-defined sleep in the light phase. Both genotypes held on TRF exhibited an increase in sleep duration and in sleep bout length as well as a reduction in sleep fragmentation, measured by the number of sleep bouts, compared to their ALF counterparts. Data are shown as the means ± SEM; two-way ANOVA followed by Holm-Sidak’s multiple comparisons test with genotype and diet as factors, *P < 0.05 significant differences within genotype - between diet regimens; #P < 0.05 significant differences between genotypes - same feeding regimen. See Table 7.

The benefits of TRF were also evident on other aberrant behaviors in the Fmr1 KO model, and after two weeks of treatment, this feeding regimen improved social memory recognition while reducing the inappropriate grooming behavior (Table 8). Scheduled feeding improved social memory in the KO mice, by restoring the preference for the novel over the familiar mice to WT-like level (Fig. 8A; Table 8), suggesting that, perhaps, the treated mutants were able to distinguish the novel mouse from the familiar mouse. Grooming behavior was also significantly reduced by TRF in the Fmr1 KO mice, which showed a significant reduction in the time spent self-grooming (Fig. 8B), without affecting their locomotor activity (Fig. 8C). These data suggest that TRF has ameliorating effects on sleep/wake rhythms as well as autistic-like behaviors in the Fmr1 KO model.

Improved social memory and stereotypic grooming behavior in the Fmr1 KO mice after 2 weeks of TRF.

(A) Social memory was evaluated with the 5-trial social interaction test as described above. The social memory recognition was significantly augmented in the Fmr1 KO by the intervention, suggesting that the treated mutants were, perhaps, able to distinguish the novel mouse from the familiar mouse. The time spent in social interactions with the novel mouse in the 5th-trial was increased to WT-like levels in the mutants on TRF. Paired t-tests were used to evaluate significant differences in the time spent interacting with the test mouse in the 4th (familiar mouse) and 5th (novel mouse) trials. *P < 0.05 indicates the significant time spent with the novel mouse compared to the familiar mouse. (B) Grooming was assessed in a novel arena in mice of each genotype (WT, Fmr1 KO) under each feeding condition and the resulting data analyzed by two-way ANOVA followed by the Holm-Sidak’s multiple comparisons test with feeding regimen and genotype as factors. *P <0.05 indicates the significant difference within genotype - between diet regimens, and #P < 0.05 those between genotypes - same feeding regimen. (C) TRF did not alter the overall locomotion in the treated mice. See Table 8.

Attenuation of the Elevated Levels of Pro-inflammatory Markers in the Fmr1 KO Mice by TRF

An abnormal immune response has been suggested to play a role in FXS pathophysiology (Reynolds et al., 2021; Dias et al., 2022; Robinson-Agramonte et al., 2022). Hence, blood was collected at the end of the two weeks of the scheduled feeding regimen to investigate whether TRF would impact the profile of pro-inflammatory markers in the Fmr1 KO mice. The mutants on ALF displayed elevated levels of several cytokine in comparison with their WT counterpart (Table 9), in particular, Interleukin-12, (IL-12), Interferon-gamma (IFN_), and the Chemokine Ligand-9 (CXCL-9). These genotypic differences were abolished in the TRF-treated groups (Fig. 9A). Scheduled feeding also repristinated the Fmr1 KO levels of IL-2 to WT levels. Furthermore, elevated IL-12 and IFN-γ levels were significantly associated with markers of poor sleep, such as higher activity in the light phase, shorter sleep time/sleep bout length, and higher numbers of sleep bouts (Table 10), as well as with social memory deficits and excessive grooming (Fig. 9B, C; Table 10). These data suggest that TRF-dampening of the pro-inflammatory signaling in Fmr1 KO mice provides a potential mechanism to explain, at in least part, the observed behavioral improvements.

TRF repristinates the levels IL-12 and IFND in the plasma of Fmr1 KO mice to WT levels.

(A) The levels of selected plasma pro-inflammatory markers are shown. The increase in IL-12 and IFN_ in the mutants is lowered by TRF to WT levels. The full list of the assayed makers is reported in Table 9. Data were analyzed with two-way ANOVA followed by the Holm-Sidak’s multiple comparisons test with feeding regimen and genotype as factors. *P < 0.05 indicates the significant difference within genotype - between feeding regimen, and #P < 0.05 between genotypes – same feeding regimen. (B-G) Correlations between IL-12 or IFN-γ levels and social recognition or grooming behavior. IL-12 levels correlated with both behaviors. Data were analyzed using the Pearson Correlation, and the coefficients are reported in Table 10.

Discussion

Sleep and circadian rhythm disruptions are prevalent in individuals with NDDs, including FXS. Common issues include difficulty initiating and maintaining sleep and irregular sleep patterns, which significantly impair the quality of life of FXS individuals and their families/caretakers (Budimirovic et al., 2022; Minhas et al., 2024).

Notably, individuals exhibiting more severe behavioral symptoms often experience more pronounced sleep problems (Schreck et al., 2004; Taylor et al., 2012; Kaufmann et al., 2024), suggesting that circadian and sleep dysfunction exacerbate core NDD features. These disturbances are detectable quite early, even in infants and toddlers carrying the FMR1 mutation (D’Souza et al., 2020), raising the possibility that disrupted sleep/wake cycles may be an initial and influential feature of disease expression. Mechanistically, such disruptions could stem from direct effects of FMR1 loss on sleep and circadian circuits, or indirectly from secondary symptoms such as anxiety, sensory hypersensitivity, and repetitive behaviors.

Animal models of FXS, such as the Fmr1 KO mouse, are essential for dissecting the underlying mechanisms and testing interventions (Sandoval et al., 2024). In the present study, we demonstrate that loss of Fmr1 results in disrupted sleep and circadian rhythms, including higher fragmentation of the rest-phase sleep and abnormal diurnal activity patterns. These disruptions were associated with deficits in social memory and increased repetitive behaviors. Importantly, we found that a circadian-based intervention (TRF) not only improved rhythmicity and sleep consolidation but also ameliorated behavioral deficits and reduced peripheral inflammatory markers.

Circadian and Sleep Disturbances in Fmr1 KO Mice

Our behavioral sleep assay revealed reduced sleep duration and increased fragmentation during the light/rest phase in Fmr1 KO mice (Fig. 1), consistent with a previous report by Saré and colleagues (2017). In contrast, Westmark et al. (2023), using EEG, did not observe baseline sleep differences between genotypes, possibly reflecting methodological differences or subtle phenotypic variability. A key gap in the field is the lack of studies probing sleep homeostasis in Fmr1 KO mice; future studies using sleep deprivation protocols may reveal underlying vulnerabilities. Notably, the Fmr1 KO mice exhibit neural oscillation abnormalities, including increased resting gamma power and enhanced auditory-evoked responses, paralleling the EEG signatures in FXS patients (Jonak et al., 2024).

Assessment of circadian locomotor activity revealed weakened rhythmicity, elevated daytime activity, and a greater onset variability in Fmr1 KO mice (Fig. 2), consistent with prior studies (Bonasera et al., 2017; Angelakos et al., 2019). These deficits persisted in constant darkness (DD), indicating a compromised endogenous circadian clock. Differences in activity rhythms measured in LD conditions could be driven by the circadian clock or the direct effects of light and dark. Although earlier studies using Fmr1/Fxr2 double mutants reported severe rhythm loss (Zhang et al., 2008), the present study demonstrates that the loss of Fmr1 alone is sufficient to impair circadian regulation.

Impaired Light Response and SCN Connectivity in the Fmr1 KO Mice

To explore the basis of the observed dysregulations, we evaluated the photic input to the circadian system. Fmr1 KO mice displayed impaired negative masking, delayed re-entrainment to phase shifts, blunted light-induced phase shifts, and poor entrainment to a skeleton photoperiod (Figs. 3–4). These behavioral deficits suggest a compromised signal transmission from the retina to the SCN via the ipRGCs. Such hypothesis was confirmed by anatomical and functional assays; indeed, the ipRGC innervation of the SCN was lessened in the Fmr1 KO mice (Fig. 5A), and the light-evoked c-Fos induction substantially reduced (Fig. 5B). Given that the FMRP is expressed in the ipRGCs (Zhang et al., 2020), its loss may underline the deficient light response observed in the mutants. Interestingly, this reduced sensitivity to circadian light input contrasts with the general sensory hypersensitivity typical of FXS, highlighting a unique dissociation in sensory processing pathways (Rais et al., 2018). Whether these circadian-specific photic impairments are present in FXS patients warrants further investigation.

Despite these light input abnormalities, the Fmr1 KO mice did not exhibit a shortened free-running period (tau) in DD (Fig. 2D; Angelakos et al., 2019), nor were the rhythmic expressions of the clock genes altered (Zhang et al., 2008). These findings suggest that the core molecular clock remains intact, and the circadian dysfunction may arise from an altered excitatory/inhibitory (E/I) balance or impaired network synchronization within the SCN (Olde Engberink et al., 2023). Hence, our present results reinforce that loss of Fmr1 impairs the circadian system at the level of input and output regulation rather than at the core molecular oscillation.

Sleep and Behavior Correlate in the Fmr1 KO Mice

The Fmr1 KO model reproduces core behavioral deficits of FXS, including impaired social recognition and increased repetitive behaviors (Melancia & Trezza, 2018; Kat et al., 2022). By analyzing sleep, circadian activity, and behavior in the same animals, we identified significant correlations: mice with more disrupted sleep and weaker circadian rhythms exhibited more severe social and repetitive behavior impairments (Fig. 6). These findings parallel clinical studies showing that disrupted sleep predicts next-day behavioral difficulties in FXS and autism spectrum disorders (Cohen et al., 2018; Robinson-Shelton & Malow, 2016; Schwichtenberg et al., 2022; Minhas et al., 2024). Although correlational, these results suggest that targeting circadian and sleep dysfunction may offer a novel route for ameliorating some behavioral phenotypes present in certain NDDs.

Time-Restricted Feeding as a Therapeutic Strategy

Given the link between circadian disruption and behavior, we tested whether restoring rhythmicity could improve outcomes. TRF, a circadian-aligned feeding schedule, significantly improved the power of the rhythms, reduced cycle-to-cycle (onset) variability, and consolidated daytime sleep (Fig. 7). Importantly, TRF reduced repetitive behavior and improved social memory and interactions in the Fmr1 KO mice (Fig. 8), establishing proof-of-principle that circadian-based behavioral interventions can ameliorate core phenotypes of FXS.

Ketogenic diets have previously been shown to improve both sleep and seizure control in the Fmr1 KO mice, albeit with limited effects on other behaviors (Westmark et al., 2020, 2023, 2024). TRF, by contrast, appears to directly impact both behavioral and physiological endpoints, potentially through its entraining effects on peripheral and central clocks. Prior studies have shown that TRF improves behavior in aging and animal models of neurodegenerative disorders (Wang et al., 2018; Whittaker et al., 2018, 2023), but this is the first report to demonstrate benefits in a model of NDDs.

TRF Reduces Inflammatory Signatures in Fmr1 KO Mice

An emerging literature suggests that immune dysregulation accompanies FXS pathology (Ashwood et al., 2010; Careaga et al., 2014; Bertrand et al., 2021). In the present study, TRF reduced the abnormal levels of IL-12 and IFN-γ in the Fmr1 KO mice (Fig. 9A; Table 9), and the increased levels of these cytokines correlated with both sleep disturbances and behavioral impairments (Fig. 9B, C; Table 10). While most cytokines did not differ between genotypes, the normalization of IL-12 and IFN-γ levels following TRF supports a mechanistic link between inflammation and behavioral symptoms in FXS. These results are consistent with prior evidence implicating IL-12 in social and repetitive behavior in autism spectrum disorders (Ashwood et al., 2011; Fallah et al., 2020) and suggest that circadian-based interventions may exert part of their benefit by dampening low-grade inflammation.

Whether TRF exerts direct anti-inflammatory effects in the central nervous system remains unknown. Future studies should assess microglia activation, blood–brain barrier integrity, and central cytokine levels following TRF in the Fmr1 KO and related models. Moreover, defining the causal role of specific cytokines such as IL-12 and IFN-γ in behavioral regulation may offer novel therapeutic targets for FXS and related disorders.

Translational Implications

The present findings support the view that circadian disruption is not merely a downstream consequence of disease processes but actively contributes to symptom expression. Hence, the possibility that interventions designed to reinforce circadian rhythms can hold therapeutic value for individuals with FXS and related neurodevelopmental conditions. Given that sleep and circadian dysfunction are detectable early in development and are predictive of more severe clinical phenotypes, circadian-based interventions may be particularly beneficial if applied during periods of heightened neural plasticity. Importantly, time-restricted feeding represents a relatively low-cost, non-invasive strategy that could be feasibly implemented in real-world settings. Further translational work is needed to evaluate whether the mechanistic links identified here— between circadian misalignment, immune dysregulation, and behavioral impairments—are conserved in humans, and similar approaches can be implemented for clinical use.

Methods

Animals and Time-restricted feeding (TRF) paradigm

All experimental procedures were approved by the UCLA Animal Research Committee and conformed to guidelines from the UCLA Division of Laboratory Animal Medicine (DLAM) and the National Institutes of Health (NIH). Wild-type (WT) (JAX ID: 000664) and Fmr1 KO mice (Fmr1tm4Cgr) on the C57BL/6J background (JAX ID: 003025) were acquired from the Jackson laboratory (Bar Harbor, ME). These mutants have a neomycin resistance cassette replacing exon 5 of the fragile X mental retardation syndrome 1 (Fmr1) gene. The studies were carried out in male adult mice (3 to 5 months old) singly housed in light-tight ventilated cabinets in temperature- and humidity-controlled conditions, with ad libitum access to food and water unless otherwise stated. Cages were equipped with either running wheel or infrared motion sensors to monitor the sleep/wake behavior.

TRF was conducted as previously described (Wang et al 2018; Whittaker et al 2018; 2023). Briefly, WT and Fmr1 KO mice were singly housed and entrained to a 12h:12h light/dark (LD) cycle (300 lx vs 0 lx respectively) for a minimum of 2 weeks, then randomly assigned to one of two feeding conditions: food (standard chow) available ad libitum (ad lib) or available for 6 hr during the middle of the active phase from Zeitgeber Time (ZT) 15 to ZT 21. By definition ZT0 and ZT12 are, respectively, the onsets of lights turning on and off. Scheduled feeding was achieved by manually adding and removing food from the mouse cages, and a careful examination was carried out to ensure no small food fragments were dropped and remained in the cages.

Food consumption was measured by weighing food at the beginning and the end of the feeding cycles (ALF: 24 hr vs TRF: 6 hr). The ALF and TRF mice groups were held in these conditions for at least two weeks (14 days) plus 3 additional days to measure immobility-based sleep behavior (Suppl. Fig. 1).

This study used two different cohorts of mice for a total of 135 mice. The experiments shown in Fig. 1 through 6 were designed to compare the circadian phenotype of the Fmr1 KO mice to the WT and were carried out between 2017-2020 before the Covid pandemic research shutdown. After the restart of research, new cohorts of mice were obtained from Jackson Laboratory, and used to test the effects of scheduled feeding on the mutants’ behavioral and circadian deficits (Fig. 7 through 9).

Immobility-Based Sleep Behavior

Sleep Behavior was assessed under a 12h:12h LD cycle by video recording in combination with an automated mouse tracking analysis software system (Anymaze, Stoelting Co., Wood Dale, IL) as previously described (Lee et al., 2018; Whittaker et al., 2018). Sleep was defined as the periods when 95% of the animal remained immobile for at least 40 seconds or longer, a threshold shown to correlate with >99% accuracy to EEG-defined sleep (Fisher et al., 2016). For each animal, continuous recordings were performed over 5 days and data from two consecutive days with the fewest artifact-free epochs were averaged and used for further analysis. Data were exported in 1-min bins and analyzed separately for the light and dark phases. Total sleep time was determined by summing the immobility durations in the rest phase (ZT 0-12) or active phase (ZT 12-24). Sleep fragmentation was determined by the number of sleep bouts, which were operationally defined as episodes of continuous immobility with a sleep count greater than 3 per minute, persisting for at least 60 secs.

Locomotor cage activity rhythms

Methods used to characterize the sleep/wake cycles have been previously described (Lee et al., 2018; Wang et al., 2018; Whittaker et al., 2018). Single housed WT and the Fmr1 KO mice were habituated to a 12h:12h LD cycle for two weeks to ensure entrainment to the appointed LD schedule before data collection.

For the phenotypical comparisons between WT and Fmr1 KO, cage activity was recorded with running wheel sensors for at least 14 days in LD, then, after the assessment of sleep behavior (3 days), the animals were released for two additional weeks into constant darkness (12h:12h dark-dark, DD) to obtain measures of the endogenous rhythms (free-running activity, Suppl. Fig. 1). To investigate the effects of TRF, cage activity rhythms were monitored using overhead passive infrared (IR) motion sensors for two weeks (Wang et al., 2018; Whittaker et al., 2018).

Activity data were collected via the VitalView data recording system (Mini Mitter, Bend, OR) and ten days of recordings were used to obtain the period (tau), the rhythmic power as well as the waveform presentations. The analysis was conducted using the El Temps (A. Diez-Nogura, Barcelona, Spain) and ClockLab (Actimetrics, Lafayette Instruments, Lafayette, IN) programs as previously described (Lee et al 2018;

Wang et al., 2018; Whittaker et al., 2018). The periodogram generated by the El Temps uses χ2 test with a threshold of 0.001 significance, from these the amplitude of the periodicities is calculated at the circadian harmonic to obtain the power of the rhythmicity. The rhythmic power, or percentage of variation (%V), provides a measure of the strength of the mouse’s periodicity corrected for the activity amount and normalized to the percentage of variance derived from peak significance (P=0.05). The amount of cage activity over a 24hr period was averaged over 10 days and reported here as the arbitrary units (a.u.)/h. Fragmentation and imprecision of the daily onset of sleep/wake cycles were determined using ClockLab. Fragmentation was determined by the number of activity bouts per day, with one bout counted when activity was separated by a gap of 21 min or more (maximum gap: 21 min; threshold 3 counts/min). Imprecision, a measure of cycle-to-cycle variability or onset variability, was determined by calculating the daily variability in the time of activity onset from a best-fit regression line drawn through 10 days of activity in both LD and DD conditions using the ClockLab program (Actimetrics, Wilmette, IL, United States).

Photic Regulation of Circadian Behavior

Cohorts of age-matched WT and Fmr1 KO mice were single housed in cages with running wheels under a 12h:12h LD cycle (300 lx, 4500K) for two weeks and, after stable entrainment, exposed to four behavioral assays to test the photic regulation of their circadian system.

Photic suppression on nocturnal activity (negative light masking): mice were exposed to 1 hr of white light (300 lx, 4500K) at ZT 14 to determine their response to light exposure during the dark period (masking of nocturnal activity). The animals’ activity level, measured as the number of wheel revolutions, during the light pulse was compared to the level at the same hour (ZT 14 to 15) on the day before the exposure. Data are reported as percent of suppression= % [(rev during the 1-hr light exposure) – (rev during the dark baseline)]/(rev during the dark baseline).

Re-entrainment to a 6hr advance of the LD cycle: the mice underwent a 6-hr phase advance. The activity levels were continuously monitored by the wheel-running system and the activity onset determined with VitalView. The ability of the animals to re-entrain was quantified by the difference between the new ZT 12 (lights-off) and the onset of the running activity for each subsequent recording day. A mouse was considered fully re-entrained when the activity onset aligned to the new lights-off time for the consecutive 5 days.

Skeleton Photoperiod (SPP): the mice were transferred from the 12h:12h LD cycle to an SPP consisting of 1hL:11hD:1L:11hD cycles. Mice were exposed to the 1-hr light (300 lx) at the beginning (ZT 0) and the end (ZT 11) of the original light phase. The mice were kept in this photic condition for at least 2 weeks and activity was recorded and analyzed to determine rhythm power, period (tau) and other basic activity rhythm parameters using ClockLab.

Light-induced phase shift of activity rhythms: Mice were released into DD for 10 days. The circadian time (CT) of their free-running activity rhythms was determined using VitalView and El Temps software. The time of activity onset under DD was defined as CT 12. On day 11, the mice were exposed to light (300 lx, 4500K) at CT 16 for 15 min, and after the light pulse, they were held in DD for an additional 10 days. The best-fit lines of the activity onsets of sleep/wake cycles before and after light exposure were measured and compared.

Behavioral Assays

All these tests were conducted under dim red light (<2 lx) during the dark (active) phase between 2 and 4 h after lights-off (ZT 14–16) to avoid sleep disruptions after the sleep and activity recordings had been completed (Suppl. Fig. 1). Animals were habituated to the testing environment for at least 30 minutes prior to the behavioral assays.

The Stereotypic Behavior was assessed using two tests:

The marble burying test was used to evaluate repetitive digging behavior (Yrigollen et al., 2019). An array of 4 by 6 marbles was placed in the testing arena over a layer of 4-cm deep shavings. The testing mouse was introduced to the arena from the corner and its behavior recorded for 30 mins. After that, the testing mouse was carefully removed from the arena with special precaution to not move the marbles and returned to the home cages. The number of marbles buried more than ⅔ of their area was counted, and these were cleaned with 70% ethanol before the next use. Time spent on digging was manually scored by an observer masked to the experimental groups, whilst the distance travelled was derived from the automatic mouse-tracking system (Anymaze software, Stoelting Co., Wood Dale, IL).

The grooming test was conducted as in our previous work (Wang et al 2020). The mouse was introduced to the testing arena and its behavior recorded by a camcorder for 30 mins. The time spent on self-grooming, defined as the cleaning, licking, or washing of the limbs, tail, and body surface areas, typically from a head to tail direction, but excluding bouts of scratching, was manually scored by an observer masked to the experimental groups. The distance travelled was derived from the automatic mouse-tracking system (Anymaze software).

Social Behavior was assessed using two tests:

The three-chamber test was performed as previously described (Wang et al 2020). The testing mice were allowed to freely explore an arena with three chambers where the central chamber remained empty. When being habituated to the three-chamber arena, both the WT and the mutants explored the arena evenly with no preference toward the left or the right chamber (left/right ratio: WT: 0.95 ± 0.14; P = 0.43 by paired t-test. Fmr1 KO: 0.82 ± 0.1; P = 0.085 by paired t-test). The three-chamber test consisted of two parts: the first stage assessed the preference of social approach toward the stranger mouse, and the second stage assessed the ability of social discrimination of the testing mouse. In the first stage (social approach), an up-turned metal-grid pencil cup was placed in the side chambers: one was left empty as the novel object (the object chamber), while a never-met stranger mouse matched by sex, age, and genotype with the testing mouse was placed in the second up-turned cup (the social chamber). Thus, the testing mouse was tested for its preference between the object chamber and the social chamber in this first testing stage. In the second stage (social discrimination), the first stranger mouse and the cup remained the same, while a second stranger mouse matched by sex, age, and genotype with the testing mouse was placed in the second up-turned cup. In other words, the social chamber in the first stage became a familiar chamber in the second stage, and the object chamber of the first stage a novel chamber in the second stage. Thus, the preference between the familiar chamber and the novel chamber of the testing mouse was assessed in this second stage. Time spent in each chamber and the distance travelled were derived from the automatic mouse-tracking system (Anymaze software).

The 5-trial social test was conducted as previously described (Mineur et al., 2006). The testing mouse was first habituated to the arena for 30 mins and then introduced to a never-met stranger mouse for 4 trials. The testing mouse was allowed to explore and interact with the stranger mouse for 2 mins in each trial, then it was introduced to a second stranger mouse during the 5th trial for 2 mins. The resting interval between trials was 5 mins. Active social behaviors such as physical contacts (e.g. crawling over, social grooming), nose-to-nose sniffing, nose-to-anus from the testing mouse to the stranger mice were scored manually. Testing mice as well as stranger mice that showed aggressive behavior were withdrawn from the experiment.

Retinal-SCN Connectivity: Injection, Visualization and Analyses of the Neuroanatomical tracer Cholera Toxin

WT and Fmr1 KO mice (4 months-old) received a bilateral injection of CholeraToxin (β subunit) conjugated to Alexa Fluor™ 555 Conjugate (catalog number: C34776; Invitrogen™, Carlsbad, CA). Prior to the injections, the animals received a drop of a local ophthalmic anesthetic (Proparacaine HCl, 0.5%, Sandoz, Holzkirchen Germany) and an intraperitoneal (i.p.) injection of a non-steroidal anti-inflammatory drug (Carprofen, Zoetis, Parsippany-Troy Hills, NJ). The mice were then anesthetized with isoflurane and a 30G needle (BD PrecisionGlideTM Needle; Becton Dickinson, Franklin Lakes NJ) was inserted at a 45° angle into the sclera and the vitreous chamber, towards the base of the retina to allow leakage of vitreous humor. The Cholera Toxin (2 μg in 2μl of sterile PBS) was injected into the vitreous chamber using a 32G Hamilton syringe (Hamilton, Reno NV). The needle was left in place for ten seconds before being retracted. Seventy-two hours after the injection, the mice were euthanized with isoflurane (30%–32%) and transcardially perfused with phosphate-buffered saline (PBS, 0.1 M, pH 7.4) containing 4% (w/v) paraformaldehyde (PFA, Sigma). The brains were rapidly dissected out, post-fixed overnight in 4% PFA at 4°C, and cryoprotected in 15% sucrose.

Sequential coronal sections (40-50 μm) containing the left and right SCN were mounted, and the cover-slips applied with a drop of Vectashield-containing DAPI (4’,6-diamidino-2-phenylinodole; catalog number: H-1200; Vector Laboratories, Burlingame, CA). Sections were visualized on a Zeiss AxioImager M2 microscope equipped with an AxioCam MRm and the ApoTome imaging system, and images acquired with the Zeiss Zen software and a 10x objective to include both left and right SCN. Two methods of analyses were carried out on the images of 5 consecutive sections per animal containing the middle SCN (Lee et al., 2018). First, the relative intensity of the Cholera Toxin fluorescent processes was quantified in the whole SCN, both left and right separately, by scanning densitometry using the Fiji image processing package of the NIH ImageJ software (https://imagej.net). A single ROI of fixed size (575.99 μm x 399.9 μm, width x height) was used to measure the relative integrated density (mean gray values x area of the ROI) in all the images. The values from the left and right SCN were averaged per section and 5 sections per animal were averaged to obtain one value per animal. Second, the retinal innervation of the SCN is strongest in the ventral aspect, where the retino-hypothalamic fibers reach the nuclei, hence, the distribution of the Cholera Toxin fluorescent signal was also obtained for each left and right ventral SCN separately in the same 4-5 consecutive sections per animal using the Profile Plot Analysis feature of ImageJ (Lee et al., 2018). Briefly, a rectangular box of fixed size (415.38μm x 110.94 μm, width x height, Suppl. Fig. 2) to include the ventral part of the SCN was set for each side, and a column plot profile was generated whereby the x-axis represents the horizontal distance through the SCN (lateral to medial for the left and medial to lateral for the right; Suppl. Fig. 2) and the y-axis represents the average pixel intensity per vertical line within the rectangular box. Subsequent processing of the resulting profiles was performed for left and right SCN images separately. To average the profiles of the 5 sections and obtain a single curve per animal, fifth-order polynomial curves were fit to best estimate the position of the intensity peak on the x-axis and, using this position, the original y-axis values were aligned and averaged arithmetically [1 profile per section (either left or right), 5 sections per animal]. Analyses were performed by two observers masked to the genotype of the animals. Data are shown as the average profile ± SD of 3 animals per genotype.

Photic induction of cFos in the SCN and cFos-positive Cell Counting

A separate cohort of male WT and Fmr1 KO mice (3–4 months-old) was housed in DD conditions and exposed to light (300 lx, 4500K, 15 min) at CT 16. Forty-five mins later, the mice were euthanized with isoflurane (30%–32%) and transcardially perfused with phosphate-buffered saline (PBS, 0.1 M, pH 7.4) containing 4% (w/v) PFA. The brains were rapidly dissected out, post-fixed overnight in 4% PFA at 4°C, and cryoprotected in 15% sucrose until further processing. Sequential coronal sections (40-50 μm), containing the middle SCN, were collected on a cryostat (Leica, Buffalo Grove, IL) and further processed for cFos immunofluorescence as previously described (Wang et al., 2020; 2023; Longcore et al., 2024). Briefly, free-floating coronal sections, paired on the rostral-caudal axis and containing both the left and right SCN, were blocked for 1 h at room temperature (1% Bovine Serum Albumin, 0.3% Triton X-100, 10% normal donkey serum in 1xPBS) and then incubated overnight at 4°C with a rabbit polyclonal antiserum against cFos (1:1000, RRID: AB_2247211; Cell Signaling) followed by a Cy3-conjugated donkey-anti-rabbit secondary antibody (Jackson ImmunoResearch Laboratories, Bar Harbor, ME). Sections were mounted and coverslips applied with Vectashield mounting medium containing DAPI, and visualized on a Zeiss AxioImager M2 microscope (Zeiss, Thornwood NY) equipped with a motorized stage, an AxioCam MRm and the ApoTome imaging system. Z-Stack Images (35 images; 34mm, 1.029mm interval) of both the left and right middle SCN were acquired with a 20X objective using the Zeiss Zen digital imaging software, and two observers masked to the experimental groups performed the cell counting. The boundaries of the SCN were visualized using the DAPI nuclear staining and the cells immuno-positive for cFos counted with the aid of the Zen software tool ‘marker’ in three to five consecutive sections. The numbers obtained from the left and right SCN were averaged to obtain one value per section, and those from three-five sections averaged to obtain one value per animal and are presented as the mean ± standard deviation (SD) of 4 animals per genotype.

Histomorphometrical analyses of the SCN

Photographs of DAPI-stained sections generated from the WT and Fmr1 KO mice as described above were used to estimate the area, the perimeter, height and width of the SCN as previously reported (Li et al., 2015; Lee et al., 2018). For each animal, the four measurements were performed in three consecutive sections containing the middle SCN and acquired with a 10X objective and the Zen software. Measurements (in μm) of both the left and right SCN were obtained with the auxilium of the AxioVision software (Zeiss, Pleasanton, CA, USA). Because the borders of the DAPI-defined SCN are somewhat arbitrary, measurements were performed independently by two observers masked to the genotype of the animals. The area of the SCN in the three sections was summed, whilst the perimeter, height and width were averaged to obtain one value per side. No significant differences were found between the left and right SCN, therefore the values of the left and right SCN were averaged to obtain one value per animal. Data are shown as the mean ± standard deviation (SD) of 6 animals per genotype.

Blood sampling and measurements of plasma immune molecules

Blood was collected (∼0.5 mL per animal) in the beginning of the light phase via cheek puncture into microvette tubes coated with EDTA (Sarstedt, Numbrecht, Germany; Suppl. Fig. 1). Tubes were gently inverted a few times and placed immediately on wet ice. Within 1 h following collection, samples were centrifuged at 2500 rpm for 15 min at 4° C and the plasma collected into prelabeled Eppendorf tubes (Fisher Scientific, Hampton, NH), and immediately stored at −80° C until further processing using the Luminex Multiplex Assay at the UCLA Immune Assessment Core (https://www.uclahealth.org/pathology/services-immunoassays).

Statistical Analyses

Data analyses were performed using SigmaPlot 14.5 (Grafiti LLC, Palo Alto, CA) or Prism 10 (GraphPad Software, La Jolla, CA). The impact of the loss of Fmr1 on the waveforms of sleep/wake cycles was analyzed using repeated measures two-way analysis of variance (ANOVA) with time and genotype factors. While a two-way ANOVA with genotype and treatment as factors was used for the differences in sleep bouts between light phase and dark phase. The Holm-Sidak’s multiple comparisons test was applied to determine significant differences among the groups. The effect of TRF on the activity waveforms was analyzed using a three-way ANOVA with genotype, feeding regimen, and time as factors followed by Holm-Sidak’s multiple comparisons test. The datasets were examined for normality (Shapiro–Wilk test) and equal variance (Brown–Forsythe test); genotypic differences in the behavioral tests were determined by Student t-test or the Mann-Whitney test. Correlations between circadian/sleep parameters and other behaviors were examined by applying the Pearson correlation analysis. Normal distribution of the histomorphological datasets and the relative intensity of the Cholera Toxin fluorescent processes in the whole SCN were assessed using the Shapiro-Wilk test, and since these did not pass the normality test, a two-tailed Mann-Whitney test was employed to identify significant differences between groups. The effect of the loss of FMRP on the photic induction of cFos cells was assessed by one-way ANOVA followed by Bonferroni’s multiple comparisons test. Values are reported as the mean ± standard error of the mean (SEM) or mean ± standard deviation (SD).

Differences were determined significant if P < 0.05.

Data Availability

Any data will be available upon request. The raw data will be also deposited in Dryad, University of California Curation Center (UC3), California Digital Library.